Simple decoding of behavior from a complicated neural manifold

Version 1

Abstract

Decoders for brain-computer interfaces (BCIs) assume constraints on neural activity, chosen to reflect scientific beliefs while yielding tractable computations. We document how low tangling – a typical property of motor-cortex neural trajectories – yields unusual neural geometries. We designed a decoder, MINT, to embrace statistical constraints that are appropriate for these geometries. MINT takes a trajectory-centric approach: a library of neural trajectories (rather than a set of neural dimensions) provides a scaffold approximating the neural manifold. Each neural trajectory has a corresponding behavioral trajectory, allowing straightforward but highly nonlinear decoding. MINT consistently outperformed other interpretable methods, and outperformed expressive machine learning methods in 37 of 42 comparisons. Yet unlike these expressive methods, MINT’s constraints are known rather than the implicit result of optimizing decoder output. MINT performed well across tasks, suggesting its assumptions are generally well-matched to the statistics of neural data. Despite embracing highly nonlinear relationships between behavior and potentially complex neural trajectories, MINT’s computations are simple, scalable, and provide interpretable quantities such as data likelihoods. MINT’s performance and simplicity suggest it may be an excellent candidate for clinical BCI applications.

Introduction

A common goal in neural engineering and neuroscience is to estimate target variables, in real time, from observations of neural spiking activity. For brain-computer interfaces (BCIs), target variables may correspond to prosthetic motion [1–6], muscle activity [7–10], cursor control [11–17], navigation [18–20], speech [21–23], handwriting [24], or cognitive states [25–29]. In other applications, one may wish to estimate the state of the brain, a latent variable, for clinical monitoring [30], data visualization [31], or closed-loop neurally contingent experiments [32]. Spiking activity is high-dimensional (dimensionality matching the number of neurons) and can be variable across repeated trials of a nominally identical behavior. This complexity can be managed by utilizing statistical models that simplify the assumed relationship between spiking activity and variables of interest. For example, early BCI decoders estimated horizontal and vertical cursor velocity via weighted sums of neural activity, allowing velocity to be readily decoded even for spike patterns that were never observed during training. Ideally, estimation should leverage assumptions that are both accurate and strongly simplifying.

Perhaps the most common constraint assumed by models of spiking activity is the notion of ‘rate’. Rates are estimated in various ways by binning, filtering, and/or trial-averaging. Fundamentally, a neuron’s rate is a value determining its instantaneous probability of spiking. Statistically, this assumption has been powerful, especially in brain areas where spiking shows considerable variability. Furthermore, in artificial spiking networks, rate is a well-defined quantity whose properties justify those statistical models [33]. It is thus common practice to assume that neural activity, at each moment in time, can be summarized by a vector of continuously varying rates with one entry per neuron (the ‘neural state’). Spikes are then treated as noisy observations of the neural state.

A variety of methods for estimating the neural state and decoding behavior on single trials have been proposed [34–40]. Many of these leverage an additional assumption: that the neural state is low-dimensional, with the vector of rates exploring far fewer dimensions than the number of recorded neurons. Low-dimensional state estimation and decoding are often implicitly combined – e.g. linear decoders consider a neural state whose dimensionality matches the number of decoded variables [1–3, 5, 11, 12].

Many studies have usefully assumed further constraints on the neural state, beyond those related to dimensionality. For example, studies have leveraged assumptions regarding linear [37] or nonlinear [38] neural dynamics that constrain how the neural state can evolve. Multiple studies have assumed constraints on neural activity related to structure in behavior [13, 16, 17, 41–49]. Sani et al. [50] leveraged the assumption that only a subspace within the low-dimensional neural state is relevant to behavior. Schneider, Lee, and Mathis [51] and Zhou and Wei [52] assumed behaviorally-relevant structure in latent variables that relate nonlinearly to the full-dimensional neural state. Nonlinear decoders (including neural networks) can relax some constraints (e.g. linear relationships between behavior and neural state) while imposing or learning others [21, 23, 24, 38, 40, 53–61].

Each of these prior approaches succeeded by imposing structure on the estimation problem via constraints (explicit or learned) on the neural state or its relationship to behavior. Here, we propose a novel algorithm, Mesh of Idealized Neural Trajectories (MINT), that leverages different constraints. These constraints derive from advances – some recent and some presented here – in our understanding of the geometry of population activity and its relationship to movement. MINT takes a trajectory-centric view of neural geometry, in which the manifold of observable neural states is best approximated using stereotyped trajectories and interpolations between them. MINT assumes that the relationship between neural activity and behavior, while highly nonlinear, can be approximated by creating a correspondence between neural and behavioral trajectories. MINT is well-suited to situations where neural activity is shaped by dynamics, yet makes no assumption regarding the form of those dynamics. Because MINT’s assumptions were chosen to respect scientific beliefs rather than tractability, they might be expected to complicate computation. Fortuitously the opposite is true: MINT’s assumptions make it easier to compute desired quantities such as data likelihoods.

Below we document and quantify properties of motor-cortex population geometry – and its relationship with behavior – that are relevant to any decoder (and more generally to many scientific analyses and hypotheses). We describe how MINT was designed to respect and leverage these properties. We compare MINT’s decode performance with multiple other decode methods. We do so across a variety of tasks and behavioral variables, to document MINT’s flexibility. We also evaluate MINT as a method for neural-state estimation. MINT’s overall high level of performance illustrates the value of a trajectory-centric view that respects the empirical structure of population geometry.

Results

Neural trajectories are sparsely distributed, stereotyped, and can be high-dimensional

We analyzed 8 datasets consisting of simultaneous neural recordings and behavioral measurements from primates performing motor, sensory, or cognitive tasks. To begin, we focus on one dataset (MC_Cycle) recorded during a task where a primate grasps a hand-pedal and moves it cyclically forward or backward to navigate a virtual environment. We use this dataset to document some unusual properties of motor-cortex population geometry – properties that have implications for any BCI-decoding method. A variety of studies indicate that the dominant features of neural trajectories do not reflect encoded parameters or network outputs, but rather the computational needs of network solutions (e.g. [62, 63]). For example, noise-robust solutions require low trajectory tangling [64], resulting in neural trajectories that look very different from the outputs they create [65].

By definition, low-tangled neural trajectories never come close to crossing while traveling in different directions. For example, forward-cycling trajectories and backward-cycling trajectories appear to overlap when viewed in two-dimensions (Fig. 1a, top). Yet the apparent crossings do not result in high trajectory tangling; there exist other dimensions where forward and backward trajectories are well separated (Fig. 1a, bottom, same scale used for both). Separation keeps trajectory tangling extremely low – far lower than for the corresponding muscle trajectories.

Properties of neural trajectories in motor cortex, illustrated for data recorded during the cycling task (MC_Cycle dataset).

a) Low tangling implies separation between trajectories that might otherwise be close. Top. Neural trajectories for forward (purple) and backward (orange) cycling. Trajectories begin 600 ms before movement onset and end 600 ms after movement offset. Trajectories are based on trial-averaged firing rates, projected onto two dimensions. Dimensions were hand-selected to highlight apparent trajectory crossings while also capturing considerable variance (11.5%, comparable to the 11.6% captured by PCs 3 and 4). Gray region highlights one set of apparent crossings. Bottom. Trajectories during the restricted range of times in the gray region, but projected onto different dimensions: the top two principal components when PCA was applied to this subset of data. The same scale is used in top and bottom subpanels. b) Examples of similar behavioral states (inset) corresponding to well-separated neural states (main panel). Colored trajectory tails indicate the previous 150 ms of states. Data from 7-cycle forward condition. c) Joint distribution of pairwise distances for muscle and neural trajectories. Analysis considered all states across all conditions. For both muscle and neural trajectories, we computed all possible pairwise distances across all states. Each muscle state has a corresponding neural state, from the same time within the same condition. Thus, each pairwise muscle-state distance has a corresponding neural-state distance. The color of each pixel indicates how common it was to observe a particular combination of muscle-state distance and neural-state distance. Muscle trajectories are based on seven z-scored intramuscular EMG recordings. Correspondence between neural and muscle state pairs included a 50 ms lag to account for physiological latency. Results were not sensitive to the presence or size of this lag. Neural and muscle distances were normalized (separately) by average pairwise distance. d) Same analysis for neural and kinematic distances (based on phase and angular velocity). Correspondence between neural and kinematic state pairs included a 150 ms lag. Results were not sensitive to the presence or size of this lag. e) Control analysis to assess the impact of sampling error. If two sets of trajectories (e.g. neural and kinematic) are isometric and can be estimated perfectly, their joint distribution should fall along the diagonal. To estimate the impact of sampling error, we repeated the above analysis comparing neural distances across two data partitions, each containing 15-18 trials/condition.

Under the commonly assumed low-dimensional constraint, the manifold of observable neural states is equivalent to a subspace. This assumption is so standard that the terms ‘manifold’ and ‘subspace’ have often been used interchangeably. From that perspective, all states within that subspace are possible, with probabilities that may be described by a covariance matrix. For example, if two states are likely, the state between them is also likely. A family of low-tangled trajectories will often display the opposite property. Low-tangled trajectories typically require large voids that are never occupied. For example, circular trajectories are common [63, 64, 66]. Their intrinsically low tangling would increase if other trajectories (e.g. from other conditions) intruded into their center, but empirically this is avoided. Thus, the manifold of observable states may be far more sparsely distributed than is suggested by the simple constraint of a subspace or covariance matrix.

Maintaining low trajectory tangling can require adding ‘new’ dimensions to separate trajectories. Thus, neural-state dimensionality may be higher than the number of variables one wishes to decode [67, 68]. Furthermore, low trajectory tangling requires that two similar neural states lead to similar future neural states, even over timescales longer than those governing temporal smoothness. Thus, sequences of neural states over time may be stereotyped. Recent work used this assumption to compare network and biological solutions [69]. Stereotyped trajectories would also be consistent with recent BCI-adaptation results. BCI adaptation involves reusing existing patterns of neural activity [70], and trajectories do not reverse even when that would be advantageous [71, 72].

These observations argue that the appropriate constraints on the neural state have little to do with dimensionality per se – dimensionality may actually be quite high. Rather, constraints should reflect that neural states are sparsely distributed and belong to families of stereotyped trajectories. MINT makes assumptions that embrace these aspects of state-trajectory geometry. We briefly summarize these assumptions here, leaving a detailed description of MINT for a later section. First, MINT accommodates any dimensionality (high or low) by operating directly on the neural state rather than attempting to reduce its dimensionality. Second, MINT assumes the manifold of observable states is extremely sparsely distributed and therefore best approximated by leveraging the trajectories themselves rather than their covariance matrix or corresponding subspace. Third, MINT assumes that neural states are traversed in a stereotyped order (i.e. each neural state determines the subsequent state). MINT embodies these assumptions by approximating the manifold of observable neural states – and the transitions between them – with a ‘mesh’ of neural trajectories. A library of idealized trajectories samples the underlying manifold. Interpolations between neural states connect these trajectories into a mesh that spans the manifold more completely, allowing MINT to decode neural and behavioral states beyond those observed during training.

Neural and behavioral trajectories are non-isometric

The geometry of neural trajectories also has implications regarding how the neural state, once it is estimated, should be decoded into a behavioral state. To explore, we compared distances in neural and behavioral space. Well-separated neural states frequently corresponded to the same behavioral state. For example, in Figure 1b, the two red circles correspond to moments when behavior is essentially identical: cycling at ∼1.7 Hz with the hand approaching the ‘bottom’ of the cycle. Nevertheless, the corresponding moments are distant in neural state space. Similarly, the two gray squares correspond to two moments when the hand is stopped at the bottom (immediately before and after bouts of cycling) and are also quite distant. The neural and kinematic dimensions in Figure 1b were selected and oriented to highlight some resemblance between neural and behavioral trajectories. Yet, there are geometric differences even in this view, and these would only become more pronounced if more conditions (e.g. backward cycling) were included. As a consequence, the neural-to-behavioral-state mapping is many-to-one.

To compare trajectory-geometries quantitatively, we leveraged the fact that each neural distance has a corresponding behavioral distance, taken for the same pair of conditions and times (allowing for a latency shift). If geometries are similar, behavioral and neural distances should be strongly correlated. To establish a baseline that takes into account the impact of sampling error, we compared neural distances across two partitions. As expected, nearly all data fell along the diagonal (Fig. 1e). The joint distribution was strikingly non-diagonal when comparing neural versus muscle-population trajectories (Fig. 1c) and neural versus kinematic trajectories (Fig. 1d).

Thus, neural trajectories are isometric to neither muscle nor kinematic trajectories. Similar behavioral states frequently corresponded to dissimilar neural states. Defining similar kinematic states to be those with normalized distance <0.1, corresponding neural states were as close as 0.02 and as far as 1.46. Nevertheless, dissimilar behavioral states did not correspond to similar neural states. Defining dissimilar kinematic states to be those within 0.1 of distance 2, neural states were never closer than 0.53. These relationships facilitate decoding in some ways – different behavioral states imply different neural states – while also creating a challenge – decoding must map dissimilar neural states to similar behavioral states.

In principle, this many-to-one mapping could be accomplished linearly. Suppose neural trajectories mirrored behavioral trajectories, but with additional non-behavior-encoding dimensions. A linear decoder could simply ignore those additional dimensions. Yet this will be suboptimal if the ignored dimensions capture the majority of response variance. MINT leverages all dimensions by learning, directly from training data, which behavioral state is associated with each neural state in a library of idealized trajectories. This allows highly nonlinear mappings – including many-to-one mappings – without needing to choose a nonlinear function class and fit it to the data.

Training and decoding procedure

For training, MINT requires a library of neural trajectories (Ω) and a corresponding library of behavioral trajectories (Φ). Libraries can be learned in a variety of ways, depending on the nature of the task and training data. Because trial-averaging is particularly easy to apply, we focus first on libraries of neural trajectories learned in this way.

Consider a center-out reaching task. Neural and behavioral trajectories (Fig. 2a) have different geometries: stacked loops versus radially arranged and relatively straight. Trajectories are learned from a training set containing repeated reaching trials to each target location. Each moment in the training data corresponds to a condition (c) and an index (k) indicating progress through the movement. For example, c = 3 corresponds to the purple reach condition and k = 240 indicates a moment 240 ms into that movement. There will typically be many such moments in the training data: one for each trial of that type. Averaging across trials yields a behavioral state . Similarly, temporally filtering spikes and averaging across trials yields a neural state , containing each neuron’s estimated rate. Rate estimation can be further improved by a variety of means, including additional smoothing across neurons and/or conditions (see Methods). For some tasks, training data may not be organized by discrete conditions. Rates would then be estimated using methods devised for that purpose (e.g. [38, 40]).

Example training (top panel) and decoding (bottom panels) procedures for MINT, illustrated using four conditions from a reaching task.

a) Libraries of neural and behavioral trajectories are learned such that each neural state corresponds to a behavioral state . b) Spiking observations are binned spike counts (20 ms bins for all main analyses). contains the spike count of each neuron for the present bin, t′, and τ ′ bins in the past. c) At each time step during decoding, the log-likelihood of observing is computed for each and every state in the library of neural trajectories. Log-likelihoods decompose across neurons and time bins into Poisson log-likelihoods that can be queried from a precomputed lookup table. A recursive procedure (not depicted) further improves computational efficiency. Despite utilizing binned spiking observations, log-likelihoods can be updated at millisecond resolution (Methods). d) Two candidate neural states (purple and blue) are identified. The first is the state within the library of trajectories that maximizes the log-likelihood of the observed spikes. The second state similarly maximizes that log-likelihood, with the restriction that the second state must not come from the same trajectory as the first (i.e. must be from a different condition). Interpolation identifies an intermediate state that maximizes log-likelihood. e) The optimal interpolation is applied to candidate neural states – yielding the final neural-state estimate – and their corresponding behavioral states – yielding decoded behavior.

The library of neural trajectories (Ω) and the corresponding library of behavioral trajectories (Φ) are the parameters of MINT. They specify which states can be occupied and the order in which they are traversed. Prior work has utilized stereotyped behavioral trajectories [41, 46] or a mixture of condition-specific behavioral trajectory models that capture trial-by-trial movement variability [49]. For tractability, those methods assumed low-dimensional neural states that are roughly isometric to arm velocities. With MINT, neural trajectories can take any geometry, are learned empirically, and are related to behavioral trajectories via c and k, rather than by an assumed geometric isometry. The relationship between neural and behavioral states can be highly nonlinear and many-to-one. Although in Figure 2a there is one trajectory per target location, in practice there may be multiple trajectories per target (e.g. different speeds or curvatures) to enrich the library by allowing for movement variations. At the same time, it is not necessary for every potential target location and movement variation to be present in the training set. As will be discussed below, MINT can generalize in a handful of ways, including between any two conditions where neural activity and behavior both change smoothly. Furthermore, although the design of MINT is inspired by observations related to low tangling during movement, it is perfectly acceptable if there are moments of high tangling within Ω. Indeed this often occurs at transitions, e.g. from not-yet preparing to preparing a specific reach.

During decoding, incoming observations of spiking activity are binned (Fig. 2b, Δ = 20 ms for all analyses here). Suppose we are at time-bin t′ within a session where behavior is being decoded. Recent spiking observations are denoted by , where τ ′ specifies the number of previous time bins that are retained in memory. We compute the log-likelihood that we would have observed given each possible neural state, under a Poisson spiking model (Fig. 2c). The set of possible neural states is initially restricted to the states within the library of neural trajectories. This is a strong, but important assumption. If one were to consider all possible neural states in a 100-dimensional space, with each rate discretized into 10 possible values, it would be necessary to compute log-likelihoods for 10100 potential neural states. Yet computing log-likelihoods becomes tractable if trajectories are sparsely distributed. For example, even with 1000 samples per neural trajectory for the 4 neural trajectories in Figure 2, there would still be only 4000 possible neural states for which log-likelihoods must be computed (in practice it is fewer still, see Methods). Beyond reducing computational complexity, MINT’s assumptions regarding where the neural state can (and can’t) reside should aid decoding if those assumptions are well-aligned with the data.

MINT also leverages the order in which states are traversed, providing another massive reduction in complexity. MINT assumes that each neural state along each neural trajectory has a unique recent history that can be compared to the recent history of spike counts when computing log-likelihoods. Fortuitously, spike counts in non-overlapping bins are conditionally independent given rate in a Poisson spiking model. Thus, the log-likelihood associated with a particular neural state decomposes across time bins and neurons into Poisson log-likelihoods (which can be queried from a precomputed lookup table). Furthermore, many terms computed for decoding in one time bin can be reused at the next time bin. A recursive solution for exploiting this redundancy reduces the complexity of computing these log-likelihoods, leading to a lightweight procedure amenable to real-time applications. The assumption of stereotypy trajectory also enables neural states (and decoded behaviors) to be updated in between time bins. While waiting for the next set of spiking observations to arrive, MINT simply assumes that each neural state advances deterministically along the trajectory to which it is assigned.

Using the above procedure, one could simply take the maximum-likelihood neural state from the library, then render a behavioral decode based on the corresponding behavioral state. However, we wish the trajectory library to serve not as an exhaustive set of possible states, but as the basic structure defining a ‘mesh’ of possible states. Suppose the subject made a reach that split the difference between purple and blue reach conditions in Figure 2a. MINT would identify two candidate states with high likelihoods, along different neural trajectories, and interpolate (Fig. 2d). A line segment connecting those states is parameterized by α, ranging from 0 (purple state) to 1 (blue state). Because the log-likelihood of the observed spikes is a concave function of α, the optimal neural state can be rapidly identified online using Newton’s method. The same interpolation produces the final decode from the corresponding behavioral states (Fig. 2e).

The ‘mesh’ in MINT conveys that interpolation fills in the space between candidate states along different trajectories. The mesh is not fully defined during training because it is unknown which candidate states to consider until the spiking observations arrive. Rather, as spikes are observed during decoding, they will often be consistent with more than one trajectory. It is at that juncture that MINT utilizes interpolation to consider a space of intermediate trajectories that could have given rise to these observations. It is easiest to think of the mesh as connecting nearby trajectories, as in Figure 2d,e, and this is indeed what occurs in practice. While no assumption prevents MINT from interpolating between very different trajectories, this is empirically rare: the vast majority of decoded states are near an on-trajectory state. For on-trajectory states, direct decoding should be nearly perfect in determining the associated behavior. For states near trajectories, linear interpolation should provide good generalization because neural activity and behavior typically vary together smoothly.

Behavioral decoding in a multi-task network

MINT can be used in the same way for different tasks – the libraries differ but all other operations remain the same. If training data include multiple tasks, MINT can readily and implicitly switch tasks ‘on the fly’ based only on neural observations. MINT can do so even when the nonlinear relationship between neural activity and behavior is very different across tasks. If desired, MINT can report which type of behavior is presently being produced and decode only those variables relevant to that behavior. This could be useful, for example, if a subject were controlling both a computer cursor and a wheelchair. It is presently rare for physiological data to be recorded across multiple tasks in the same session. We thus demonstrate this task-switching ability for an artificial spiking network that performs two tasks on interleaved trials. This evaluation is also useful because the ground-truth neural state is known.

We evaluated MINT’s performance on data produced by an artificial recurrent network built from leaky-integrate-and-fire spiking neurons [33]. Network neurons exhibit realistic firing rates and realistic levels of spiking variability (i.e. spiking statistics are roughly Poisson). The network was trained to generate posterior deltoid activity (Fig. 3a) for a reaching task (similar to that in Fig. 2) and a cycling task (similar to that in Fig. 1). The network uses different strategies to perform the two tasks, and neural activity occupies distinct subspaces for each (Fig. 3b). This provides a stringent challenge; correlations between neural activity and behavioral output are completely different across tasks.

Examples of behavioral decoding provided by MINT for one simulated dataset and four empirically recorded datasets.

All decoding is causal; only spikes from before the decoded moment are used. a) MINT was applied to spiking data from an artificial spiking network. That network was trained to generate posterior deltoid activity and to switch between reaching and cycling tasks. Based on spiking observations, MINT approximately decoded the true network output at each moment. ‘R3’, ‘R4’, and ‘R6’ indicate three different reach conditions. ‘C’ indicates a cycling bout. MINT used no explicit task-switching, but simply tracked neural trajectories across tasks as if they were conditions within a task. b) Illustration of the challenging nature, from a decoding perspective, of network trajectories. Trajectories are shown for two dimensions that are strongly occupied during cycling. Trajectories for the 8 reaching conditions (pink) are all nearly orthogonal to the trajectory for cycling (brown) and thus appear compressed in this projection. c) Decoded behavioral variables (green) compared to actual behavioral variables (black) across four empirical datasets. MC_Cycle and MC_RTT show 10 seconds of continuous decoding. MC_Maze and Area2_Bump show randomly selected trials, demarcated by vertical dashed lines.

Decoding performance was excellent: R2 = .968 over ∼7.1 minutes of test trials. Decoded deltoid activity (green) was virtually identical to network output (black). This was true during both reaching (R3, R4 and R6 correspond to three reach directions) and cycling (C), with smooth transitions between them. MINT’s decoding approach was not hindered by the very different relationship between neural trajectories and behavioral output during the two tasks, nor by the dramatically different neuron-neuron covariance. Rather, MINT simply identified the most likely neural state at each moment and decoded the associated behavioral output.

The fact that the ground truth is known in this network illuminates a challenge that may impact decoding in many real-world situations. Even when a neural population or brain area exists to generate a particular output, that output may be encoded in low-variance dimensions and it may therefore be difficult to decode unless other dimensions can also be leveraged. For the present network, the output is simply a weighted sum of spikes, across all 1200 neurons. Thus, a linear readout (learned via ridge regression) that leveraged all 1200 neurons decoded network output exceptionally well (R2 = .982). However, the output dimension in this network captures exceedingly little of the total variance of network responses (∼2%), and thus is unlikely to perform well if only a subset of neurons were recorded. Indeed, when only using 5% of neurons in the network, linear decoding performance dropped to R2 = .773. In contrast, MINT’s performance only dropped by .001 to R2 = .967. Thus, even when the ground-truth readout is known, MINT can provide improved performance by leveraging all neural dimensions, not only those that directly impact behavioral output. Furthermore, MINT can readily decode other variables (as will be shown below) that are unlikely to be directly controlled by linear sums of spiking activity.

Behavioral decoding across multiple datasets

We investigated how MINT performed on four datasets in which primates performed a variety of motor tasks. All tasks involved simultaneously collected spiking activity and behavioral measurements. MC_Cycle is the same dataset from Figure 1. MC_Maze involves neural recordings from primary motor cortex (M1) and dorsal premotor cortex (PMd) during straight and curved reaches [73]. Area2_Bump involves neural recordings from area 2 of somatosensory cortex during active and passive center-out-reaches [74]. MC_RTT involves M1 recordings during point-to-point reaches without condition structure [59]. All decoding was offline, to allow evaluation across a variety of task types and comparison with other decode methods (below) across multiple decoded variables. All decoding was causal; i.e. based on spikes from before the decoded time. All decoding examples and performance quantifications utilized held-out test sets. Discussion of how MINT’s (few) hyperparameters were selected for the analyses in Figure 3 will be provided in the next section.

Decoding closely tracked actual behavior (Fig. 3c). For the MC_Cycle dataset, MINT closely tracked pedal phase and angular velocity. Decoding was virtually lag-free (on average): phase was decoded with near-zero lag, and transitions between rest and cycling occurred with near-zero latency. This was true despite the fact that all decoding was causal. During periods of non-movement, decoded angular velocity remained flat with few false movements. To illustrate, consider a 70 second stretch of cycling (Supp. Video) that included ∼30 seconds of true non-movement (true angular speed <.05 Hz). Small decode mistakes (decoded angular speed >0.1 Hz) did occasionally occur (e.g. Fig. 3c, small ‘blip’ before the middle cycling bout). Yet such mistakes were rare: decoded angular speed was <0.1 Hz for ∼99% of non-moving times.

A subtle point is that MINT assumes stereotyped trajectories when computing data likelihoods, but this does not cause the maximum-likelihood state to follow a stereotyped trajectory during decoding. For example, individual trials rarely have a cycling speed that perfectly matches the library trajectory. Yet this does not cause a growing discrepancy between decoded and actual phase. If cycling is slightly faster than the library trajectory, MINT advances along that trajectory at a faster-than-usual rate. This flexibility is essential to good performance, but also reveals itself during errors when behavioral trajectories exhibit features (e.g. the small blip noted above) not present in any library trajectory.

In the MC_Maze dataset, decoded velocity tracked actual reach velocity across a wide variety of straight and curved reaches (108 conditions). Decoded and actual velocities were highly correlated (Rx = .962, Ry = .950). We quantified error in absolute terms using the mean absolute error (MAE). Absolute errors were low (MAEx = 4.6 cm/s, MAEy = 4.8 cm/s) relative to the range of x- and y-velocities (224.8 and 205.7 cm/s, respectively).

Decoding performance was also high for Area2_Bump (Rx = .939, Ry = .895; MAEx = 4.1 cm/s, MAEy = 4.4 cm/s). Errors were slightly smaller than for the MC_Maze dataset in absolute terms, but were modestly larger in relative terms because the behavioral range of velocities was smaller. Errors are thus slightly clearer in the traces plotted in Fig. 3c. The overall high performance – and minimal lag – is notable given that the decode was based on spiking in a primary sensory area.

The task used in the MC_RTT dataset was not designed to include repeated trials of the same movement. Thus, when learning neural trajectories, we used a method that doesn’t rely on averaging: AutoLFADS [39]. This yielded one long neural trajectory during training (only broken by recording gaps). MINT’s interpolation procedure was modified to 1) occur across states >1000 ms apart, and 2) perform multiple pairwise interpolations between candidate-state pairs and use that which yielded the highest log-likelihood. The first modification ensures interpolation must consider different movements, even without a condition structure. The second modification accommodates the fact that this less-constrained task has many degrees of behavioral freedom. There may thus be more than two high-probability candidate states that are worth considering.

This strategy yielded decoded velocities (Fig. 3c) that correlated well (Rx = .786, Ry = .843) with actual velocities, but less well than for the other three datasets. Absolute errors were lower than for the other reaching datasets (MAEx = 2.7 cm/s, MAEy = 2.1 cm/s). However, x- and y-velocity ranges were also lower (86.6 and 61.3 cm/s). As will be documented below, decode performance for the MC_RTT dataset was lower, relative to that for the other three datasets, not only for MINT but also for other decode algorithms. There are two likely reasons. First, MC_RTT contained by far the slowest reaches of the three datasets. When analyzing only movement periods (reach speed >1 cm/s), the median speed across all movement times was 38.1, 16.8, and 5.9 cm/s for MC_Maze, Area2_Bump, and MC_RTT, respectively. Second, the ability of AutoLFADS to denoise trajectories based on inferred dynamics may have been limited by the relatively short (8.1 minutes) amount of training data. This issue aside, the MC_RTT task provides a useful demonstration that the neural trajectories needed to parameterize MINT can be obtained without a standard condition structure and without trial averaging.

Comparison to other decoders

We compared MINT to four other decode algorithms: the Kalman filter, Wiener filter, feedforward neural network, and recurrent neural network (GRU). The Kalman filter and Wiener filter are historically popular BCI algorithms [1, 43] that are computationally simple and make linear assumptions. The feedforward neural network and GRU are expressive nonlinear function approximators that have been proposed for neural decoding [54, 75] and can be implemented causally. At each decoding time, both the feedforward neural network and the GRU leverage recent spiking history. The primary difference between the two is whether the networks process that history all at once (feedforward network) or sequentially (GRU).

Each method was evaluated on each of four datasets, generating predictions for behavioral variables on held-out test sets (Fig. 4). All decoders were provided with a trailing window of spiking observations (binned in 20 millisecond increments). The length of this trailing window was optimized separately for each decoder and dataset (and, like all hyperparameters, was chosen using validation data). For non-MINT decoders, hyperparameters were tuned, using Bayesian optimization, to maximize performance [76]. MINT’s few hyperparameters were less extensively optimized. For example, window length was optimized once per dataset, rather than separately for different sets of behavioral variables (as was done for the non-MINT decoders). This choice embraces an unusual and useful aspect of MINT: there is no need to retrain across behavioral variables. Once the neural state has been estimated, all behavioral variables are readily and equally decodable. Less-extensive optimization is unlikely to have put MINT at a meaningful disadvantage; MINT is robust across reasonable hyperparameters choices (Fig. S1).

Comparison of decoding performance, for MINT and four additional algorithms, for four datasets and multiple decoded variables.

On a given trial, MINT decodes all behavioral variables in a unified way based on the same inferred neural state. For non-MINT algorithms, separate decoders were trained for each behavioral group. E.g. separate GRUs were trained to output ‘position’ and ‘velocity’ in MC_Maze. Parentheticals indicate the number of behavioral variables within a group. E.g. ‘position (2)’ has two components: x- and y-position. R2 is averaged across behavioral variables within a group. ‘Overall’ plots performance averaged across all behavioral groups. R2 for feedforward networks and GRUs are additionally averaged across runs for 10 random seeds. The Kalman filter is traditionally utilized for position- and velocity-based decoding and was therefore only used to predict these behavioral groups. Accordingly, the ‘overall’ category excludes the Kalman filter for datasets in which the Kalman filter did not contribute predictions for every behavioral group. Vertical offsets and vertical ticks are used to increase visibility of data.

By necessity and desire, all comparisons were made offline, enabling benchmarked performance across a variety of tasks and decoded variables, where each decoder had access to the exact same data and recording conditions. There were 15 total ‘behavioral groups’ across the four datasets. For example, MC_Maze yielded two behavioral groups, position and velocity, each of which had horizontal and vertical components.

MINT yielded the best performance, of all decoders, for 11/15 behavioral groups. For 3 of the 4 behavioral groups where MINT’s performance was not the highest, the deficit relative to the best method was very small: ΔR2 < .004. We first consider position and velocity, two variables commonly focused on in BCI decoding. For position, three decoders – MINT, the GRU, and the feedforward network – provided extremely similar and very accurate performance. For velocity, MINT provided a small but noticeable improvement over the GRU and the feedforward network for two datasets (MC_Maze and Area2_Bump), was modestly worse for one dataset (MC_RTT), and was involved in essentially a three-way tie for one dataset (MC_Cycle). MINT tended to display the largest improvements, relative to the next-best decoder, when decoding muscle-related variables (EMG, force, muscle velocity, and muscle length) and phase (circular variables can be challenging for some methods but are not for MINT). Relative to the other computationally ‘lightweight’ methods – the Wiener and Kalman filter – MINT’s performance was consistently better, often by a large margin.

For each dataset, we computed overall performance by simply averaging across behavioral groups. In three of the four datasets (all but MC_RTT, discussed further below), overall performance was highest for MINT. Thus, despite using simple and interpretable operations, MINT is extremely competitive with highly expressive machine learning methods. Within these three datasets (MC_Cycle, MC_Maze, Area2_Bump), MINT performed well for every individual behavioral group. This consistency reflects a feature noted above: once the neural state is estimated, it becomes possible to decode any behavioral variable that has a relationship with the neural state (though some, such as EMG, may of course be intrinsically ‘noisier’).

MINT’s lowest performance, relative to other decoders, occurred for the MC_RTT dataset. MINT outperformed the Kalman filter and Wiener filter. Yet improved decoding was achieved by the feedforward network and GRU. Thus, this was the only situation, across all datasets and behavioral groups, where MINT was ever at a meaningful performance disadvantage. A likely explanation is that the training data in the MC_RTT task are not ideally suited for estimating the library of trajectories, limiting subsequent performance on test data. Thus, if using MINT is a goal for tasks of this type, more structured training data (e.g. repeated trials of the same type) might be helpful. Alternatively, it may be sufficient simply to collect more training data. This dataset contains only 8.1 minutes of training data, which may limit the ability of AutoLFADS to estimate single-trial rates. If collecting more training data is impractical, there exist methods that improve estimation of single-trial rates (and thus trajectories) based on data from other sessions using the same task [38]. We do not pursue these avenues here, but simply stress that the principal limitation on MINT’s performance is likely to be the ability to estimate an accurate library of trajectories. When doing so is challenging, MINT may be at a disadvantage. Conversely, if a library can be accurately estimated, MINT is likely to perform well across a broad range of decoded variables. In this way, MINT can be seen as largely reducing the problem of decoding to a problem of estimating a library of trajectories.

When is generalization possible across tasks?

A natural hope of BCI decoding is to find a mapping between neural and behavioral states that generalizes across tasks, eliminating the need for task-specific training data. Whether this is achievable is not known; a basic challenge is that neural activity can sometimes occupy different subspaces across tasks. To explore the implications of such geometry, we leveraged the MC_Cycle dataset. Neural activity during forward and backward cycling typically occurs in rather different subspaces [64]. In the dataset used here, those subspaces are quite close to orthogonal (principal angles of 88°and 73°between the top two principal components for forward steady-state cycling and the top two principal components for backward steady-state cycling). When asked to generalize across cycling directions, all decode methods (including MINT) generalized poorly. Across all behavioral groups and all decode methods, generalization R2 values were typically negative and the largest positive value was 0.042. This was true even though the marginal distributions of x- and y-position and x- and y-velocity were similar across cycling directions.

The failure of all methods makes sense given the near-orthogonality of forward and backward neural trajectories (Fig. S2). MINT has no decoding advantage in this regard, but its focus on the geometry of neural activity may make it easier to anticipate when generalization is and isn’t likely to be possible. When neural trajectories are nearly orthogonal to those learned during training, the existing ‘mesh’ will not support generalization. This knowledge allows such situations to be recognized and handled appropriately.

Modeling & preprocessing choices

MINT uses a direct mapping from neural states to behavioral states and uses interpolation to potentially improve decoding. To determine the impact of these choices, we ran a full factorial analysis comparing the direct mapping to a linear mapping (Fig. S3a), and assessing the impact of interpolation (Fig. S3b). Both choices did indeed improve performance. The analysis also assessed how acausal decoding performed relative to causal decoding (Fig. S3c). Causal decoding is a requirement for real-time BCIs, and thus is employed for all our main decoding analyses. However, future applications could potentially introduce a small lag into online decoding, effectively allowing some ‘acausal decoding’ at the expense of longer-latency responses. As expected, acausal decoding provided modest performance increases, raising the possibility of an interesting tradeoff. Decoding can be zero-lag (causal), positive-lag (acausal), or even negative-lag. The latter would result in a very ‘snappy’ decoder that anticipated actions at the expense of decoding accuracy. The lag hyperparameter wouldn’t need to be specified prior to training and could be freely adjusted at any time.

All datasets were curated to contain sorted spikes. Yet during online performance, decoding must typically rely on unsorted threshold crossings. Prior studies have found that moving from sorted spikes to threshold crossings produces a modest but tolerable drop in performance [77]. We similarly found that moving from sorted spikes to threshold crossings (selected to be from ‘good’ channels with reasonable signal-to-noise) produced only a small drop in performance (Fig. S3d). The operations of MINT highlight why robustness is expected. When spikes from two neurons are combined, the resulting ‘unit’ acts statistically as if it were a high firing-rate neuron. Spiking is still approximately Poisson, because the sum of Poisson processes is a Poisson process. Thus, MINT’s statistical assumptions remain appropriate.

Neural state estimation

MINT’s success at behavioral decoding suggests underlying success in neural-state estimation. However, that would not necessarily have to be true. For example, the GRU also typically supported excellent behavioral decodes, but doesn’t provide neural state estimation. MINT does (by its very construction) but whether those estimates are accurate must be determined directly.

To evaluate neural-state estimates, we submitted firing-rate predictions to the Neural Latents Benchmark [78], an online benchmark for latent variable models that compares acausal neural-state estimates across methods and datasets. The four primary datasets for the benchmark are MC_Maze, Area2_Bump, MC_RTT, and an additional dataset, DMFC_RSG, that contains neural recordings from dorsomedial frontal cortex while a monkey performed a cognitive timing task [79]. Three secondary datasets (MC_Maze-L, MC_Maze-M, and MC_Maze-S) contain additional sessions of the maze task with progressively fewer training trials (500, 250, 100, respectively).

By submitting to the Neural Latents Benchmark, we can ask how MINT performs relative to a set of ‘baseline’ methods. One baseline method (smoothed spikes) is traditional and simple. Other baseline methods, Gaussian Process Factor Analysis (GPFA) and Switching Linear Dynamical System (SLDS), are modern yet well-established. Finally, two baseline methods are quite new and cutting-edge: Neural Data Transformers (NDT) [40] and AutoLFADS [39]. These are expected to provide a high bar by which to judge the performance of other approaches.

Although the neural state is a well-defined quantity in spiking models [33], experimental data does not typically provide a ground-truth neural state against which estimates can be compared (unlike for behavioral-state decoding). Yet one can define attributes the neural state should possess, then determine whether estimates display those attributes. One critical attribute is prediction of spiking activity. This attribute is assessed via bits per spike, which is closely related to the log-likelihood of spiking observations given a set of firing rates (assuming Poisson spiking). MINT performed similarly to AutoLFADS and was consistently slightly better than NDT on the four primary datasets (Fig. 5a, left). MINT performed slightly better than AutoLFADS and NDT on the largest of the secondary datasets, and that advantage grew as dataset-size became smaller (Fig. 5a, right). MINT outperformed the other three baseline methods by modest-to-sizeable margins for all seven datasets.

Evaluation of neural state estimates for seven datasets.

a) Performance quantified using bits per spike. The benchmark’s baseline methods have colored markers. All other submissions have gray markers. Vertical offsets and vertical ticks are used to increase visibility of data. Results with negative values are designated by markers to the left of zero, but their locations don’t reflect the magnitude of the negative values. Data from Neural Latents Benchmark (https://neurallatents.github.io/). b) Performance quantified using PSTH R2. c) Performance quantified using velocity R2, after velocity was decoded linearly from the neural state estimate. A linear decode is used simply as a way of evaluating the quality of neural state estimates, especially in dimensions relevant to behavior.

MINT was not designed to maximize bits per spike – good performance in this regard simply results from the primary goal of estimating a neural state that is expressed in terms of firing rates. This matters because there exist spiking features – e.g. synchrony beyond that due to correlated rates – that some methods may be able to leverage. Doing so is reasonable but separate from MINT’s goal of estimating a rate-based neural state. It is thus worth considering other attributes expected of good neural-state estimates. The Neural Latents Benchmark includes two such attributes.

First, good neural state estimates should be correct on average. Although ground-truth trial-averaged neural states are also unavailable, they can be approximated by PSTHs (trial-averaged smoothed spikes for each condition). Thus, trial-averaged neural state estimates should resemble PSTHs from the same condition, an attribute assessed by PSTH R2. MINT achieved higher R2 values than every other method on every dataset (Fig. 5b). This result relates to the fact that MINT estimates neural states near idealized trajectories ‘built’ from trial-averaged rates. Yet strong performance is not a trivial consequence of MINT’s design; PSTH R2 will be high only if condition (c), index (k), and the interpolation parameter (α) are accurately estimated.

Second, in sensorimotor areas during motor tasks, behavior should be decodable from the neural state. The benchmark thus measured R2 for a linear decode of velocity from the neural state. MINT outperformed all baseline approaches (Fig. 5c) for all datasets. (MINT could of course have achieved even better decoding using its typical direct association, but that would have undermined the utility of this comparison).

In addition to the five baseline methods, the Neural Latents Benchmark solicited submissions from other methods and many were submitted. These submissions spanned a broad range in terms of performance (gray vertical lines in Fig. 5), highlighting the challenging nature of this problem. Some submissions involved ensemble approaches, some of which outperformed both MINT and the baseline methods in terms of bits per spike. For the primary datasets, the highest bits per spike overall was achieved by an ensemble of SpatioTemporal Neural Data Transformers [80], achieving .386 bits per spike on the MC_Maze dataset, compared to .330 for MINT. For the smaller secondary datasets, the very best submissions provided minimal-to-modest improvement over MINT in terms of bits per spike (Fig. 5a, right), and the majority performed less well.

The ability of some expressive neural networks (and some ensembles of these networks) to improve the bits per spike metric indicates there exists variability in the neural recordings that MINT does not capture. That variability could arise from behaviorally irrelevant fluctuations in the neural state, from synchrony based effects, or from instabilities in the neural recordings. These are ‘real’ things and it is thus valid for a method to leverage them to predict spiking when that is the goal. Yet as noted above, they may often be incidental to one’s goals in estimating the neural state, highlighting the need to consider additional attributes such as PSTH R2 and velocity R2. For PSTH R2, MINT outperformed all other methods for all six datasets. For velocity R2, MINT had the best performance for three out of six datasets and was very close to the highest R2 values for the remaining three datasets (the largest deficit was ΔR2 = .012) (Fig. 5c).

In summary, MINT’s strong decode performance (Fig. 4) does indeed follow from good neural-state estimation. MINT provided estimates that were competitive with the best present methods. This includes existing well-documented methods, as well as a bevy of additional submissions. This implies that MINT could be used in situations where neural-state estimation is the goal, possibly including monitoring of the neural state for clinical purposes. MINT uses simple operations and is readily implemented causally – indeed this is typically how we would expect to see it used. Yet MINT’s performance is typically similar to – and often better than – that of complex methods that would likely have to be limited to offline situations.

Practical considerations

Training & execution times

MINT is typically very fast to train (Table 1), on the order of seconds using generic hardware (no GPUs). Fast training is a consequence of the simple operations involved in constructing the library of neural-state trajectories: principally filtering of spikes and averaging across trials. Thus, MINT can be trained extremely quickly. At the same time we stress that MINT is a method for leveraging a trajectory library, not a method for constructing it. One may sometimes wish to use alternatives to trial-averaging, either of necessity or because they improve trajectory estimates. For example, for the MC_RTT task we used AutoLFADS to infer the library and training was consequently much slower (hours rather than seconds). Training time could be reduced back to seconds using a different approach – grouping into pseudo-conditions – but performance was then reduced by a modest but meaningful amount. Thus, training will typically be very fast, but one may choose time-consuming methods when appropriate.

Execution times were well below the threshold needed for real-time applications (Table 1), even using a laptop computer. State estimation in MINT is 𝒪 (NC) when making the simplifying assumption that all conditions are of equal length. Thus, the computational burden grows only linearly with neurons or with conditions. Additionally, much of the estimation procedure in MINT is condition-specific and therefore parallelizable. This makes MINT a good candidate for decoding in BCI applications that involve large neuron counts and/or large behavioral repertoires.

Performance on small training sets

To investigate robustness to training with small trial-counts, we leveraged the fact that the four maze-reaching datasets – one primary (MC_Maze), and three secondary (MC_Maze-L, MC_Maze-M, and MC_Maze-S) – contained training sets of decreasing size. For comparison, we chose the GRU and the feedforward neural network because of their generally good position- and velocity-decoding performance (Fig. 4).

On the largest training set, MINT and the GRU achieved similar performance for position decoding (Fig. 6a, R2 = .963 and R2 = .962, respectively). The performance gap widened as training sets became smaller (R2 = .926 versus R2 = .880 for the smallest dataset). Similar results were found for velocity decoding: the performance gap between MINT and the GRU widened for smaller training sets. Thus, MINT was quite robust to small training sets: an approximately five-fold reduction in training data led to a decline in R2 of only .037 (position) and .050 (velocity). The feedforward network did not show as consistent a performance rolloff with number of training trials, but also underperformed overall.

Decoding robustness in the face of small training sets and neuron counts.

a) R2 values for MINT and two neural network decoders (GRU and the feedforward network) when decoding position and velocity from four maze datasets with progressively fewer training trials per condition. b) R2 values for MINT and the Wiener filter when decoding position and velocity from MC_Maze-L with progressively fewer neurons. ‘Retrained’ results (solid lines) show mean performance (across 50 random drops at each neuron count) when the decoders are trained and tested on reduced neuron sets. ‘Zeroed’ results (dashed lines) show mean performance when the decoders are trained on all 162 neurons and, during testing, the ‘dropped’ neurons’ spike counts are set to zero without retraining the decoder. Shaded regions depict standard errors across repeated droppings with different neurons.

Recording instabilities

With respect to recording quality, one desires two types of robustness. First, allowing for retraining, decoding should be robust to the across-session loss of neurons. To simulate such loss, we performed a neuron dropping analysis on the MC_Maze-L dataset (Fig. 6b, solid lines). Training was repeated after neurons were dropped. Average R2 for position decoding remained above 95% of peak performance until fewer than 58 neurons (of 162) remained. Results were similar for velocity decoding: performance was >95% of its peak until fewer than 62 neurons remained. We repeated this analysis for the Wiener filter. Despite its overall lower performance, the Wiener filter provides a useful comparison because (like MINT) it is easy to retrain quickly, with reliable results, across the many differently sized training sets examined here. The Wiener filter was also fairly robust, but less so than MINT. To remain above 95% of peak performance, the Wiener filter needed approximately twice as many remaining neurons: 118 (Wiener filter) versus 58 (MINT) for position and 126 (Wiener filter) versus 62 (MINT) for velocity.

Second, because there can be occasional within-session loss of isolations or individual channels, one desires robustness to modest neuron loss even without retraining. We simulated within-session neuron loss by eliminating all spikes from a subset of neurons (Fig. 6b, dashed lines) with no retraining. MINT’s performance was robust to modest neuron loss. Losing fewer than 20 neurons caused essentially no performance loss. Position and velocity decoding remained above 95% of peak performance until fewer than 113 and 109 neurons remained, respectively. In contrast, the Wiener filter had a more linear performance roll off.

Thus, MINT is robust to modest neuron-loss without retraining. At the same time, we stress that ‘retraining’ MINT to account for neuron loss is incredibly simple – it simply involves ignoring that neuron when computing spike likelihoods. Unlike nearly all other methods (including the Wiener filter) the loss of one neuron does not require changing parameters that apply to other neurons. Thus, if neuron loss is detected, retraining can occur on the fly with no need to pause decoding.

Interpretability

One form of interpretability relates to whether an algorithm uses ‘black box’ optimization to learn a solution. Neural networks can be highly expressive and learn constraints directly from data. This approach is useful when appropriate constraints are unclear. However, when constraints are known, one can rely more on model assumptions and less on expressivity, yielding robust methods that perform accurately with less training data. MINT is an interpretable algorithm that relies on strong assumptions.

The performance of interpretable methods can reflect the extent to which their assumptions match the statistics of the data. Consider the Kalman filter – an interpretable algorithm – and its performance on the MC_Maze reaching dataset. The Kalman filter produces a relatively poor decode (R2 = .609 for hand velocity). This doesn’t imply the Kalman filter is a poor algorithm, simply that its assumptions may not be well-aligned with the data. We confirmed this by generating synthetic data whose properties are closer to those assumed by the Kalman filter. The Kalman filter then performed quite competitively (Fig. S4). While expected, this result underscores an important point. MINT does not outperform the Kalman filter because it is intrinsically a better method. Instead, MINT performs well because its assumptions, guided by an understanding of neural geometry, are presumably a good match to the underlying properties of the data.

MINT is also interpretable regarding why each prediction is made. Suppose that, at a given moment during decoding, MINT selects neural state as the most likely state, and thus renders its decode using the associated behavioral state . Selection of comes with an associated likelihood that conveys information regarding how well the spiking data matched that neural state. Likelihoods are known not only for this ‘selected’ state, but for all non-selected states on all trajectories in the library. These likelihoods can be analyzed to understand why the selected state was chosen over the others, how close the algorithm was to selecting a different state, and which neurons contributed most strongly to state selection. Because of its probability-based approach, MINT is readily modifiable and extensible in a variety of ways (Fig. S5).

Discussion

Implications regarding neural geometry

All neural decode algorithms and state estimators make assumptions regarding the structure of neural activity and/or its relationship to behavior. Correctly specifying these assumptions can yield accurate, interpretable methods. MINT is the outcome of our endeavor to codify observations regarding motor-cortex population geometry into statistical assumptions for a decoder. Some observations are recent and some are documented here for the first time.

Low trajectory tangling [64] creates sparsely distributed trajectories whose geometry differs from behavioral trajectories, both muscle and kinematic. In particular, distance is poorly preserved between neural and behavioral trajectories. Neighboring states in behavioral space can be quite distant in neural space. These observations challenge two common assumptions regarding motor cortex activity. The first is that there exist strong isometries between neural activity and to-be-decoded parameters. Traditional tuning-centric hypotheses embody this assumption. Yet even as the field has moved towards factor-dynamics based hypotheses, it has remained reasonable to assume strong isometries in a subset of high-variance dimensions. Yet as the model spiking network in Figure 3b illustrates, it is also likely that motor-cortex outputs are encoded in low-variance dimensions. If so, isometry-based decoding becomes limited to low-variance dimensions, or to isometries that are incidental and task-specific [20]. MINT avoids the need to make assumptions regarding isometries by using a direct decode.

The second common assumption is that the range of observable neural states can be approximately described by a covariance matrix, and/or by the associated low-dimensional subspace obtained from a subset of its eigenvectors (the principal components). There has always been some tension associated with this assumption. Dimensionality reduction has been – and almost certainly will continue to be – a powerful analysis tool. Competing hypotheses often make contrasting predictions regarding what occurs in high-variance dimensions. Dimensionality reduction methods are ideally suited for testing those predictions. At the same time, it is unclear when and whether motor-cortex trajectories are themselves low-dimensional. Even in studies that explicitly consider an occupied subspace, that subspace is high-dimensional relative to traditional assumptions [81], and learning can ‘open up’ new dimensions [82]. Studies have also stressed the possibility of a nonlinear manifold embedded in a higher-dimensional subspace [83–86]. Nonlinear dimensionality reduction techniques can reveal a manifold’s topology or how the manifold relates to external behavioral variables (e.g. [51]). Multiple studies have stressed the remarkably high-dimensional nature of motor cortex activity [67, 68, 87, 88]. Network models of motor cortex activity [89, 90] have a high ratio of across-neuron dimensionality to across-condition dimensionality [87]. Networks that produce extended temporally structured outputs can leverage trajectories whose geometry resembles complex and well-separated ‘tubes’ embedded in a very high-dimensional space [91].

Present results – both analysis of data and the successful decode provided by MINT – support the idea that ‘subspace dimensionality’ may not be the best way to describe the constraints governing neural trajectories. Low trajectory tangling can, depending on the circumstances, be compatible with either low or high dimensionality. What is more likely to be consistent are properties of the not-easily-described nonlinear manifold, where large ‘voids’ (e.g. the middle of circular trajectories) remain unoccupied [64, 65]. Precisely because this geometry is difficult to describe with a model, MINT approximates the manifold using the empirical trajectories as a ‘scaffolding’, with interpolation producing a mesh of states that are considered during decoding.

Embracing these aspects of trajectory geometry produced excellent performance: MINT accurately estimated the neural state and decoded a broad range of behavioral variables across a variety of tasks and brain areas. Good decode performance, on its own, is insufficient to validate the assumptions of a decode method and we stress caution in that regard. This is particularly true because many of the scientific findings that inspired MINT are not, strictly speaking, assumptions of the method itself. MINT is designed to work well when trajectories are sparsely distributed, with low tangling, and spread out in a high-dimensional space. Yet MINT can perform well in the absence of these features. For example, MINT performed well on simulated data that was instead well-matched to the assumptions of a Kalman filter (MINT and the Kalman filter performed comparably well in that case).

Thus, good decode performance does not necessarily imply MINT’s assumptions are uniquely correct. Nevertheless, comparisons can aid interpretation. For the empirical data, MINT consistently outperformed the Kalman filter, often by a wide margin. This indicates that the data contain properties that MINT, but not the Kalman filter, is poised to leverage. Given the analysis in Figure 1, some of the most likely properties are those that motivated MINT’s design: sparsely distributed low-tangled trajectories. MINT’s assumption of Poisson spiking statistics may also have contributed, yet MINT had a much smaller advantage for simulated data where spiking was Poisson but neural geometry matched the Kalman filter’s assumptions Fig. S4. Thus, it is likely that MINT gains much of its advantage via constraints on the neural-state geometry that are more accurate than those imposed by the Kalman filter. Comparison with expressive machine-learning approaches makes a similar point from the other direction. Highly expressive methods, through training, should be able to learn and leverage a range of neural constraints. Given that MINT was competitive with these methods, its assumptions are likely to be a good match to the statistics of the data.

Generalization

MINT readily performs certain types of generalization. Interpolation allows MINT to generalize spatially, to neural states (and corresponding behavioral states) that are between trajectories observed during training. MINT can also generalize temporally, by selecting mostly-likely states that move faster (or slower) in neural state space than those present during training. MINT can also generalize compositionally: switching between trajectories or exiting them early. As a simple example, a library consisting of a single 4-cycle trajectory from the cycling task would allow MINT to decode a 2-cycle movement or a 200-cycle movement. Thus, MINT is not stuck on the library trajectories, nor is it stuck traversing trajectories at fixed speeds or for fixed durations.

Yet while MINT can follow trajectories through the mesh that are quite different from those in the library, all decoded states will be on the mesh. MINT is not capable of generalizing to trajectories that traverse regions of state space far from the mesh. This is relevant, because empirical evidence indicates that ‘new’ behaviors can sometimes be associated with neural trajectories that are very different from all previously observed trajectories – often occupying orthogonal subspaces [64, 92–94]. MINT could be modified to allow extrapolation outside the library, but we suspect that this form of generalization will still not typically be available. As illustrated in Figure S2, it will be extremely difficult for any decoder to successfully decode neural trajectories that are orthogonal to those in the training set. For example, none of the methods we evaluated were able to generalize from forward to backward cycling, which were associated with nearly orthogonal neural trajectories. However, MINT’s computation of log-likelihoods does lend it a potential advantage in handling novel or unexpected behaviors. Using the training set, one can roughly estimate the distribution of log-likelihoods. If, during decoding, log-likelihoods become vastly lower than expected, that suggests decoding is struggling with a movement too far from the training set. This knowledge would give the technician options, including decoding zero movement (on the grounds that no movement is better than a large error) or pausing to retrain. The latter could leverage another advantage of MINT: adding a trajectory to the library – and thus expanding the mesh – is simple and no other retraining or modification of parameters is needed.

Multi-task decoding

Across-task generalization may often be limited by the above considerations. Yet this does not imply that MINT cannot decode in a multi-task regime. On the contrary, MINT is ideally suited to decoding from a behavioral repertoire spanning two or more tasks – each task simply needs to be represented in the mesh with an appropriate set of trajectories. And as noted above, trajectories corresponding to a new task can be easily added to the library.

MINT’s facility with multi-task decoding is particularly relevant because, historically, decoding approaches have had to vary across tasks. For example, linear decoding of velocity works well during reaching but not cycling [20]. Nevertheless, Schroeder et al. [20] were able to achieve accurate closed-loop decoding of cycling by nonlinearly leveraging statistical regularities in a task-specific neural subspace. MINT embraces both task-specificity and unification. Each task contributes a unique set of trajectories to the mesh. A unified decoding strategy then applies both within and across tasks. This avoids the need to engineer bespoke decoders for each task individually and facilitates multi-task decoding that doesn’t rely on an additional meta-decoder to identify the appropriate task-specific decoder at each moment.

Future avenues

MINT’s probability-based framework is amenable to principled extensions. Rather than simply decode the state with the highest data likelihood, one could incorporate priors. Priors could be fixed or guided by incoming information (e.g. an eye tracker). Decoding could focus on utility functions rather than probabilities; a patient may value avoiding certain types of errors more than others. For example, erroneous transitions from stationary to moving may be jarring or unsafe, and could be dissuaded via an appropriately designed utility function.

As noted above, MINT is amenable to a variety of methods – simple or sophisticated – for inferring the trajectory library. Possible future approaches include leveraging the typical neural geometry for a particular task to allow a patient-specific mesh to be learned using fewer trials. This might even be done in an unsupervised way (similar to Dyer et al. [95]): a trajectory library could be built without behavioral labels, then appropriately linked to a ‘reference’ behavioral library via prior knowledge of task-specific neural geometry. As a simple example, cycling across different speeds produces a tube-shaped manifold with different velocities at each end [65]. An empirically observed tube might be given behavioral labels based on this knowledge. These observations may be relevant to a primary goal of MINT: potential use as an online decoder for real-time closed-loop performance. In situations involving paralyzed patients, training may involve asking subjects to observe and/or attempt movements, both of which engage neurons in motor cortex [96], but may create uncertainty regarding the exact behavioral label for each moment. The above strategies may aid in this regard. The approach of Gilja et al. [13] would also likely be helpful. They found that re-fitting Kalman filter parameters online improved closed-loop performance. One could similarly use an initial mesh to allow decoding to gain traction, then ‘re-mesh’ MINT online to improve performance.

Supporting information

eLife Assessment

This paper presents a new method called MINT that is simple yet effective at BCI-style decoding tasks in stereotyped settings. While the reviewers raise caveats, overall they believe the work is a valuable study for the field of motor control, and the evidence to support their claims is solid.

Reviewer #1 (Public Review):

Summary:

This paper presents an innovative decoding approach for brain-computer interfaces (BCIs), introducing a new method named MINT. The authors develop a trajectory-centric approach to decode behaviors across several different datasets, including eight empirical datasets from the Neural Latents Benchmark. Overall, the paper is well written and their method shows impressive performance compared to more traditional decoding approaches that use a simpler approach. While there are some concerns (see below), the paper's strengths, particularly its emphasis on a trajectory-centric approach and the simplicity of MINT, provide a compelling contribution to the field.

Strengths:

The adoption of a trajectory-centric approach that utilizes statistical constraints presents a substantial shift in methodology, potentially revolutionizing the way BCIs interpret and predict neural behaviour. This is one of the strongest aspects of the paper.

The thorough evaluation of the method across various datasets serves as an assurance that the superior performance of MINT is not a result of overfitting. The comparative simplicity of the method in contrast to many neural network approaches is refreshing and should facilitate broader applicability.

Weaknesses:

Scope: Despite the impressive performance of MINT across multiple datasets, it seems predominantly applicable to M1/S1 data. Only one of the eight empirical datasets comes from an area outside the motor/somatosensory cortex. It would be beneficial if the authors could expand further on how the method might perform with other brain regions that do not exhibit low tangling or do not have a clear trial structure (e.g. decoding of position or head direction from hippocampus)

When comparing methods, the neural trajectories of MINT are based on averaged trials, while the comparison methods are trained on single trials. An additional analysis might help in disentangling the effect of the trial averaging. For this, the authors could average the input across trials for all decoders, establishing a baseline for averaged trials. Note that inference should still be done on single trials. Performance can then be visualized across different values of N, which denotes the number of averaged trials used for training.

Reviewer #2 (Public Review):

Summary:

The goal of this paper is to present a new method, termed MINT, for decoding behavioral states from neural spiking data. MINT is a statistical method which, in addition to outputting a decoded behavioral state, also provides soft information regarding the likelihood of that behavioral state based on the neural data. The innovation in this approach is neural states are assumed to come from sparsely distributed neural trajectories with low tangling, meaning that neural trajectories (time sequences of neural states) are sparse in the high-dimensional space of neural spiking activity and that two dissimilar neural trajectories tend to correspond to dissimilar behavioral trajectories. The authors support these assumptions through analysis of previously collected data, and then validate the performance of their method by comparing it to a suite of alternative approaches. The authors attribute the typically improved decoding performance by MINT to its assumptions being more faithfully aligned to the properties of neural spiking data relative to assumptions made by the alternatives.

Strengths:

The paper did an excellent job critically evaluating common assumptions made by neural analytical methods, such as neural state being low-dimensional relative to the number of recorded neurons. The authors made strong arguments, supported by evidence and literature, for potentially high-dimensional neural states and thus the need for approaches that do not rely on an assumption of low dimensionality.

The paper was thorough in considering multiple datasets across a variety of behaviors, as well as existing decoding methods, to benchmark the MINT approach. This provided a valuable comparison to validate the method. The authors also provided nice intuition regarding why MINT may offer performance improvement in some cases and in which instances MINT may not perform as well.

In addition to providing a philosophical discussion as to the advantages of MINT and benchmarking against alternatives, the authors also provided a detailed description of practical considerations. This included training time, amount of training data, robustness to data loss or changes in the data, and interpretability. These considerations not only provided objective evaluation of practical aspects but also provided insights to the flexibility and robustness of the method as they relate back to the underlying assumptions and construction of the approach.

Weaknesses:

The authors posit that neural and behavioral trajectories are non-isometric. To support this point, they look at distances between neural states and distances between the corresponding behavioral states, in order to demonstrate that there are differences in these distances in each respective space. This supports the idea that neural states and behavioral states are non-isometric but does not directly address their point. In order to say the trajectories are non-isometric, it would be better to look at pairs of distances between corresponding trajectories in each space.

With regards to the idea of neural and behavioral trajectories having different geometries, this is dependent on what behavioral variables are selected. In the example for Fig 2a, the behavior is reach position. The geometry of the behavioral trajectory of interest would look different if instead the behavior of interest was reach velocity. The paper would be strengthened by acknowledgement that geometries of trajectories are shaped by extrinsic choices rather than (or as much as they are) intrinsic properties of the data.

The approach is built up on the idea of creating a "mesh" structure of possible states. In the body of the paper the definition of the mesh was not entirely clear and I could not find in the methods a more rigorous explicit definition. Since the mesh is integral to the approach, the paper would be improved with more description of this component.

Impact:

This work is motivated by brain-computer interfaces applications, which it will surely impact in terms of neural decoder design. However, this work is also broadly impactful for neuroscientific analysis to relate neural spiking activity to observable behavioral features. Thus, MINT will likely impact neuroscience research generally. The methods are made publicly available, and the datasets used are all in public repositories, which facilitates adoption and validation of this method within the greater scientific community.

Reviewer #3 (Public Review):

Summary:

This manuscript develops a new method termed MINT for decoding of behavior. The method is essentially a table-lookup rather than a model. Within a given stereotyped task, MINT tabulates averaged firing rate trajectories of neurons (neural states) and corresponding averaged behavioral trajectories as stereotypes to construct a library. For a test trial with a realized neural trajectory, it then finds the closest neural trajectory to it in the table and declares the associated behavior trajectory in the table as the decoded behavior. The method can also interpolate between these tabulated trajectories. The authors mention that the method is based on three key assumptions: (1) Neural states may not be embedded in a low-dimensional subspace, but rather in a high-dimensional space. (2) Neural trajectories are sparsely distributed under different behavioral conditions. (3) These neural states traverse trajectories in a stereotyped order.

The authors conducted multiple analyses to validate MINT, demonstrating its decoding of behavioral trajectories in simulations and datasets (Figures 3, 4). The main behavior decoding comparison is shown in Figure 4. In stereotyped tasks, decoding performance is comparable (M_Cycle, MC_Maze) or better (Area 2_Bump) than other linear/nonlinear algorithms (Figure 4). However, MINT underperforms for the MC_RTT task, which is less stereotyped (Figure 4).

This paper is well-structured and its main idea is clear. The fact that performance on stereotyped tasks is high is interesting and informative, showing that these stereotyped tasks create stereotyped neural trajectories. The task-specific comparisons include various measures and a variety of common decoding approaches, which is a strength. However, I have several major concerns. I believe several of the conclusions in the paper, which are also emphasized in the abstract, are not accurate or supported, especially about generalization, computational scalability, and utility for BCIs. MINT is essentially a table-lookup algorithm based on stereotyped task-dependent trajectories and involves the tabulation of extensive data to build a vast library without modeling. These aspects will limit MINT's utility for real-world BCIs and tasks. These properties will also limit MINT's generalizability from task to task, which is important for BCIs and thus is commonly demonstrated in BCI experiments with other decoders without any retraining. Furthermore, MINT's computational and memory requirements can be prohibitive it seems. Finally, as MINT is based on tabulating data without learning models of data, I am unclear how it will be useful in basic investigations of neural computations. I expand on these concerns below.

Main comments:

1. MINT does not generalize to different tasks, which is a main limitation for BCI utility compared with prior BCI decoders that have shown this generalizability as I review below. Specifically, given that MINT tabulates task-specific trajectories, it will not generalize to tasks that are not seen in the training data even when these tasks cover the exact same space (e.g., the same 2D computer screen and associated neural space).

First, the authors provide a section on generalization, which is inaccurate because it mixes up two fundamentally different concepts: 1) collecting informative training data and 2) generalizing from task to task. The former is critical for any algorithm, but it does not imply the latter. For example, removing one direction of cycling from the training set as the authors do here is an example of generating poor training data because the two behavioral (and neural) directions are non-overlapping and/or orthogonal while being in the same space. As such, it is fully expected that all methods will fail. For proper training, the training data should explore the whole movement space and the associated neural space, but this does not mean all kinds of tasks performed in that space must be included in the training set (something MINT likely needs while modeling-based approaches do not). Many BCI studies have indeed shown this generalization ability using a model. For example, in Weiss et al. 2019, center-out reaching tasks are used for training and then the same trained decoder is used for typing on a keyboard or drawing on the 2D screen. In Gilja et al. 2012, training is on a center-out task but the same trained decoder generalizes to a completely different pinball task (hit four consecutive targets) and tasks requiring the avoidance of obstacles and curved movements. There are many more BCI studies, such as Jarosiewicz et al. 2015 that also show generalization to complex real-world tasks not included in the training set. Unlike MINT, these works can achieve generalization because they model the neural subspace and its association to movement. On the contrary, MINT models task-dependent neural trajectories, so the trained decoder is very task-dependent and cannot generalize to other tasks. So, unlike these prior BCIs methods, MINT will likely actually need to include every task in its library, which is not practical.

I suggest the authors remove claims of generalization and modify their arguments throughout the text and abstract. The generalization section needs to be substantially edited to clarify the above points. Please also provide the BCI citations and discuss the above limitation of MINT for BCIs.

2. MINT is shown to achieve competitive/high performance in highly stereotyped datasets with structured trials, but worse performance on MC_RTT, which is not based on repeated trials and is less stereotyped. This shows that MINT is valuable for decoding in repetitive stereotyped use-cases. However, it also highlights a limitation of MINT for BCIs, which is that MINT may not work well for real-world and/or less-constrained setups such as typing, moving a robotic arm in 3D space, etc. This is again due to MINT being a lookup table with a library of stereotyped trajectories rather than a model. Indeed, the authors acknowledge that the lower performance on MC_RTT (Figure 4) may be caused by the lack of repeated trials of the same type. However, real-world BCI decoding scenarios will also not have such stereotyped trial structure and will be less/un-constrained, in which MINT underperforms. Thus, the claim in the abstract or lines 480-481 that MINT is an "excellent" candidate for clinical BCI applications is not accurate and needs to be qualified. The authors should revise their statements according and discuss this issue. They should also make the use-case of MINT on BCI decoding clearer and more convincing.

3. Related to 2, it may also be that MINT achieves competitive performance in offline and trial-based stereotyped decoding by overfitting to the trial structure in a given task, and thus may not generalize well to online performance due to overfitting. For example, a recent work showed that offline decoding performance may be overfitted to the task structure and may not represent online performance (Deo et al. 2023). Please discuss.

4. Related to 2, since MINT requires firing rates to generate the library and simple averaging does not work for this purpose in the MC_RTT dataset (that does not have repeated trials), the authors needed to use AutoLFADS to infer the underlying firing rates. The fact that MINT requires the usage of another model to be constructed first and that this model can be computationally complex, will also be a limiting factor and should be clarified.

5. I also find the statement in the abstract and paper that "computations are simple, scalable" to be inaccurate. The authors state that MINT's computational cost is O(NC) only, but it seems this is achieved at a high memory cost as well as computational cost in training. The process is described in section "Lookup table of log-likelihoods" on line [978-990]. The idea is to precompute the log-likelihoods for any combination of all neurons with discretization x all delay/history segments x all conditions and to build a large lookup table for decoding. Basically, the computational cost of precomputing this table is O(V^{Nτ} x TC) and the table requires a memory of O(V^{Nτ}), where V is the number of discretization points for the neural firing rates, N is the number of neurons, τ is the history length, T is the trial length, and C is the number of conditions. This is a very large burden, especially the V^{Nτ} term. This cost is currently not mentioned in the manuscript and should be clarified in the main text. Accordingly, computation claims should be modified including in the abstract.

6. In addition to the above technical concerns, I also believe the authors should clarify the logic behind developing MINT better. From a scientific standpoint, we seek to gain insights into neural computations by making various assumptions and building models that parsimoniously describe the vast amount of neural data rather than simply tabulating the data. For instance, low-dimensional assumptions have led to the development of numerous dimensionality reduction algorithms and these models have led to important interpretations about the underlying dynamics (e.g., fixed points/limit cycles). While it is of course valid and even insightful to propose different assumptions from existing models as the authors do here, they do not actually translate these assumptions into a new model. Without a model and by just tabulating the data, I don't believe we can provide interpretation or advance the understanding of the fundamentals behind neural computations. As such, I am not clear as to how this library building approach can advance neuroscience or how these assumptions are useful. I think the authors should clarify and discuss this point.

7. Related to 6, there seems to be a logical inconsistency between the operations of MINT and one of its three assumptions, namely, sparsity. The authors state that neural states are sparsely distributed in some neural dimensions (Figure 1a, bottom). If this is the case, then why does MINT extend its decoding scope by interpolating known neural states (and behavior) in the training library? This interpolation suggests that the neural states are dense on the manifold rather than sparse, thus being contradictory to the assumption made. If interpolation-based dense meshes/manifolds underlie the data, then why not model the neural states through the subspace or manifold representations? I think the authors should address this logical inconsistency in MINT, especially since this sparsity assumption also questions the low-dimensional subspace/manifold assumption that is commonly made.

References

Weiss, Jeffrey M., Robert A. Gaunt, Robert Franklin, Michael L. Boninger, and Jennifer L. Collinger. 2019. "Demonstration of a Portable Intracortical Brain-Computer Interface." Brain-Computer Interfaces 6 (4): 106-17. https://doi.org/10.1080/2326263X.2019.1709260.

Gilja, Vikash, Paul Nuyujukian, Cindy A. Chestek, John P. Cunningham, Byron M. Yu, Joline M. Fan, Mark M. Churchland, et al. 2012. "A High-Performance Neural Prosthesis Enabled by Control Algorithm Design." Nature Neuroscience 15 (12): 1752-1757. https://doi.org/10.1038/nn.3265.

Jarosiewicz, Beata, Anish A. Sarma, Daniel Bacher, Nicolas Y. Masse, John D. Simeral, Brittany Sorice, Erin M. Oakley, et al. 2015. "Virtual Typing by People with Tetraplegia Using a Self-Calibrating Intracortical Brain-Computer Interface." Science Translational Medicine 7 (313): 313ra179-313ra179. https://doi.org/10.1126/scitranslmed.aac7328.

Darrel R. Deo, Francis R. Willett, Donald T. Avansino, Leigh R. Hochberg, Jaimie M. Henderson, and Krishna V. Shenoy. 2023. "Translating Deep Learning to Neuroprosthetic Control." BioRxiv, 2023.04.21.537581. https://doi.org/10.1101/2023.04.21.537581.

Author response:

The following is the authors’ response to the original reviews.

Summary of reviewers’ comments and our revisions:

We thank the reviewers for their thoughtful feedback. This feedback has motivated multiple revisions and additions that, in our view, have greatly improved the manuscript. This is especially true with regard to a major goal of this study: clearly defining existing scientific perspectives and delineating their decoding implications. In addition to building on this conceptual goal, we have expanded existing analyses and have added a new analysis of generalization using a newly collected dataset. We expect the manuscript will be of very broad interest, both to those interested in BCI development and to those interested in fundamental properties of neural population activity and its relationship with behavior.

Importantly, all reviewers were convinced that MINT provided excellent performance, when benchmarked against existing methods, across a broad range of standard tasks:

“their method shows impressive performance compared to more traditional decoding approaches” (R1)

“The paper was thorough in considering multiple datasets across a variety of behaviors, as well as existing decoding methods, to benchmark the MINT approach. This provided a valuable comparison to validate the method.” (R2)

“The fact that performance on stereotyped tasks is high is interesting and informative…” (R3)

This is important. It is challenging to design a decoder that performs consistently across multiple domains and across multiple situations (including both decoding and neural state estimation). MINT does so. MINT consistently outperformed existing lightweight ‘interpretable’ decoders, despite being a lightweight interpretable decoder itself. MINT was very competitive with expressive machine-learning methods, yet has advantages in flexibility and simplicity that more ‘brute force’ methods do not. We made a great many comparisons, and MINT was consistently a strong performer. Of the many comparisons we made, there was only one where MINT was at a modest disadvantage, and it was for a dataset where all methods performed poorly. No other method we tested was as consistent. For example, although the GRU and the feedforward network were often competitive with MINT (and better than MINT in the one case mentioned above), there were multiple other situations where they performed less well and a few situations where they performed poorly. Moreover, no other existing decoder naturally estimates the neural state while also readily decoding, without retraining, a broad range of behavioral variables.

R1 and R2 were very positive about the broader impacts of the study. They stressed its impact both on decoder design, and on how our field thinks, scientifically, about the population response in motor areas:

“This paper presents an innovative decoding approach for brain-computer interfaces” (R1)

“presents a substantial shift in methodology, potentially revolutionizing the way BCIs interpret and predict neural behaviour” (R1)

“the paper's strengths, particularly its emphasis on a trajectory-centric approach and the simplicity of MINT, provide a compelling contribution to the field” (R1)

“The authors made strong arguments, supported by evidence and literature, for potentially high-dimensional neural states and thus the need for approaches that do not rely on an assumption of low dimensionality” (R2)

“This work is motivated by brain-computer interfaces applications, which it will surely impact in terms of neural decoder design.” (R2)

“this work is also broadly impactful for neuroscientific analysis... Thus, MINT will likely impact neuroscience research generally.” (R2)

We agree with these assessments, and have made multiple revisions to further play into these strengths. As one example, the addition of Figure 1b (and 6b) makes this the first study, to our knowledge, to fully and concretely illustrate this emerging scientific perspective and its decoding implications. This is important, because multiple observations convince us that the field is likely to move away from the traditional perspective in Figure 1a, and towards that in Figure 1b. We also agree with the handful of weaknesses R1 and R2 noted. The manuscript has been revised accordingly. The major weakness noted by R1 was the need to be explicit regarding when we suspect MINT would (and wouldn’t) work well in other brain areas. In non-motor areas, the structure of the data may be poorly matched with MINT’s assumptions. We agree that this is likely to be true, and thus agree with the importance of clarifying this topic for the reader. The revision now does so. R1 also wished to know whether existing methods might benefit from including trial-averaged data during training, something we now explore and document (see detailed responses below). R2 noted two weaknesses: 1) The need to better support (with expanded analysis) the statement that neural and behavioral trajectories are non-isometric, and 2) The need to more rigorously define the ‘mesh’. We agree entirely with both suggestions, and the revision has been strengthened by following them (see detailed responses below).

R3 also saw strengths to the work, stating that:

“This paper is well-structured and its main idea is clear.”

“The fact that performance on stereotyped tasks is high is interesting and informative, showing that these stereotyped tasks create stereotyped neural trajectories.”

“The task-specific comparisons include various measures and a variety of common decoding approaches, which is a strength.”

However, R3 also expressed two sizable concerns. The first is that MINT might have onerous memory requirements. The manuscript now clarifies that MINT has modest memory requirements. These do not scale unfavorably as the reviewer was concerned they might. The second concern is that MINT is:

“essentially a table-lookup rather than a model.”

Although we don’t agree, the concern makes sense and may be shared by many readers, especially those who take a particular scientific perspective. Pondering this concern thus gave us the opportunity to modify the manuscript in ways that support its broader impact. Our revisions had two goals: 1) clarify the ways in which MINT is far more flexible than a lookup-table, and 2) better describe the dominant scientific perspectives and their decoding implications.

The heart of R3’s concern is the opinion that MINT is an effective but unprincipled hack suitable for situations where movements are reasonably stereotyped. Of course, many tasks involve stereotyped movements (e.g. handwriting characters), so MINT would still be useful. Nevertheless, if MINT is not principled, other decode methods would often be preferable because they could (unlike MINT in R3’s opinion) gain flexibility by leveraging an accurate model. Most of R3’s comments flow from this fundamental concern:

“This is again due to MINT being a lookup table with a library of stereotyped trajectories rather than a model.”

“MINT models task-dependent neural trajectories, so the trained decoder is very task-dependent and cannot generalize to other tasks.”

“Unlike MINT, these works can achieve generalization because they model the neural subspace and its association to movement.”

“given that MINT tabulates task-specific trajectories, it will not generalize to tasks that are not seen in the training data even when these tasks cover the exact same space (e.g., the same 2D computer screen and associated neural space).”

“For proper training, the training data should explore the whole movement space and the associated neural space, but this does not mean all kinds of tasks performed in that space must be included in the training set (something MINT likely needs while modeling-based approaches do not).”

The manuscript has been revised to clarify that MINT is considerably more flexible than a lookup table, even though a lookup table is used as a first step. Yet, on its own, this does not fully address R3’s concern. The quotes above highlight that R3 is making a standard assumption in our field: that there exists a “movement space and associated neural space”. Under this perspective, one should, as R3 argues fully explore the movement space. This would perforce fully explore the associated neural subspace. One can then “model the neural subspace and its association to movement”. MINT does not use a model of this type, and thus (from R3’s perspective) does not appear to use a model at all. A major goal of our study is to question this traditional perspective. We have thus added a new figure to highlight the contrast between the traditional (Figure 1a) and new (Figure 1b) scientific perspectives, and to clarify their decoding implications.

While we favor the new perspective (Figure 1b), we concede that R3 may not share our view. This is fine. Part of the reason we believe this study is timely, and will be broadly read, is that it raises a topic of emerging interest where there is definitely room for debate. If we are misguided – i.e. if Figure 1a is the correct perspective – then many of R3’s concerns would be on target: MINT could still be useful, but traditional methods that make the traditional assumptions in Figure 1a would often be preferable. However, if the emerging perspective in Figure 1b is more accurate, then MINT’s assumptions would be better aligned with the data than those of traditional methods, making it a more (not less) principled choice.

Our study provides new evidence in support of Figure 1b, while also synthesizing existing evidence from other recent studies. In addition to Figure 2, the new analysis of generalization further supports Figure 1b. Also supporting Figure 1b is the analysis in which MINT’s decoding advantage, over a traditional decoder, disappears when simulated data approximate the traditional perspective in Figure 1a.

That said, we agree that the present study cannot fully resolve whether Figure 1a or 1b is more accurate. Doing so will take multiple studies with different approaches (indeed we are currently preparing other manuscripts on this topic). Yet we still have an informed scientific opinion, derived from past, present and yet-to-be-published observations. Our opinion is that Figure 1b is the more accurate perspective. This possibility makes it reasonable to explore the potential virtues of a decoding method whose assumptions are well-aligned with that perspective. MINT is such a method. As expected under Figure 1b, MINT outperforms traditional interpretable decoders in every single case we studied.

As noted above, we have added a new generalization-focused analysis (Figure 6) based on a newly collected dataset. We did so because R3’s comments highlight a deep point: which scientific perspective one takes has strong implications regarding decoder generalization. These implications are now illustrated in the new Figure 6a and 6b. Under Figure 6a, it is possible, as R3 suggests, to explore “the whole movement space and associated neural space” during training. However, under Figure 6b, expectations are very different. Generalization will be ‘easy’ when new trajectories are near the training-set trajectories. In this case, MINT should generalize well as should other methods. In contrast, generalization will be ‘hard’ when new neural trajectories have novel shapes and occupy previously unseen regions / dimensions. In this case, all current methods, including MINT, are likely to fail. R3 points out that traditional decoders have sometimes generalized well to new tasks (e.g. from center-out to ‘pinball’) when cursor movements occur in the same physical workspace. These findings could be taken to support Figure 6a, but are equally consistent with ‘easy’ generalization in Figure 6b. To explore this topic, the new analysis in Figure 6c-g considers conditions that are intended to span the range from easy to hard. Results are consistent with the predictions of Figure 6b.

We believe the manuscript has been significantly improved by these additions. The revisions help the manuscript achieve its twin goals: 1) introduce a novel class of decoder that performs very well despite being very simple, and 2) describe properties of motor-cortex activity that will matter for decoders of all varieties.

Reviewer #1:

Summary:

This paper presents an innovative decoding approach for brain-computer interfaces (BCIs), introducing a new method named MINT. The authors develop a trajectory-centric approach to decode behaviors across several different datasets, including eight empirical datasets from the Neural Latents Benchmark. Overall, the paper is well written and their method shows impressive performance compared to more traditional decoding approaches that use a simpler approach. While there are some concerns (see below), the paper's strengths, particularly its emphasis on a trajectory-centric approach and the simplicity of MINT, provide a compelling contribution to the field.

We thank the reviewer for these comments. We share their enthusiasm for the trajectory-centric approach, and we are in complete agreement that this perspective has both scientific and decoding implications. The revision expands upon these strengths.

Strengths:

The adoption of a trajectory-centric approach that utilizes statistical constraints presents a substantial shift in methodology, potentially revolutionizing the way BCIs interpret and predict neural behaviour. This is one of the strongest aspects of the paper.

Again, thank you. We also expect the trajectory-centric perspective to have a broad impact, given its relevance to both decoding and to thinking about manifolds.

The thorough evaluation of the method across various datasets serves as an assurance that the superior performance of MINT is not a result of overfitting. The comparative simplicity of the method in contrast to many neural network approaches is refreshing and should facilitate broader applicability.

Thank you. We were similarly pleased to see such a simple method perform so well. We also agree that, while neural-network approaches will always be important, it is desirable to also possess simple ‘interpretable’ alternatives.

Weaknesses:

Comment 1) Scope: Despite the impressive performance of MINT across multiple datasets, it seems predominantly applicable to M1/S1 data. Only one of the eight empirical datasets comes from an area outside the motor/somatosensory cortex. It would be beneficial if the authors could expand further on how the method might perform with other brain regions that do not exhibit low tangling or do not have a clear trial structure (e.g. decoding of position or head direction from hippocampus)

We agree entirely. Population activity in many brain areas (especially outside the motor system) presumably will often not have the properties upon which MINT’s assumptions are built. This doesn’t necessarily mean that MINT would perform badly. Using simulated data, we have found that MINT can perform surprisingly well even when some of its assumptions are violated. Yet at the same time, when MINT’s assumptions don’t apply, one would likely prefer to use other methods. This is, after all, one of the broader themes of the present study: it is beneficial to match decoding assumptions to empirical properties. We have thus added a section on this topic early in the Discussion:

“In contrast, MINT and the Kalman filter performed comparably on simulated data that better approximated the assumptions in Figure 1a. Thus, MINT is not a ‘better’ algorithm – simply better aligned with the empirical properties of motor cortex data. This highlights an important caveat. Although MINT performs well when decoding from motor areas, its assumptions may be a poor match in other areas (e.g. the hippocampus). MINT performed well on two non-motor-cortex datasets – Area2_Bump (S1) and DMFC_RSG (dorsomedial frontal cortex) – yet there will presumably be other brain areas and/or contexts where one would prefer a different method that makes assumptions appropriate for that area.”

Comment 2) When comparing methods, the neural trajectories of MINT are based on averaged trials, while the comparison methods are trained on single trials. An additional analysis might help in disentangling the effect of the trial averaging. For this, the authors could average the input across trials for all decoders, establishing a baseline for averaged trials. Note that inference should still be done on single trials. Performance can then be visualized across different values of N, which denotes the number of averaged trials used for training.

We explored this question and found that the non-MINT decoders are harmed, not helped, by the inclusion of trial-averaged responses in the training set. This is presumably because the statistics of trialaveraged responses don’t resemble what will be observed during decoding. This statistical mismatch, between training and decoding, hurts most methods. It doesn’t hurt MINT, because MINT doesn’t ‘train’ in the normal way. It simply needs to know rates, and trial-averaging is a natural way to obtain them. To describe the new analysis, we have added the following to the text.

“We also investigated the possibility that MINT gained its performance advantage simply by having access to trial-averaged neural trajectories during training, while all other methods were trained on single-trial data. This difference arises from the fundamental requirements of the decoder architectures: MINT needs to estimate typical trajectories while other methods don’t. Yet it might still be the case that other methods would benefit from including trial-averaged data in the training set, in addition to single-trial data. Alternatively, this might harm performance by creating a mismatch, between training and decoding, in the statistics of decoder inputs. We found that the latter was indeed the case: all non-MINT methods performed better when trained purely on single-trial data.”

Reviewer #2:

Summary:

The goal of this paper is to present a new method, termed MINT, for decoding behavioral states from neural spiking data. MINT is a statistical method which, in addition to outputting a decoded behavioral state, also provides soft information regarding the likelihood of that behavioral state based on the neural data. The innovation in this approach is neural states are assumed to come from sparsely distributed neural trajectories with low tangling, meaning that neural trajectories (time sequences of neural states) are sparse in the high-dimensional space of neural spiking activity and that two dissimilar neural trajectories tend to correspond to dissimilar behavioral trajectories. The authors support these assumptions through analysis of previously collected data, and then validate the performance of their method by comparing it to a suite of alternative approaches. The authors attribute the typically improved decoding performance by MINT to its assumptions being more faithfully aligned to the properties of neural spiking data relative to assumptions made by the alternatives.

We thank the reviewer for this accurate summary, and for highlighting the subtle but important fact that MINT provides information regarding likelihoods. The revision includes a new analysis (Figure 6e) illustrating one potential way to leverage knowledge of likelihoods.

Strengths:

The paper did an excellent job critically evaluating common assumptions made by neural analytical methods, such as neural state being low-dimensional relative to the number of recorded neurons. The authors made strong arguments, supported by evidence and literature, for potentially high-dimensional neural states and thus the need for approaches that do not rely on an assumption of low dimensionality.

Thank you. We also hope that the shift in perspective is the most important contribution of the study. This shift matters both scientifically and for decoder design. The revision expands on this strength. The scientific alternatives are now more clearly and concretely illustrated (especially see Figure 1a,b and Figure 6a,b). We also further explore their decoding implications with new data (Figure 6c-g).

The paper was thorough in considering multiple datasets across a variety of behaviors, as well as existing decoding methods, to benchmark the MINT approach. This provided a valuable comparison to validate the method. The authors also provided nice intuition regarding why MINT may offer performance improvement in some cases and in which instances MINT may not perform as well.

Thank you. We were pleased to be able to provide comparisons across so many datasets (we are grateful to the Neural Latents Benchmark for making this possible).

In addition to providing a philosophical discussion as to the advantages of MINT and benchmarking against alternatives, the authors also provided a detailed description of practical considerations. This included training time, amount of training data, robustness to data loss or changes in the data, and interpretability. These considerations not only provided objective evaluation of practical aspects but also provided insights to the flexibility and robustness of the method as they relate back to the underlying assumptions and construction of the approach.

Thank you. We are glad that these sections were appreciated. MINT’s simplicity and interpretability are indeed helpful in multiple ways, and afford opportunities for interesting future extensions. One potential benefit of interpretability is now explored in the newly added Figure 6e.

Impact:

This work is motivated by brain-computer interfaces applications, which it will surely impact in terms of neural decoder design. However, this work is also broadly impactful for neuroscientific analysis to relate neural spiking activity to observable behavioral features. Thus, MINT will likely impact neuroscience research generally. The methods are made publicly available, and the datasets used are all in public repositories, which facilitates adoption and validation of this method within the greater scientific community.

Again, thank you. We have similar hopes for this study.

Weaknesses (1 & 2 are related, and we have switched their order in addressing them):

Comment 2) With regards to the idea of neural and behavioral trajectories having different geometries, this is dependent on what behavioral variables are selected. In the example for Fig 2a, the behavior is reach position. The geometry of the behavioral trajectory of interest would look different if instead the behavior of interest was reach velocity. The paper would be strengthened by acknowledgement that geometries of trajectories are shaped by extrinsic choices rather than (or as much as they are) intrinsic properties of the data.

We agree. Indeed, we almost added a section to the original manuscript on this exact topic. We have now done so:

“A potential concern regarding the analyses in Figure 2c,d is that they require explicit choices of behavioral variables: muscle population activity in Figure 2c and angular phase and velocity in Figure 2d. Perhaps these choices were misguided. Might neural and behavioral geometries become similar if one chooses ‘the right’ set of behavioral variables? This concern relates to the venerable search for movement parameters that are reliably encoded by motor cortex activity [69, 92–95]. If one chooses the wrong set of parameters (e.g. chooses muscle activity when one should have chosen joint angles) then of course neural and behavioral geometries will appear non-isometric. There are two reasons why this ‘wrong parameter choice’ explanation is unlikely to account for the results in Figure 2c,d. First, consider the implications of the left-hand side of Figure 2d. A small kinematic distance implies that angular position and velocity are nearly identical for the two moments being compared. Yet the corresponding pair of neural states can be quite distant. Under the concern above, this distance would be due to other encoded behavioral variables – perhaps joint angle and joint velocity – differing between those two moments. However, there are not enough degrees of freedom in this task to make this plausible. The shoulder remains at a fixed position (because the head is fixed) and the wrist has limited mobility due to the pedal design [60]. Thus, shoulder and elbow angles are almost completely determined by cycle phase. More generally, ‘external variables’ (positions, angles, and their derivatives) are unlikely to differ more than slightly when phase and angular velocity are matched. Muscle activity could be different because many muscles act on each joint, creating redundancy. However, as illustrated in Figure 2c, the key effect is just as clear when analyzing muscle activity. Thus, the above concern seems unlikely even if it can’t be ruled out entirely. A broader reason to doubt the ‘wrong parameter choice’ proposition is that it provides a vague explanation for a phenomenon that already has a straightforward explanation. A lack of isometry between the neural population response and behavior is expected when neural-trajectory tangling is low and output-null factors are plentiful [55, 60]. For example, in networks that generate muscle activity, neural and muscle-activity trajectories are far from isometric [52, 58, 60]. Given this straightforward explanation, and given repeated failures over decades to find the ‘correct’ parameters (muscle activity, movement direction, etc.) that create neural-behavior isometry, it seems reasonable to conclude that no such isometry exists.”

Comment 1) The authors posit that neural and behavioral trajectories are non-isometric. To support this point, they look at distances between neural states and distances between the corresponding behavioral states, in order to demonstrate that there are differences in these distances in each respective space. This supports the idea that neural states and behavioral states are non-isometric but does not directly address their point. In order to say the trajectories are non-isometric, it would be better to look at pairs of distances between corresponding trajectories in each space.

We like this idea and have added such an analysis. To be clear, we like the original analysis too: isometry predicts that neural and behavioral distances (for corresponding pairs of points) should be strongly correlated, and that small behavioral distances should not be associated with large neural distances. These predictions are not true, providing a strong argument against isometry. However, we also like the reviewer’s suggestion, and have added such an analysis. It makes the same larger point, and also reveals some additional facts (e.g. it reveals that muscle-geometry is more related to neural-geometry than is kinematic-geometry). The new analysis is described in the following section:

“We further explored the topic of isometry by considering pairs of distances. To do so, we chose two random neural states and computed their distance, yielding dneural1. We repeated this process, yielding dneural2. We then computed the corresponding pair of distances in muscle space (dmuscle1 and dmuscle2) and kinematic space (dkin1 and dkin2). We considered cases where dneural1 was meaningfully larger than (or smaller than) dneural2, and asked whether the behavioral variables had the same relationship; e.g. was dmuscle1 also larger than dmuscle2? For kinematics, this relationship was weak: across 100,000 comparisons, the sign of dkin1 − dkin2 agreed with dneural1 − dneural2 only 67.3% of the time (with 50% being chance). The relationship was much stronger for muscles: the sign of dmuscle1 − dmuscle2 agreed with dneural1 − dneural2 79.2% of the time, which is far more than expected by chance yet also far from what is expected given isometry (e.g. the sign agrees 99.7% of the time for the truly isometric control data in Figure 2e). Indeed there were multiple moments during this task when dneural1 was much larger than dneural2, yet dmuscle1 was smaller than dmuscle2. These observations are consistent with the proposal that neural trajectories resemble muscle trajectories in some dimensions, but with additional output-null dimensions that break the isometry [60].”

Comment 3) The approach is built up on the idea of creating a "mesh" structure of possible states. In the body of the paper the definition of the mesh was not entirely clear and I could not find in the methods a more rigorous explicit definition. Since the mesh is integral to the approach, the paper would be improved with more description of this component.

This is a fair criticism. Although MINTs actual operations were well-documented, how those operations mapped onto the term ‘mesh’ was, we agree, a bit vague. The definition of the mesh is a bit subtle because it only emerges during decoding rather than being precomputed. This is part of what gives MINT much more flexibility than a lookup table. We have added the following to the manuscript.

“We use the term ‘mesh’ to describe the scaffolding created by the training-set trajectories and the interpolated states that arise at runtime. The term mesh is apt because, if MINT’s assumptions are correct, interpolation will almost always be local. If so, the set of decodable states will resemble a mesh, created by line segments connecting nearby training-set trajectories. However, this mesh-like structure is not enforced by MINT’s operations.

Interpolation could, in principle, create state-distributions that depart from the assumption of a sparse manifold. For example, interpolation could fill in the center of the green tube in Figure 1b, resulting in a solid manifold rather than a mesh around its outer surface. However, this would occur only if spiking observations argued for it. As will be documented below, we find that essentially all interpolation is local”

We have also added Figure 4d. This new analysis documents the fact that decoded states are near trainingset trajectories, which is why the term ‘mesh’ is appropriate.

Reviewer #3:

Summary:

This manuscript develops a new method termed MINT for decoding of behavior. The method is essentially a table-lookup rather than a model. Within a given stereotyped task, MINT tabulates averaged firing rate trajectories of neurons (neural states) and corresponding averaged behavioral trajectories as stereotypes to construct a library. For a test trial with a realized neural trajectory, it then finds the closest neural trajectory to it in the table and declares the associated behavior trajectory in the table as the decoded behavior. The method can also interpolate between these tabulated trajectories. The authors mention that the method is based on three key assumptions: (1) Neural states may not be embedded in a lowdimensional subspace, but rather in a high-dimensional space. (2) Neural trajectories are sparsely distributed under different behavioral conditions. (3) These neural states traverse trajectories in a stereotyped order.

The authors conducted multiple analyses to validate MINT, demonstrating its decoding of behavioral trajectories in simulations and datasets (Figures 3, 4). The main behavior decoding comparison is shown in Figure 4. In stereotyped tasks, decoding performance is comparable (M_Cycle, MC_Maze) or better (Area 2_Bump) than other linear/nonlinear algorithms

(Figure 4). However, MINT underperforms for the MC_RTT task, which is less stereotyped (Figure 4).

This paper is well-structured and its main idea is clear. The fact that performance on stereotyped tasks is high is interesting and informative, showing that these stereotyped tasks create stereotyped neural trajectories. The task-specific comparisons include various measures and a variety of common decoding approaches, which is a strength. However, I have several major concerns. I believe several of the conclusions in the paper, which are also emphasized in the abstract, are not accurate or supported, especially about generalization, computational scalability, and utility for BCIs. MINT is essentially a table-lookup algorithm based on stereotyped task-dependent trajectories and involves the tabulation of extensive data to build a vast library without modeling. These aspects will limit MINT's utility for real-world BCIs and tasks. These properties will also limit MINT's generalizability from task to task, which is important for BCIs and thus is commonly demonstrated in BCI experiments with other decoders without any retraining. Furthermore, MINT's computational and memory requirements can be prohibitive it seems. Finally, as MINT is based on tabulating data without learning models of data, I am unclear how it will be useful in basic investigations of neural computations. I expand on these concerns below.

We thank the reviewer for pointing out weaknesses in our framing and presentation. The comments above made us realize that we needed to 1) better document the ways in which MINT is far more flexible than a lookup-table, and 2) better explain the competing scientific perspectives at play. R3’s comments also motivated us to add an additional analysis of generalization. In our view the manuscript is greatly improved by these additions. Specifically, these additions directly support the broader impact that we hope the study will have.

For simplicity and readability, we first group and summarize R3’s main concerns in order to better address them. (These main concerns are all raised above, in addition to recurring in the specific comments below. Responses to each individual specific comment are provided after these summaries.)

(1) R3 raises concerns about ‘computational scalability.’ The concern is that “MINT's computational and memory requirements can be prohibitive.” This point was expanded upon in a specific comment, reproduced below:

I also find the statement in the abstract and paper that "computations are simple, scalable" to be inaccurate. The authors state that MINT's computational cost is O(NC) only, but it seems this is achieved at a high memory cost as well as computational cost in training. The process is described in section "Lookup table of log-likelihoods" on line [978-990]. The idea is to precompute the log-likelihoods for any combination of all neurons with discretization x all delay/history segments x all conditions and to build a large lookup table for decoding. Basically, the computational cost of precomputing this table is O(V^{Nτ} x TC) and the table requires a memory of O(V^{Nτ}), where V is the number of discretization points for the neural firing rates, N is the number of neurons, τ is the history length, T is the trial length, and C is the number of conditions. This is a very large burden, especially the V^{Nτ} term. This cost is currently not mentioned in the manuscript and should be clarified in the main text. Accordingly, computation claims should be modified including in the abstract.

The revised manuscript clarifies that our statement (that computations are simple and scalable) is absolutely accurate. There is no need to compute, or store, a massive lookup table. There are three tables: two of modest size and one that is tiny. This is now better explained:

“Thus, the log-likelihood of , for a particular current neural state, is simply the sum of many individual log-likelihoods (one per neuron and time-bin). Each individual log-likelihood depends on only two numbers: the firing rate at that moment and the spike count in that bin. To simplify online computation, one can precompute the log-likelihood, under a Poisson model, for every plausible combination of rate and spike-count. For example, a lookup table of size 2001 × 21 is sufficient when considering rates that span 0-200 spikes/s in increments of 0.1 spikes/s, and considering 20 ms bins that contain at most 20 spikes (only one lookup table is ever needed, so long as its firing-rate range exceeds that of the most-active neuron at the most active moment in Ω). Now suppose we are observing a population of 200 neurons, with a 200 ms history divided into ten 20 ms bins. For each library state, the log-likelihood of the observed spike-counts is simply the sum of 200 × 10 = 2000 individual loglikelihoods, each retrieved from the lookup table. In practice, computation is even simpler because many terms can be reused from the last time bin using a recursive solution (Methods). This procedure is lightweight and amenable to real-time applications.”

In summary, the first table simply needs to contain the firing rate of each neuron, for each condition, and each time in that condition. This table consumes relatively little memory. Assuming 100 one-second-long conditions (rates sampled every 20 ms) and 200 neurons, the table would contain 100 x 50 x 200 = 1,000,000 numbers. These numbers are typically stored as 16-bit integers (because rates are quantized), which amounts to about 2 MB. This is modest, given that most computers have (at least) tens of GB of RAM. A second table would contain the values for each behavioral variable, for each condition, and each time in that condition. This table might contain behavioral variables at a finer resolution (e.g. every millisecond) to enable decoding to update in between 20 ms bins (1 ms granularity is not needed for most BCI applications, but is the resolution used in this study). The number of behavioral variables of interest for a particular BCI application is likely to be small, often 1-2, but let’s assume for this example it is 10 (e.g. x-, y-, and z-position, velocity, and acceleration of a limb, plus one other variable). This table would thus contain 100 x 1000 x 10 = 1,000,000 floating point numbers, i.e. an 8 MB table. The third table is used to store the probability of s spikes being observed given a particular quantized firing rate (e.g. it may contain probabilities associated with firing rates ranging from 0 – 200 spikes/s in 0.1 spikes/s increments). This table is not necessary, but saves some computation time by precomputing numbers that will be used repeatedly. This is a very small table (typically ~2000 x 20, i.e. 320 KB). It does not need to be repeated for different neurons or conditions, because Poisson probabilities depend on only rate and count.

(2) R3 raises a concern that MINT “is essentially a table-lookup rather than a model.’ R3 states that MINT

“is essentially a table-lookup algorithm based on stereotyped task-dependent trajectories and involves the tabulation of extensive data to build a vast library without modeling.”

and that,

“as MINT is based on tabulating data without learning models of data, I am unclear how it will be useful in basic investigations of neural computations.”

This concern is central to most subsequent concerns. The manuscript has been heavily revised to address it. The revisions clarify that MINT is much more flexible than a lookup table, even though MINT uses a lookup table as its first step. Because R3’s concern is intertwined with one’s scientific assumptions, we have also added the new Figure 1 to explicitly illustrate the two key scientific perspectives and their decoding implications.

Under the perspective in Figure 1a, R3 would be correct in saying that there exist traditional interpretable decoders (e.g. a Kalman filter) whose assumptions better model the data. Under this perspective, MINT might still be an excellent choice in many cases, but other methods would be expected to gain the advantage when situations demand more flexibility. This is R3’s central concern, and essentially all other concerns flow from it. It makes sense that R3 has this concern, because their comments repeatedly stress a foundational assumption of the perspective in Figure 1a: the assumption of a fixed lowdimensional neural subspace where activity has a reliable relationship to behavior that can be modeled and leveraged during decoding. The phrases below accord with that view:

“Unlike MINT, these works can achieve generalization because they model the neural subspace and its association to movement.”

“it will not generalize… even when these tasks cover the exact same space (e.g., the same 2D computer screen and associated neural space).”

“For proper training, the training data should explore the whole movement space and the associated neural space”

“I also believe the authors should clarify the logic behind developing MINT better. From a scientific standpoint, we seek to gain insights into neural computations by making various assumptions and building models that parsimoniously describe the vast amount of neural data rather than simply tabulating the data. For instance, low-dimensional assumptions have led to the development of numerous dimensionality reduction algorithms and these models have led to important interpretations about the underlying dynamics”

Thus, R3 prefers a model that 1) assumes a low-dimensional subspace that is fixed across tasks and 2) assumes a consistent ‘association’ between neural activity and kinematics. Because R3 believes this is the correct model of the data, they believe that decoders should leverage it. Traditional interpretable method do, and MINT doesn’t, which is why they find MINT to be unprincipled. This is a reasonable view, but it is not our view. We have heavily revised the manuscript to clarify that a major goal of our study is to explore the implications of a different, less-traditional scientific perspective.

The new Figure 1a illustrates the traditional perspective. Under this perspective, one would agree with R3’s claim that other methods have the opportunity to model the data better. For example, suppose there exists a consistent neural subspace – conserved across tasks – where three neural dimensions encode 3D hand position and three additional neural dimensions encode 3D hand velocity. A traditional method such as a Kalman filter would be a very appropriate choice to model these aspects of the data.

Figure 1b illustrates the alternative scientific perspective. This perspective arises from recent, present, and to-be-published observations. MINT’s assumptions are well-aligned with this perspective. In contrast, the assumptions of traditional methods (e.g. the Kalman filter) are not well-aligned with the properties of the data under this perspective. This does not mean traditional methods are not useful. Yet under Figure 1b, it is traditional methods, such as the Kalman filter, that lack an accurate model of the data. Of course, the reviewer may disagree with our scientific perspective. We would certainly concede that there is room for debate. However, we find the evidence for Figure 1b to be sufficiently strong that it is worth exploring the utility of methods that align with this scientific perspective. MINT is such a method. As we document, it performs very well.

Thus, in our view, MINT is quite principled because its assumptions are well aligned with the data. It is true that the features of the data that MINT models are a bit different from those that are traditionally modeled. For example, R3 is quite correct that MINT does not attempt to use a biomimetic model of the true transformation from neural activity, to muscle activity, and thence to kinematics. We see this as a strength, and the manuscript has been revised accordingly (see paragraph beginning with “We leveraged this simulated data to compare MINT with a biomimetic decoder”).

(3) R3 raises concerns that MINT cannot generalize. This was a major concern of R3 and is intimately related to concern #2 above. The concern is that, if MINT is “essentially a lookup table” that simply selects pre-defined trajectories, then MINT will not be able to generalize. R3 is quite correct that MINT generalizes rather differently than existing methods. Whether this is good or bad depends on one’s scientific perspective. Under Figure 1a, MINT’s generalization would indeed be limiting because other methods could achieve greater flexibility. Under Figure 1b, all methods will have serious limits regarding generalization. Thus, MINT’s method for generalizing may approximate the best one can presently do. To address this concern, we have made three major changes, numbered i-iii below:

i) Large sections of the manuscript have been restructured to underscore the ways in which MINT can generalize. A major goal was to counter the impression, stated by R3 above, that:

“for a test trial with a realized neural trajectory, [MINT] then finds the closest neural trajectory to it in the table and declares the associated behavior trajectory in the table as the decoded behavior”.

This description is a reasonable way to initially understand how MINT works, and we concede that we may have over-used this intuition. Unfortunately, it can leave the misimpression that MINT decodes by selecting whole trajectories, each corresponding to ‘a behavior’. This can happen, but it needn’t and typically doesn’t. As an example, consider the cycling task. Suppose that the library consists of stereotyped trajectories, each four cycles long, at five fixed speeds from 0.5-2.5 Hz. If the spiking observations argued for it, MINT could decode something close to one of these five stereotyped trajectories. Yet it needn’t. Decoded trajectories will typically resemble library trajectories locally, but may be very different globally. For example, a decoded trajectory could be thirty cycles long (or two, or five hundred) perhaps speeding up and slowing down multiple times across those cycles.

Thus, the library of trajectories shouldn’t be thought of as specifying a limited set of whole movements that can be ‘selected from’. Rather, trajectories define a scaffolding that outlines where the neural state is likely to live and how it is likely to be changing over time. When we introduce the idea of library trajectories, we are now careful to stress that they don’t function as a set from which one trajectory is ‘declared’ to be the right one:

“We thus designed MINT to approximate that manifold using the trajectories themselves, rather than their covariance matrix or corresponding subspace. Unlike a covariance matrix, neural trajectories indicate not only which states are likely, but also which state-derivatives are likely. If a neural state is near previously observed states, it should be moving in a similar direction. MINT leverages this directionality.

Training-set trajectories can take various forms, depending on what is convenient to collect. Most simply, training data might include one trajectory per condition, with each condition corresponding to a discrete movement. Alternatively, one might instead employ one long trajectory spanning many movements. Another option is to employ many sub-trajectories, each briefer than a whole movement. The goal is simply for training-set trajectories to act as a scaffolding, outlining the manifold that might be occupied during decoding and the directions in which decoded trajectories are likely to be traveling.”

Later in that same section we stress that decoded trajectories can move along the ‘mesh’ in nonstereotyped ways:

“Although the mesh is formed of stereotyped trajectories, decoded trajectories can move along the mesh in non-stereotyped ways as long as they generally obey the flow-field implied by the training data. This flexibility supports many types of generalization, including generalization that is compositional in nature. Other types of generalization – e.g. from the green trajectories to the orange trajectories in Figure 1b – are unavailable when using MINT and are expected to be challenging for any method (as will be documented in a later section).”

The section “Training and decoding using MINT” has been revised to clarify the ways in which interpolation is flexible, allowing decoded movements to be globally very different from any library trajectory.

“To decode stereotyped trajectories, one could simply obtain the maximum-likelihood neural state from the library, then render a behavioral decode based on the behavioral state with the same values of c and k. This would be appropriate for applications in which conditions are categorical, such as typing or handwriting. Yet in most cases we wish for the trajectory library to serve not as an exhaustive set of possible states, but as a scaffolding for the mesh of possible states. MINT’s operations are thus designed to estimate any neural trajectory – and any corresponding behavioral trajectory – that moves along the mesh in a manner generally consistent with the trajectories in Ω.”

“…interpolation allows considerable flexibility. Not only is one not ‘stuck’ on a trajectory from Φ, one is also not stuck on trajectories created by weighted averaging of trajectories in Φ. For example, if cycling speed increases, the decoded neural state could move steadily up a scaffolding like that illustrated in Figure 1b (green). In such cases, the decoded trajectory might be very different in duration from any of the library trajectories. Thus, one should not think of the library as a set of possible trajectories that are selected from, but rather as providing a mesh-like scaffolding that defines where future neural states are likely to live and the likely direction of their local motion. The decoded trajectory may differ considerably from any trajectory within Ω.”

This flexibility is indeed used during movement. One empirical example is described in detail:

“During movement… angular phase was decoded with effectively no net drift over time. This is noteworthy because angular velocity on test trials never perfectly matched any of the trajectories in Φ. Thus, if decoding were restricted to a library trajectory, one would expect growing phase discrepancies. Yet decoded trajectories only need to locally (and approximately) follow the flow-field defined by the library trajectories. Based on incoming spiking observations, decoded trajectories speed up or slow down (within limits).

This decoding flexibility presumably relates to the fact that the decoded neural state is allowed to differ from the nearest state in Ω. To explore… [the text goes on to describe the new analysis in Figure 4d, which shows that the decoded state is typically not on any trajectory, though it is typically close to a trajectory].”

Thus, MINT’s operations allow considerable flexibility, including generalization that is compositional in nature. Yet R3 is still correct that there are other forms of generalization that are unavailable to MINT. This is now stressed at multiple points in the revision. However, under the perspective in Figure 1b, these forms of generalization are unavailable to any current method. Hence we made a second major change in response to this concern… ii) We explicitly illustrate how the structure of the data determines when generalization is or isn’t possible. The new Figure 1a,b introduces the two perspectives, and the new Figure 6a,b lays out their implications for generalization. Under the perspective in Figure 6a, the reviewer is quite right: other methods can generalize in ways that MINT cannot. Under the perspective in Figure 6b, expectations are very different. Those expectations make testable predictions. Hence the third major change… iii) We have added an analysis of generalization, using a newly collected dataset. This dataset was collected using Neuropixels Probes during our Pac-Man force-tracking task. This dataset was chosen because it is unusually well-suited to distinguishing the predictions in Figure 6a versus Figure 6b. Finding a dataset that can do so is not simple. Consider R3’s point that training data should “explore the whole movement space and the associated neural space”. The physical simplicity of the Pac-Man task makes it unusually easy to confirm that the behavioral workspace has been fully explored. Importantly, under Figure 6b, this does not mean that the neural workspace has been fully explored, which is exactly what we wish to test when testing generalization. We do so, and compare MINT with a Wiener filter. A Wiener filter is an ideal comparison because it is simple, performs very well on this task, and should be able to generalize well under Figure 1a. Additionally, the Wiener filter (unlike the Kalman Filter) doesn’t leverage the assumption that neural activity reflects the derivative of force. This matters because we find that neural activity does not reflect dforce/dt in this task. The Wiener filter is thus the most natural choice of the interpretable methods whose assumptions match Figure 1a.

The new analysis is described in Figure 6c-g and accompanying text. Results are consistent with the predictions of Figure 6b. We are pleased to have been motivated to add this analysis for two reasons. First, it provides an additional way of evaluating the predictions of the two competing scientific perspectives that are at the heart of our study. Second, this analysis illustrates an underappreciated way in which generalization is likely to be challenging for any decode method. It can be tempting to think that the main challenge regarding generalization is to fully explore the relevant behavioral space. This makes sense if a behavioral space has “an associated neural space”. However, we are increasingly of the opinion that it doesn’t. Different tasks often involve different neural subspaces, even when behavioral subspaces overlap. We have even seen situations where motor output is identical but neural subspaces are quite different. These facts are relevant to any decoder, something highlighted in the revised Introduction:

“MINT’s performance confirms that there are gains to be made by building decoders whose assumptions match a different, possibly more accurate view of population activity. At the same time, our results suggest fundamental limits on decoder generalization. Under the assumptions in Figure 1b, it will sometimes be difficult or impossible for decoders to generalize to not-yet-seen tasks. We found that this was true regardless of whether one uses MINT or a more traditional method. This finding has implications regarding when and how generalization should be attempted.”

We have also added an analysis (Figure 6e) illustrating how MINT’s ability to compute likelihoods can be useful in detecting situations that may strain generalization (for any method). MINT is unusual in being able to compute and use likelihoods in this way.

Detailed responses to R3: we reproduce each of R3’s specific concerns below, but concentrate our responses on issues not already covered above.

Main comments:

Comment 1. MINT does not generalize to different tasks, which is a main limitation for BCI utility compared with prior BCI decoders that have shown this generalizability as I review below. Specifically, given that MINT tabulates task-specific trajectories, it will not generalize to tasks that are not seen in the training data even when these tasks cover the exact same space (e.g., the same 2D computer screen and associated neural space).

First, the authors provide a section on generalization, which is inaccurate because it mixes up two fundamentally different concepts: 1) collecting informative training data and 2) generalizing from task to task. The former is critical for any algorithm, but it does not imply the latter. For example, removing one direction of cycling from the training set as the authors do here is an example of generating poor training data because the two behavioral (and neural) directions are non-overlapping and/or orthogonal while being in the same space. As such, it is fully expected that all methods will fail. For proper training, the training data should explore the whole movement space and the associated neural space, but this does not mean all kinds of tasks performed in that space must be included in the training set (something MINT likely needs while modeling-based approaches do not). Many BCI studies have indeed shown this generalization ability using a model. For example, in Weiss et al. 2019, center-out reaching tasks are used for training and then the same trained decoder is used for typing on a keyboard or drawing on the 2D screen. In Gilja et al. 2012, training is on a center-out task but the same trained decoder generalizes to a completely different pinball task (hit four consecutive targets) and tasks requiring the avoidance of obstacles and curved movements. There are many more BCI studies, such as Jarosiewicz et al. 2015 that also show generalization to complex realworld tasks not included in the training set. Unlike MINT, these works can achieve generalization because they model the neural subspace and its association to movement. On the contrary, MINT models task-dependent neural trajectories, so the trained decoder is very task-dependent and cannot generalize to other tasks. So, unlike these prior BCIs methods, MINT will likely actually need to include every task in its library, which is not practical.

I suggest the authors remove claims of generalization and modify their arguments throughout the text and abstract. The generalization section needs to be substantially edited to clarify the above points. Please also provide the BCI citations and discuss the above limitation of MINT for BCIs.

As discussed above, R3’s concerns are accurate under the view in Figure 1a (and the corresponding Figure 6a). Under this view, a method such as that in Gilja et al. or Jarosiewicz et al. can find the correct subspace, model the correct neuron-behavior correlations, and generalize to any task that uses “the same 2D computer screen and associated neural space”, just as the reviewer argues. Under Figure 1b things are quite different.

This topic – and the changes we have made to address it – is covered at length above. Here we simply want to highlight an empirical finding: sometimes two tasks use the same neural subspace and sometimes they don’t. We have seen both in recent data, and it is can be very non-obvious which will occur based just on behavior. It does not simply relate to whether one is using the same physical workspace. We have even seen situations where the patterns of muscle activity in two tasks are nearly identical, but the neural subspaces are fairly different. When a new task uses a new subspace, neither of the methods noted above (Gilja nor Jarosiewicz) will generalize (nor will MINT). Generalizing to a new subspace is basically impossible without some yet-to-be-invented approach. On the other hand, there are many other pairs of tasks (center-out-reaching versus some other 2D cursor control) where subspaces are likely to be similar, especially if the frequency content of the behavior is similar (in our recent experience this is often critical). When subspaces are shared, most methods will generalize, and that is presumably why generalization worked well in the studies noted above.

Although MINT can also generalize in such circumstances, R3 is correct that, under the perspective in Figure 1a, MINT will be more limited than other methods. This is now carefully illustrated in Figure 6a. In this traditional perspective, MINT will fail to generalize in cases where new trajectories are near previously observed states, yet move in very different ways from library trajectories. The reason we don’t view this is a shortcoming is that we expect it to occur rarely (else tangling would be high). We thus anticipate the scenario in Figure 6b.

This is worth stressing because R3 states that our discussion of generalization “is inaccurate because it mixes up two fundamentally different concepts: 1) collecting informative training data and 2) generalizing from task to task.” We have heavily revised this section and improved it. However, it was never inaccurate. Under Figure 6b, these two concepts absolutely are mixed up. If different tasks use different neural subspaces, then this requires collecting different “informative training data” for each. One cannot simply count on having explored the physical workspace.

Comment 2. MINT is shown to achieve competitive/high performance in highly stereotyped datasets with structured trials, but worse performance on MC_RTT, which is not based on repeated trials and is less stereotyped. This shows that MINT is valuable for decoding in repetitive stereotyped use-cases. However, it also highlights a limitation of MINT for BCIs, which is that MINT may not work well for real-world and/or less-constrained setups such as typing, moving a robotic arm in 3D space, etc. This is again due to MINT being a lookup table with a library of stereotyped trajectories rather than a model. Indeed, the authors acknowledge that the lower performance on MC_RTT (Figure 4) may be caused by the lack of repeated trials of the same type. However, real-world BCI decoding scenarios will also not have such stereotyped trial structure and will be less/un-constrained, in which MINT underperforms. Thus, the claim in the abstract or lines 480-481 that MINT is an "excellent" candidate for clinical BCI applications is not accurate and needs to be qualified. The authors should revise their statements according and discuss this issue. They should also make the use-case of MINT on BCI decoding clearer and more convincing.

We discussed, above, multiple changes and additions to the revision that were made to address these concerns. Here we briefly expand on the comment that MINT achieves “worse performance on MC_RTT, which is not based on repeated trials and is less stereotyped”. All decoders performed poorly on this task. MINT still outperformed the two traditional methods, but this was the only dataset where MINT did not also perform better (overall) than the expressive GRU and feedforward network. There are probably multiple reasons why. We agree with R3 that one likely reason is that this dataset is straining generalization, and MINT may have felt this strain more than the two machine-learning-based methods. Another potential reason is the structure of the training data, which made it more challenging to obtain library trajectories in the first place. Importantly, these observations do not support the view in Figure 1a. MINT still outperformed the Kalman and Wiener filters (whose assumptions align with Fig. 1a). To make these points we have added the following:

“Decoding was acceptable, but noticeably worse, for the MC_RTT dataset… As will be discussed below, every decode method achieved its worst estimates of velocity for the MC_RTT dataset. In addition to the impact of slower reaches, MINT was likely impacted by training data that made it challenging to accurate estimate library trajectories. Due to the lack of repeated trials, MINT used AutoLFADS to estimate the neural state during training. In principle this should work well. In practice AutoLFADS may have been limited by having only 10 minutes of training data. Because the random-target task involved more variable reaches, it may also have stressed the ability of all methods to generalize, perhaps for the reasons illustrated in Figure 1b.

The only dataset where MINT did not perform the best overall was the MC_RTT dataset, where it was outperformed by the feedforward network and GRU. As noted above, this may relate to the need for MINT to learn neural trajectories from training data that lacked repeated trials of the same movement (a design choice one might wish to avoid). Alternatively, the less-structured MC_RTT dataset may strain the capacity to generalize; all methods experienced a drop in velocity-decoding R2 for this dataset compared to the others. MINT generalizes somewhat differently than other methods, and may have been at a modest disadvantage for this dataset. A strong version of this possibility is that perhaps the perspective in Figure 1a is correct, in which case MINT might struggle because it cannot use forms of generalization that are available to other methods (e.g. generalization based on neuron-velocity correlations). This strong version seems unlikely; MINT continued to significantly outperform the Wiener and Kalman filters, which make assumptions aligned with Figure 1a.”

Comment 3. Related to 2, it may also be that MINT achieves competitive performance in offline and trial-based stereotyped decoding by overfitting to the trial structure in a given task, and thus may not generalize well to online performance due to overfitting. For example, a recent work showed that offline decoding performance may be overfitted to the task structure and may not represent online performance (Deo et al. 2023). Please discuss.

We agree that a limitation of our study is that we do not test online performance. There are sensible reasons for this decision:

“By necessity and desire, all comparisons were made offline, enabling benchmarked performance across a variety of tasks and decoded variables, where each decoder had access to the exact same data and recording conditions.”

We recently reported excellent online performance in the cycling task with a different algorithm

(Schroeder et al. 2022). In the course of that study, we consistently found that improvements in our offline decoding translated to improvements in our online decoding. We thus believe that MINT (which improves on the offline performance of our older algorithm) is a good candidate to work very well online. Yet we agree this still remains to be seen. We have added the following to the Discussion:

“With that goal in mind, there exist three important practical considerations. First, some decode algorithms experience a performance drop when used online. One presumed reason is that, when decoding is imperfect, the participant alters their strategy which in turn alters the neural responses upon which decoding is based. Because MINT produces particularly accurate decoding, this effect may be minimized, but this cannot be known in advance. If a performance drop does indeed occur, one could adapt the known solution of retraining using data collected during online decoding [13]. Another presumed reason (for a gap between offline and online decoding) is that offline decoders can overfit the temporal structure in training data [107]. This concern is somewhat mitigated by MINT’s use of a short spike-count history, but MINT may nevertheless benefit from data augmentation strategies such as including timedilated versions of learned trajectories in the libraries”

Comment 4. Related to 2, since MINT requires firing rates to generate the library and simple averaging does not work for this purpose in the MC_RTT dataset (that does not have repeated trials), the authors needed to use AutoLFADS to infer the underlying firing rates. The fact that MINT requires the usage of another model to be constructed first and that this model can be computationally complex, will also be a limiting factor and should be clarified.

This concern relates to the computational complexity of computing firing-rate trajectories during training. Usually, rates are estimated via trial-averaging, which makes MINT very fast to train. This was quite noticeable during the Neural Latents Benchmark competition. As one example, for the “MC_Scaling 5 ms Phase”, MINT took 28 seconds to train while GPFA took 30 minutes, the transformer baseline (NDT) took 3.5 hours, and the switching nonlinear dynamical system took 4.5 hours.

However, the reviewer is quite correct that MINT’s efficiency depends on the method used to construct the library of trajectories. As we note, “MINT is a method for leveraging a trajectory library, not a method for constructing it”. One can use trial-averaging, which is very fast. One can also use fancier, slower methods to compute the trajectories. We don’t view this as a negative – it simply provides options. Usually one would choose trial-averaging, but one does not have to. In the case of MC_RTT, one has a choice between LFADS and grouping into pseudo-conditions and averaging (which is fast). LFADS produces higher performance at the cost of being slower. The operator can choose which they prefer. This is discussed in the following section:

“For MINT, ‘training’ simply means computation of standard quantities (e.g. firing rates) rather than parameter optimization. MINT is thus typically very fast to train (Table 1), on the order of seconds using generic hardware (no GPUs). This speed reflects the simple operations involved in constructing the library of neural-state trajectories: filtering of spikes and averaging across trials. At the same time we stress that MINT is a method for leveraging a trajectory library, not a method for constructing it. One may sometimes wish to use alternatives to trial-averaging, either of necessity or because they improve trajectory estimates. For example, for the MC_RTT task we used AutoLFADS to infer the library. Training was consequently much slower (hours rather than seconds) because of the time taken to estimate rates. Training time could be reduced back to seconds using a different approach – grouping into pseudo-conditions and averaging – but performance was reduced. Thus, training will typically be very fast, but one may choose time-consuming methods when appropriate.”

Comment 5. I also find the statement in the abstract and paper that "computations are simple, scalable" to be inaccurate. The authors state that MINT's computational cost is O(NC) only, but it seems this is achieved at a high memory cost as well as computational cost in training. The process is described in section "Lookup table of log-likelihoods" on line [978-990]. The idea is to precompute the log-likelihoods for any combination of all neurons with discretization x all delay/history segments x all conditions and to build a large lookup table for decoding. Basically, the computational cost of precomputing this table is O(V^{Nτ} x TC) and the table requires a memory of O(V^{Nτ}), where V is the number of discretization points for the neural firing rates, N is the number of neurons, τ is the history length, T is the trial length, and C is the number of conditions. This is a very large burden, especially the V^{Nτ} term. This cost is currently not mentioned in the manuscript and should be clarified in the main text. Accordingly, computation claims should be modified including in the abstract.

As discussed above, the manuscript has been revised to clarify that our statement was accurate.

Comment 6. In addition to the above technical concerns, I also believe the authors should clarify the logic behind developing MINT better. From a scientific standpoint, we seek to gain insights into neural computations by making various assumptions and building models that parsimoniously describe the vast amount of neural data rather than simply tabulating the data. For instance, low-dimensional assumptions have led to the development of numerous dimensionality reduction algorithms and these models have led to important interpretations about the underlying dynamics (e.g., fixed points/limit cycles). While it is of course valid and even insightful to propose different assumptions from existing models as the authors do here, they do not actually translate these assumptions into a new model. Without a model and by just tabulating the data, I don't believe we can provide interpretation or advance the understanding of the fundamentals behind neural computations. As such, I am not clear as to how this library building approach can advance neuroscience or how these assumptions are useful. I think the authors should clarify and discuss this point.

As requested, a major goal of the revision has been to clarify the scientific motivations underlying MINT’s design. In addition to many textual changes, we have added figures (Figures 1a,b and 6a,b) to outline the two competing scientific perspectives that presently exist. This topic is also addressed by extensions of existing analyses and by new analyses (e.g. Figure 6c-g).

In our view these additions have dramatically improved the manuscript. This is especially true because we think R3’s concerns, expressed above, are reasonable. If the perspective in Figure 1a is correct, then R3 is right and MINT is essentially a hack that fails to model the data. MINT would still be effective in many circumstances (as we show), but it would be unprincipled. This would create limitations, just as the reviewer argues. On the other hand, if the perspective in Figure 1b is correct, then MINT is quite principled relative to traditional approaches. Traditional approaches make assumptions (a fixed subspace, consistent neuron-kinematic correlations) that are not correct under Figure 1b.

We don’t expect R3 to agree with our scientific perspective at this time (though we hope to eventually convince them). To us, the key is that we agree with R3 that the manuscript needs to lay out the different perspectives and their implications, so that readers have a good sense of the possibilities they should be considering. The revised manuscript is greatly improved in this regard.

Comment 7. Related to 6, there seems to be a logical inconsistency between the operations of MINT and one of its three assumptions, namely, sparsity. The authors state that neural states are sparsely distributed in some neural dimensions (Figure 1a, bottom). If this is the case, then why does MINT extend its decoding scope by interpolating known neural states (and behavior) in the training library? This interpolation suggests that the neural states are dense on the manifold rather than sparse, thus being contradictory to the assumption made. If interpolation-based dense meshes/manifolds underlie the data, then why not model the neural states through the subspace or manifold representations? I think the authors should address this logical inconsistency in MINT, especially since this sparsity assumption also questions the low-dimensional subspace/manifold assumption that is commonly made.

We agree this is an important issue, and have added an analysis on this topic (Figure 4d). The key question is simple and empirical: during decoding, does interpolation cause MINT to violate the assumption of sparsity? R3 is quite right that in principle it could. If spiking observations argue for it, MINT’s interpolation could create a dense manifold during decoding rather than a sparse one. The short answer is that empirically this does not happen, in agreement with expectations under Figure 1b. Rather than interpolating between distant states and filling in large ‘voids’, interpolation is consistently local. This is a feature of the data, not of the decoder (MINT doesn’t insist upon sparsity, even though it is designed to work best in situations where the manifold is sparse).

In addition to adding Figure 4d, we added the following (in an earlier section):

“The term mesh is apt because, if MINT’s assumptions are correct, interpolation will almost always be local. If so, the set of decodable states will resemble a mesh, created by line segments connecting nearby training-set trajectories. However, this mesh-like structure is not enforced by MINT’s operations. Interpolation could, in principle, create state-distributions that depart from the assumption of a sparse manifold. For example, interpolation could fill in the center of the green tube in Figure 1b, resulting in a solid manifold rather than a mesh around its outer surface. However, this would occur only if spiking observations argued for it. As will be documented below, we find that essentially all interpolation is local.”

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

I appreciate the detailed methods section, however, more specifics should be integrated into the main text. For example on Line 238, it should additionally be stated how many minutes were used for training and metrics like the MAE which is used later should be reported here.

Thank you for this suggestion. We now report the duration of training data in the main text:

“Decoding R^2 was .968 over ~7.1 minutes of test trials based on ~4.4 minutes of training data.”

We have also added similar specifics throughout the manuscript, e.g. in the Fig. 5 legend:

“Results are based on the following numbers of training / test trials: MC\_Cycle (174 train, 99 test), MC\_Maze (1721 train, 574 test), Area2\_Bump (272 train, 92 test), MC\_RTT (810 train, 268 test).”

Similar additions were made to the legends for Fig. 6 and 8. Regarding the request to add MAE for the multitask network, we did not do so for the simple reason that the decoded variable (muscle activity) has arbitrary units. The raw MAE is thus not meaningful. We could of course have normalized, but at this point the MAE is largely redundant with the correlation. In contrast, the MAE is useful when comparing across the MC_Maze, Area2_Bump, and MC_RTT datasets, because they all involve the same scale (cm/s).

Regarding the MC_RTT task, AutoLFADS was used to obtain robust spike rates, as reported in the methods. However, the rationale for splitting the neural trajectories after AutoLFADS is unclear. If the trajectories were split based on random recording gaps, this might lead to suboptimal performance? It might be advantageous to split them based on a common behavioural state?

When learning neural trajectories via AutoLFADS, spiking data is broken into short (but overlapping) segments, rates are estimated for each segment via AutoLFADs, and these rates are then stitched together across segments into long neural trajectories. If there had been no recording gaps, these rates could have been stitched into a single neural trajectory for this dataset. However, the presence of recording gaps left us no choice but to stitch together these rates into more than one trajectory. Fortunately, recording gaps were rare: for the decoding analysis of MC_RTT there were only two recording gaps and therefore three neural trajectories, each ~2.7 minutes in duration.

We agree that in general it is desirable to learn neural trajectories that begin and end at behaviorallyrelevant moments (e.g. in between movements). However, having these trajectories potentially end midmovement is not an issue in and of itself. During decoding, MINT is never stuck on a trajectory. Thus, if MINT were decoding states near the end of a trajectory that was cut short due to a training gap, it would simply begin decoding states from other trajectories or elsewhere along the same trajectory in subsequent moments. We could have further trimmed the three neural trajectories to begin and end at behaviorallyrelevant moments, but chose not to as this would have only removed a handful of potentially useful states from the library.

We now describe this in the Methods:

“Although one might prefer trajectory boundaries to begin and end at behaviorally relevant moments (e.g. a stationary state), rather than at recording gaps, the exact boundary points are unlikely to be consequential for trajectories of this length that span multiple movements. If MINT estimates a state near the end of a long trajectory, its estimate will simply jump to another likely state on a different trajectory (or earlier along the same trajectory) in subsequent moments. Clipping the end of each trajectory to an earlier behaviorally-relevant moment would only remove potentially useful states from the libraries.”

Are the training and execution times in Table 1 based on pure Matlab functions or Mex files? If it's Mex files as suggested by the code, it would be good to mention this in the Table caption.

They are based on a combination of MATLAB and MEX files. This is now clarified in the table caption:

“Timing measurements taken on a Macbook Pro (on CPU) with 32GB RAM and a 2.3 GHz 8-Core Intel Core i9 processor. Training and execution code used for measurements was written in MATLAB (with the core recursion implemented as a MEX file).”

As the method most closely resembles a Bayesian decoder it would be good to compare performance against a Naive Bayes decoder.

We agree and have now done so. The following has been added to the text:

“A natural question is thus whether a simpler Bayesian decoder would have yielded similar results. We explored this possibility by testing a Naïve Bayes regression decoder [85] using the MC_Maze dataset. This decoder performed poorly, especially when decoding velocity (R2 = .688 and .093 for hand position and velocity, respectively), indicating that the specific modeling assumptions that differentiate MINT from a naive Bayesian decoder are important drivers of MINT’s performance.”

Line 199 Typo: The assumption of stereotypy trajectory also enables neural states (and decoded behaviors) to be updated in between time bins.

Fixed

Table 3: It's unclear why the Gaussian binning varies significantly across different datasets. Could the authors explain why this is the case and what its implications might be?

We have added the following description in the “Filtering, extracting, and warping data on each trial” subsection of the Methods to discuss how 𝜎 may vary due to the number of trials available for training and how noisy the neural data for those trials is:

“First, spiking activity for each neuron on each trial was temporally filtered with a Gaussian to yield single-trial rates. Table 3 reports the Gaussian standard deviations σ (in milliseconds) used for each dataset. Larger values of σ utilize broader windows of spiking activity when estimating rates and therefore reduce variability in those rate estimates. However, large σ values also yield neural trajectories with less fine-grained temporal structure. Thus, the optimal σ for a dataset depends on how variable the rate estimates otherwise are.”

An implementation of the method in an open-source programming language could further enhance the widespread use of the tool.

We agree this would be useful, but have yet not implemented the method in any other programming languages. Implementation in Python is still a future goal.

Reviewer #2 (Recommendations For The Authors):

- Figures 4 and 5 should show the error bars on the horizontal axis rather than portraying them vertically.

[Note that these are now Figures 5 and 6]

The figure legend of Figure 5 now clarifies that the vertical ticks are simply to aid visibility when symbols have very similar means and thus overlap visually. We don’t include error bars (for this analysis) because they are very small and would mostly be smaller than the symbol sizes. Instead, to indicate certainty regarding MINT’s performance measurements, the revised text now gives error ranges for the correlations and MAE values in the context of Figure 4c. These error ranges were computed as the standard deviation of the sampling distribution (computed via resampling of trials) and are thus equivalent to SEMs. The error ranges are all very small; e.g. for the MC_Maze dataset the MAE for x-velocity is 4.5 +/- 0.1 cm/s. (error bars on the correlations are smaller still).

Thus, for a given dataset, we can be quite certain of how well MINT performs (within ~2% in the above case). This is reassuring, but we also don’t want to overemphasize this accuracy. The main sources of variability one should be concerned about are: 1) different methods can perform differentially well for different brain areas and tasks, 2) methods can decode some behavioral variables better than others, and 3) performance depends on factors like neuron-count and the number of training trials, in ways that can differ across decode methods. For this reason, the study examines multiple datasets, across tasks and brain areas, and measures performance for a range of decoded variables. We also examine the impact of training-set-size (Figure 8a) and population size (solid traces in Fig. 8b, see R2’s next comment below).

There is one other source of variance one might be concerned about, but it is specific to the neuralnetwork approaches: different weight initializations might result in different performance. For this reason, each neural-network approach was trained ten times, with the average performance computed. The variability around this average was very small, and this is now stated in the Methods.

“For the neural networks, the training/testing procedure was repeated 10 times with different random seeds. For most behavioral variables, there was very little variability in performance across repetitions. However, there were a few outliers for which variability was larger. Reported performance for each behavioral group is the average performance across the 10 repetitions to ensure results were not sensitive to any specific random initialization of each network.”

- For Figure 6, it is unclear whether the neuron-dropping process was repeated multiple times. If not, it should be since the results will be sensitive to which particular subsets of neurons were "dropped". In this case, the results presented in Figure 6 should include error bars to describe the variability in the model performance for each decoder considered.

A good point. The results in Figure 8 (previously Figure 6) were computed by averaging over the removal of different random subsets of neurons (50 subsets per neuron count), just as the reviewer requests. The figure has been modified to include the standard deviation of performance across these 50 subsets. The legend clarifies how this was done.

Reviewer #3 (Recommendations For The Authors):

Other comments:

(1) [Line 185-188] The authors argue that in a 100-dimensional space with 10 possible discretized values, 10^100 potential neural states need to be computed. But I am not clear on this. This argument seems to hold only in the absence of a model (as in MINT). For a model, e.g., Kalman filter or AutoLFADS, information is encoded in the latent state. For example, a simple Kalman filter for a linear model can be used for efficient inference. This 10^100 computation isn't a general problem but seems MINT-specific, please clarify.

We agree this section was potentially confusing. It has been rewritten. We were simply attempting to illustrate why maximum likelihood computations are challenging without constraints. MINT simplifies this problem by adding constraints, which is why it can readily provide data likelihoods (and can do so using a Poisson model). The rewritten section is below:

“Even with 1000 samples for each of the neural trajectories in Figure 3, there are only 4000 possible neural states for which log-likelihoods must be computed (in practice it is fewer still, see Methods). This is far fewer than if one were to naively consider all possible neural states in a typical rate- or factor-based subspace. It thus becomes tractable to compute log-likelihoods using a Poisson observation model. A Poisson observation model is usually considered desirable, yet can pose tractability challenges for methods that utilize a continuous model of neural states. For example, when using a Kalman filter, one is often restricted to assuming a Gaussian observation model to maintain computational tractability “

(2) [Figure 6b] Why do the authors set the dropped neurons to zero in the "zeroed" results of the robustness analysis? Why not disregard the dropped neurons during the decoding process?

We agree the terminology we had used in this section was confusing. We have altered the figure and rewritten the text. The following, now at the beginning of that section, addresses the reviewer’s query:

“It is desirable for a decoder to be robust to the unexpected loss of the ability to detect spikes from some neurons. Such loss might occur while decoding, without being immediately detected. Additionally, one desires robustness to a known loss of neurons / recording channels. For example, there may have been channels that were active one morning but are no longer active that afternoon. At least in principle, MINT makes it very easy to handle this second situation: there is no need to retrain the decoder, one simply ignores the lost neurons when computing likelihoods. This is in contrast to nearly all other methods, which require retraining because the loss of one neuron alters the optimal parameters associated with every other neuron.”

The figure has been relabeled accordingly; instead of the label ‘zeroed’, we use the label ‘undetected neuron loss’.

(3) Authors should provide statistical significance on their results, which they already did for Fig. S3a,b,c but missing on some other figures/places.

We have added error bars in some key places, including in the text when quantifying MINT’s performance in the context of Figure 4. Importantly, error bars are only as meaningful as the source of error they assess, and there are reasons to be careful given this. The standard method for putting error bars on performance is to resample trials, which is indeed what we now report. These error bars are very small. For example, when decoding horizontal velocity for the MC_Maze dataset, the correlation between MINT’s decode and the true velocity had a mean and SD of the sampling distribution of 0.963 +/- 0.001. This means that, for a given dataset and target variable, we have enough trials/data that we can be quite certain of how well MINT performs. However, we want to be careful not to overstate this certainty. What one really wants to know is how well MINT performs across a variety of datasets, brain areas, target variables, neuron counts, etc. It is for this reason that we make multiple such comparisons, which provides a more valuable view of performance variability.

For Figure 7, error bars are unavailable. Because this was a benchmark, there was exactly one test-set that was never seen before. This is thus not something that could be resampled many times (that would have revealed the test data and thus invalidated the benchmark, not to mention that some of these methods take days to train). We could, in principle, have added resampling to Figure 5. In our view it would not be helpful and could be misleading for the reasons noted above. If we computed standard errors using different train/test partitions, they would be very tight (mostly smaller than the symbol sizes), which would give the impression that one can be quite certain of a given R^2 value. Yet variability in the train/test partition is not the variability one is concerned about in practice. In practice, one is concerned about whether one would get a similar R^2 for a different dataset, or brain area, or task, or choice of decoded variable. Our analysis thus concentrated on showing results across a broad range of situations. In our view this is a far more relevant way of illustrating the degree of meaningful variability (which is quite large) than resampling, which produces reassuringly small but (mostly) irrelevant standard errors.

Error bars are supplied in Figure 8b. These error bars give a sense of variability across re-samplings of the neural population. While this is not typically the source of variability one is most concerned about, for this analysis it becomes appropriate to show resampling-based standard errors because a natural concern is that results may depend on which neurons were dropped. So here it is both straightforward, and desirable, to compute standard errors. (The fact that MINT and the Wiener filter can be retrained many times swiftly was also key – this isn’t true of the more expressive methods). Figure S1 also uses resampling-based confidence intervals for similar reasons.

(4) [Line 431-437] Authors state that MINT outperforms other methods with the PSTH R^2 metric (trial-averaged smoothed spikes for each condition). However, I think this measure may not provide a fair comparison and is confounded because MINT's library is built using PSTH (i.e., averaged firing rate) but other methods do not use the PSTH. The author should clarify this.

The PSTH R^2 metric was not created by us; it was part of the Neural Latents Benchmark. They chose it because it ensures that a method cannot ‘cheat’ (on the Bits/Spike measure) by reproducing fine features of spiking while estimating rates badly. We agree with the reviewer’s point: MINT’s design does give it a potential advantage in this particular performance metric. This isn’t a confound though, just a feature. Importantly, MINT will score well on this metric only if MINT’s neural state estimate is accurate (including accuracy in time). Without accurate estimation of the neural state at each time, it wouldn’t matter that the library trajectory is based on PSTHs. This is now explicitly stated:

“This is in some ways unsurprising: MINT estimates neural states that tend to resemble (at least locally) trajectories ‘built’ from training-set-derived rates, which presumably resemble test-set rates. Yet strong performance is not a trivial consequence of MINT’s design. MINT does not ‘select’ whole library trajectories; PSTH R2 will be high only if condition (c), index (k), and the interpolation parameter (α) are accurately estimated for most moments.”

Version 2

Abstract

Decoders for brain-computer interfaces (BCIs) assume constraints on neural activity, chosen to reflect scientific beliefs while yielding tractable computations. Recent scientific advances suggest that the true constraints on neural activity, especially its geometry, may be quite different from those assumed by most decoders. We designed a decoder, MINT, to embrace statistical constraints that are potentially more appropriate. If those constraints are accurate, MINT should outperform standard methods that explicitly make different assumptions. Additionally, MINT should be competitive with expressive machine learning methods that can implicitly learn constraints from data. MINT performed well across tasks, suggesting its assumptions are well-matched to the data. MINT outperformed other interpretable methods in every comparison we made. MINT outperformed expressive machine learning methods in 37 of 42 comparisons. MINT’s computations are simple, scale favorably with increasing neuron counts, and yield interpretable quantities such as data likelihoods. MINT’s performance and simplicity suggest it may be a strong candidate for many BCI applications.

Introduction

Brain-computer interfaces (BCIs) seek to estimate target variables, in real time, from observations of neural spiking activity. Target variables may correspond to prosthetic motion [1–6], muscle activity [7–10], cursor control [11–17], navigation [18–20], speech [21–26], handwriting [27], or cognitive states [28–32]. Alternatively, one may estimate the state of the brain, a latent variable, for clinical monitoring [33], data visualization [34], or closed-loop neurally contingent experiments [35]. Because neural spiking is noisy, target variables must be estimated from patterns of spikes that were unobserved during training. Doing so requires assumptions.

Figure 1a illustrates a set of common assumptions. In this view, the primary objects are neural firing rates (which determine the probability of spiking) and a manifold [36–39] that constrains the possible values of the ‘neural state’: a vector containing the rate of every neuron. While that manifold is presumed to be nonlinear, a linear approximation – i.e. a subspace – may be acceptable for practical purposes [38]. Within this subspace, there exist neural dimensions where activity correlates with target variables such as hand velocity, providing a basis for estimating those variables. Generalization (e.g. [13, 14, 40]) is thought to rely on two features. First, if distributions of behavioral variables overlap across tasks, neural-state distributions will also overlap [39]. Second, out-of-distribution generalization can leverage the reliable correlation between behavioral variables and neural activity.

Two perspectives on the structure of neural activity during motor tasks.

a) Illustration of a traditional perspective. Each neural state (colored dots) is an N-dimensional vector of firing rates. States are limited to a manifold that can be usefully approximated by a subspace. In this illustration, neural states are largely limited to a two-dimensional subspace within the full three-dimensional firing-rate space. Within that restricted subspace, there exist neural dimensions where neural activity correlates with to-be-decoded variables. BCI decoding leverages knowledge of the subspace and of those correlations. If two tasks have similar motor outputs at two particular moments, the corresponding neural states will also be similar, aiding generalization. b) An emerging perspective. Neural activity is summarized by neural factors (one per axis), with each neuron’s firing rate being a simple function of the factors. Factors may be numerous (many dozens or even hundreds), resulting in a high-dimensional subspace of neural activity. Yet most of that space is unoccupied; factor-trajectories are heavily constrained by the underlying dynamics (gray arrows), such that most states are never visited. The order in which states are visited is also heavily sculpted by dynamics; e.g. trajectories rarely ‘swim upstream’. The primary constraint is thus not a subspace, but a sparse manifold where activity largely flows in specific directions. Different tasks may often (but not always) employ different dynamics and thus different regions of the manifold. Because most aspects of neural trajectories fall in the null space of outgoing commands, there are no directions in neural state space that consistently correlate with kinematic variables, and distant states may correspond to similar motor outputs.

Decoding methods frequently assume many or all of these features. For example, the classic population vector [11, 41, 42] assumes a two- or three-dimensional manifold (for reaches in two or three physical dimensions) and leverages neural dimensions where activity correlates with hand velocity. Similarly, nearly all ‘interpretable’ methods (those that make explicit assumptions rather than learning them from training data) assume neural dimensions where activity correlates – perhaps incidentally, but at least usefully – with target variables. Such methods often assume additional constraints on neural activity related to structure in behavior [16, 17, 43–49]. E.g. a Kalman filter can embody the assumption that velocity and position are dynamically linked [13, 50], while Sani et al. [51] leveraged the assumption that only a subspace within a low-dimensional neural state is relevant to behavior.

Although the perspective in Figure 1a has been useful, its assumptions may be true only locally [52] and perhaps not even then. Under an alternative perspective (Fig. 1b), the primary objects are neural factors (which determine each neuron’s instantaneous probability of spiking) and a flow-field (gray arrows) governing factor-state trajectories. Within the space of potential factor states, the flow-field ensures that few are visited. The resulting manifold is complex, and may or may not be locally flat in any useful sense. For example, when cycling at different speeds, the manifold is an approximately seven-dimensional tube (as in Figure 1b, green) whose long-axis corresponds to cycling speed [53]. Many locations in factor-space are never visited, including the void within the tube. The manifold is thus sparse – a minority of factor-states are observable – even though individual-neuron responses are rarely sparse in the traditional sense (most neurons show time-varying activity during most movements). It is presumed that different tasks (or subtasks) may require different computations, and thus employ different regions of factor space with different local flow-fields. This may result in a complex and extremely sparse manifold. Some factors may, in some regions of the manifold, correlate incidentally with external variables such as velocity. Yet this is not expected to be consistently true.

The assumptions in Figure 1b accord with the hypothesis of factor-level dynamics [54–56] that sculpt computation-specific neural geometries [53, 57–63]. As a simple example, DePasquale et al. [55] and colleagues constructed a network of spiking neurons to generate muscle activity during the cycling task [61]. Network dynamics produced a limit cycle in 12-dimensional factor space, with any deviations being swiftly corrected by the flow-field. Thus, although the data occupy a sizable subspace, the manifold consists of only those states near the limit cycle. When the network was trained to both cycle and reach, it did so using task-specific dynamics in different neural dimensions, resulting in a sparse, complex manifold with task-specific sub-regions.

Figure 1b also accords with analyses that focused on capturing trajectories within higher-dimensional subspaces [64–67], with the assumption of highly nonlinear mappings between neural activity and behavioral variables [68, 69], and with the finding that locally linear relationships between activity and behavior are prominent during some tasks but not others [20]. It agrees with the proposal that factors (though less numerous than neurons) are plentiful [70–74], with different tasks and task-epochs often using different factors [20, 55, 75–79] (the ‘flexibility-via-subspace’ hypothesis [56]). The assumption of a strong flow-field agrees with empirical limitations on BCI-generated trajectories [80, 81], with the ability of training to ‘open up’ previously unused degrees of freedom [82], and with the finding that decoding is aided by assuming neural-state dynamics [83, 84]. More broadly, low trajectory tangling [53, 61] – a feature of motor cortex activity in most tasks – constrains neural trajectories in ways that imply many of the features in Figure 1b and thus argues for approximating the manifold using collections of trajectories rather than a subspace [66, 85].

There is thus a potential mismatch between the structure of the data and the assumptions made by traditional interpretable decoders. This may be one reason why interpretable methods are often outperformed by expressive machine-learning methods [27, 86, 87]. Expressive methods [21, 23–27, 84, 86–93] may be able to implicitly learn, during training, many of the constraints illustrated in Figure 1b. Motivated by these considerations, we constructed a novel algorithm, Mesh of Idealized Neural Trajectories (MINT), that explicitly leverages those constraints. MINT takes a trajectory-centric view, where a complicated manifold is approximated using previously observed trajectories and interpolations between them. MINT abandons any notion of neural dimensions that reliably correlate with behavioral variables. Instead, MINT creates a direct correspondence between neural and behavioral trajectories, allowing it to capture highly nonlinear relationships. These relationships can be task-specific if the data so argue. One might have expected that mathematical tractability would be compromised by embracing the unusual assumptions in Figure 1b. Yet the requisite computations are surprisingly simple. It is also straightforward to decode a variety of behavioral variables – even those that do not correlate with neural activity – and to estimate the neural state.

MINT’s performance confirms that there are gains to be made by building decoders whose assumptions match a different, possibly more accurate view of population activity. At the same time, our results suggest fundamental limits on decoder generalization. Under the assumptions in Figure 1b, it will sometimes be difficult or impossible for decoders to generalize to not-yet-seen tasks. We found that this was true regardless of whether one uses MINT or a more traditional method. This finding has implications regarding when and how generalization should be attempted.

Results

We applied MINT to a total of nine datasets, recorded from primates performing a variety of motor, sensory, and cognitive tasks. Each dataset consisted of simultaneous neural recordings and behavioral variables. We used MINT to decode, based on spiking observations, behavioral variables and the neural state. A central goal was to compare MINT’s decoding performance with existing ‘interpretable’ methods (methods such as the Kalman filter that make explicit assumptions regarding data constraints) and with ‘expressive’ methods (e.g. neural networks that can learn constraints from training data). Current interpretable methods make assumptions largely aligned with Figure 1a. Thus, if Figure 1b better describes the data, MINT should consistently outperform other interpretable methods. Highly expressive methods can, given enough training data, implicitly learn a broad range of constraints, including those in Figure 1a, Figure 1b, or some other yet-to-be imagined scenario. This provides a strong test of the assumptions in Figure 1b, because MINT should perform well if those assumptions are correct. Indeed, if Figure 1b is correct, MINT’s performance should be similar to that of highly expressive methods, and may even be better in some cases because MINT can begin with good assumptions rather than having to learn them from data.

We also document properties of the data – directly and during decoding – that bear on the scientific question of whether Figure 1a or 1b makes better assumptions. A comprehensive characterization of neural trajectory geometry is beyond the scope of this study. Yet, whenever possible, we examine features of neural trajectories relevant to MINT’s assumptions. To do so, we begin by focusing on one dataset (MC_Cycle) recorded during a task in which a primate grasps a hand-pedal and moves it cyclically forward or backward to navigate a virtual environment.

Neural trajectories are locally stereotyped and sparsely distributed

During the cycling task, empirical neural trajectories are compatible with solutions found by neural networks trained to produce muscle commands as an output [61]. Those trajectories have features that tend to argue for the assumptions in Figure 1b. Many of these features follow from the property of ‘low trajectory tangling’: during movement, it is never the case that very different factor-trajectory directions are associated with similar factor states. Trajectories are thus locally stereotyped: two trajectories that pass near one another are moving in similar directions. When trajectories travel in dissimilar directions, maintaining low tangling requires that they avoid one another, spreading out into additional dimensions and becoming more sparse in factor space. For example, the elliptical trajectories during forward and backward cycling (Fig. 2a) appear to overlap in the dominant two dimensions. Yet despite this appearance, the corresponding neural trajectories never come close to crossing. For example, at the apparent crossing-point indicated by the gray disk, forward and backward trajectories are well-separated in dimension 3 (all dimensions use the same scale). This is true at all apparent crossing points, leading to forward-cycling and backward-cycling trajectories that are well-separated. Additionally, both limit cycles contain a void in their center that is never occupied. One might expect that such voids would fill in as additional behaviors are considered. For example, perhaps the interior of the limit cycle becomes occupied when stationary, or at slower speeds? Yet when stationary, neural states are far from the limit-cycle center (Fig. 2b). Cycling at different speeds involves neural trajectories that spread out in additional ‘new’ dimensions, and thus remain sparse [53]. This illustrates a key difference between the notion of manifold in Figure 1a and Figure 1b. In Figure 1a, if two distant neural states are both likely to be observed, then the state halfway between them is also reasonably likely to be observed. In Figure 1b, this will frequently not be true.

Properties of neural trajectories in motor cortex, illustrated for data recorded during the cycling task (MC_Cycle dataset).

a) Low tangling implies separation between trajectories that might otherwise be close. Top. Neural trajectories for forward (purple) and backward (orange) cycling. Trajectories begin 600 ms before movement onset and end 600 ms after movement offset. Trajectories are based on trial-averaged firing rates, projected onto two dimensions. Dimensions were hand-selected to highlight apparent trajectory crossings while also capturing considerable variance (11.5%, comparable to the 11.6% captured by PCs 3 and 4). Gray region highlights one set of apparent crossings. Bottom. Trajectories during the restricted range of times in the gray region, but projected onto different dimensions. The same scale is used in top and bottom subpanels. b) Examples of well-separated neural states (main panel) corresponding to similar behavioral states (inset). Colored trajectory tails indicate the previous 150 ms of states. Data from 7-cycle forward condition. c) Joint distribution of pairwise distances for muscle and neural trajectories. Analysis considered all states across all conditions. For both muscle and neural trajectories, we computed all possible pairwise distances across all states. Each muscle state has a corresponding neural state, from the same time within the same condition. Thus, each pairwise muscle-state distance has a corresponding neural-state distance. The color of each pixel indicates how common it was to observe a particular combination of muscle-state distance and neural-state distance. Muscle trajectories are based on seven z-scored intramuscular EMG recordings. Correspondence between neural and muscle state pairs included a 50 ms lag to account for physiological latency. Results were not sensitive to the presence or size of this lag. Neural and muscle distances were normalized (separately) by average pairwise distance. d) Same analysis for neural and kinematic distances (based on phase and angular velocity). Correspondence between neural and kinematic state pairs included a 150 ms lag. Results were not sensitive to the presence or size of this lag. e) Control analysis to assess the impact of sampling error. If two sets of trajectories (e.g. neural and kinematic) are isometric and can be estimated perfectly, their joint distribution should fall along the diagonal. To estimate the impact of sampling error, we repeated the above analysis comparing neural distances across two data partitions, each containing 15-18 trials/condition.

Neural and behavioral trajectories are non-isometric

A potential consequence of low-tangled trajectories is that similar behavioral outputs can be associated with distant neural states. We found that this was common. As a simple example, there exist many moments when two-dimensional velocity is matched between forward and backward cycling: e.g. at a 45° phase when cycling forward versus 225°when cycling backward. Yet forward and backward trajectories remain distant at all moments. One might suspect that this is simply because angular position is different, even though velocity is matched. Yet neural states can be quite different even when both position and velocity are matched. This point is illustrated (Fig. 2b) by the neural trajectory that unfolds when transitioning from stationary, to cycling forward seven times, to stationary again. This projection was chosen to highlight dimensions where the neural trajectory resembles the behavioral trajectory (inset). Nevertheless, neural and behavioral trajectories are far from isometric. Red circles indicate two neural states where the corresponding positions and angular velocities are nearly identical: cycling at ∼1.7 Hz with the hand approaching the cycle’s bottom. Despite this behavioral match, the neural states are distant. Gray squares indicate neural states that are distant even though, in both cases, the hand is stopped at the cycle’s bottom.

These observations suggest a many-to-one mapping from a neural manifold with one geometry to a behavioral manifold with a very different geometry. To quantify, we leveraged the fact that each neural distance (between a pair of neural states) has a corresponding behavioral distance, obtained for the same pair of conditions and times (allowing for a latency shift). If behavioral and neural geometries are similar, behavioral and neural distances should be strongly correlated. If geometries differ, a given behavioral distance could be associated with a broad range of neural distances. To establish a baseline that incorporates sampling error, we compared neural distances across two partitions. As expected, the joint distribution was strongly diagonal (Fig. 2e). In contrast, the joint distribution was non-diagonal when comparing neural versus muscle-population trajectories (Fig. 2c) and neural versus kinematic trajectories (Fig. 2d). Each behavioral distance (on the x-axis) was associated with a range of different neural distances (on the y-axis). This was true even when behavioral distances were small: for kinematic states with normalized distance <0.1, the corresponding neural states were as close as 0.02 and as far as 1.46. Thus, dissimilar neural states frequently correspond to similar behavioral states, as suggested by Figure 1b (see states indicated by arrows).

Importantly, the converse was not true. It was never the case that similar neural states corresponded to dissimilar behavioral states. Once past small values on the x-axis, the white regions in Figure 2c,d never extend all the way to the x-axis. For example, when kinematic distances were ∼2, neural distances were always at least 0.53. Thus, accurate decoding should be possible as long as the decoder can handle the many-to-one mapping of neural states to behavioral variables. In principle this might be accomplished linearly. Suppose neural trajectories mirrored behavioral trajectories but with additional non-behavior-encoding dimensions. A linear decoder could simply ignore those additional dimensions. Yet this will be suboptimal if the ignored dimensions capture the majority of response variance. The above observations thus argue that one may desire quite nonlinear decoding.

A potential concern regarding the analyses in Figure 2c,d is that they require explicit choices of behavioral variables: muscle population activity in Figure 2c and angular phase and velocity in Figure 2d. Perhaps these choices were misguided. Might neural and behavioral geometries become similar if one chooses ‘the right’ set of behavioral variables? This concern relates to the venerable search for movement parameters that are reliably encoded by motor cortex activity [70, 94–97]. If one chooses the wrong set of parameters (e.g. chooses muscle activity when one should have chosen joint angles) then of course neural and behavioral geometries will appear non-isometric. There are two reasons why this ‘wrong parameter choice’ explanation is unlikely to account for the results in Figure 2c,d. First, consider the implications of the left-hand side of Figure 2d. A small kinematic distance implies that angular position and velocity are nearly identical for the two moments being compared. Yet the corresponding pair of neural states can be quite distant. Under the concern above, this distance would be due to other encoded behavioral variables – perhaps joint angle and joint velocity – differing between those two moments. However, there are not enough degrees of freedom in this task to make this plausible. The shoulder remains at a fixed position (because the head is fixed) and the wrist has limited mobility due to the pedal design [61]. Thus, shoulder and elbow angles are almost completely determined by cycle phase. More generally, ‘external variables’ (positions, angles, and their derivatives) are unlikely to differ more than slightly when phase and angular velocity are matched. Muscle activity could be different because many muscles act on each joint, creating redundancy. However, as illustrated in Figure 2c, the key effect is just as clear when analyzing muscle activity. Thus, the above concern seems unlikely even if it can’t be ruled out entirely. A broader reason to doubt the ‘wrong parameter choice’ proposition is that it provides a vague explanation for a phenomenon that already has a straightforward explanation. A lack of isometry between the neural population response and behavior is expected when neural-trajectory tangling is low and output-null factors are plentiful [56, 61]. For example, in networks that generate muscle activity, neural and muscle-activity trajectories are far from isometric [53, 59, 61]. Given this straightforward explanation, and given repeated failures over decades to find the ‘correct’ parameters (muscle activity, movement direction, etc.) that create neural-behavior isometry, it seems reasonable to conclude that no such isometry exists.

We further explored the topic of isometry by considering pairs of distances. To do so, we chose two random neural states and computed their distance, yielding dneural1. We repeated this process, yielding dneural2. We then computed the corresponding pair of distances in muscle space (dmuscle1 and dmuscle2) and kinematic space (dkin1 and dkin2). We considered cases where dneural1 was meaningfully larger than (or smaller than) dneural2, and asked whether the behavioral variables had the same relationship; e.g. was dmuscle1 also larger than dmuscle2? For kinematics, this relationship was weak: across 100,000 comparisons, the sign of dkin1 − dkin2 agreed with dneural1 − dneural2 only 67.3% of the time (with 50% being chance). The relationship was much stronger for muscles: the sign of dmuscle1 − dmuscle2 agreed with dneural1 − dneural2 79.2% of the time, which is far more than expected by chance yet also far from what is expected given isometry (e.g. the sign agrees 99.7% of the time for the truly isometric control data in Figure 2e). Indeed there were multiple moments during this task when dneural1 was much larger than dneural2, yet dmuscle1 was smaller than dmuscle2. These observations are consistent with the proposal that neural trajectories resemble muscle trajectories in some dimensions, but with additional output-null dimensions that break the isometry [61].

Leveraging trajectories to estimate the manifold

The results above, along with other recent results, argue for the perspective in Figure 1b. In this view, the manifold of observable states is complicated and sparse. We thus designed MINT to approximate that manifold using the trajectories themselves, rather than their covariance matrix or corresponding subspace. Unlike a covariance matrix, neural trajectories indicate not only which states are likely, but also which state-derivatives are likely. If a neural state is near previously observed states, it should be moving in a similar direction. MINT leverages this directionality.

Training-set trajectories can take various forms, depending on what is convenient to collect. Most simply, training data might include one trajectory per condition, with each condition corresponding to a discrete movement. Alternatively, one might instead employ one long trajectory spanning many movements. Another option is to employ many sub-trajectories, each briefer than a whole movement. The goal is simply for training-set trajectories to act as a scaffolding, outlining the manifold that might be occupied during decoding and the directions in which decoded trajectories are likely to be traveling.

Because training-set trajectories are unlikely to sample the manifold with sufficient density (e.g. one may train using eight reach directions but wish to decode any direction), we designed MINT to interpolate during decoding. We use the term ‘mesh’ to describe the scaffolding created by the training-set trajectories and the interpolated states that arise at runtime. The term mesh is apt because, if MINT’s assumptions are correct, interpolation will almost always be local. If so, the set of decodable states will resemble a mesh, created by line segments connecting nearby training-set trajectories. However, this mesh-like structure is not enforced by MINT’s operations. Interpolation could, in principle, create state-distributions that depart from the assumption of a sparse manifold. For example, interpolation could fill in the center of the green tube in Figure 1b, resulting in a solid manifold rather than a mesh around its outer surface. However, this would occur only if spiking observations argued for it. As will be documented below, we find that essentially all interpolation is local. This is a useful fact, because it implies that the estimated neural state (during decoding) will rarely be far from previously observed neural states for which the corresponding behavioral states are known. MINT finds those behavioral states using a direct (and highly nonlinear) mapping between neural and behavioral states. MINT then decodes behavior using local linear interpolation.

Although the mesh is formed of stereotyped trajectories, decoded trajectories can move along the mesh in non-stereotyped ways as long as they generally obey the flow-field implied by the training data. This flexibility supports many types of generalization, including generalization that is compositional in nature. Other types of generalization – e.g. from the green trajectories to the orange trajectories in Figure 1b – are unavailable when using MINT and are expected to be challenging for any method (as will be documented in a later section).

Training and decoding using MINT

Training MINT requires computing a library of neural trajectories (Ω) and a corresponding library of behavioral trajectories (Φ). For ease of subsequent computation, neural trajectories are expressed in firing-rate space. MINT is agnostic regarding how neural trajectories are found. For example, one can estimate factor-trajectories, then convert to firing rates via a rectifying nonlinearity [55, 84]. When training data contain repeated trials per condition, it is simplest to use trial-averaging to compute each neuron’s rate. We use this approach to illustrate MINT’s operations during a center-out reaching task. Neural and behavioral trajectories (Fig. 3a) were learned from a training set containing repeated reaches (trials) to each target location. Each moment in that training data corresponds to a condition (c) and an index (k) indicating progress through the movement. For example, c = 3 corresponds to the purple reach condition and k = 240 indicates a moment 240 ms into that movement. There were many trials for this condition, and thus many measurements of neural activity and behavior for c = 3 and k = 240. By temporally filtering spikes and averaging across trials, one obtains a neural state containing each neuron’s estimated rate. If desired, firing-rate estimation can be further improved by a variety of means (see Methods). A similar process yields a behavioral state , which may contain position, velocity, or any other variables of interest.

Example training (top panel) and decoding (bottom panels) procedures for MINT, illustrated using four conditions from a reaching task.

a) Libraries of neural and behavioral trajectories are learned such that each neural state corresponds to a behavioral state b) Spiking observations are binned spike counts (20 ms bins for all main analyses). contains the spike count of each neuron for the present bin, t′, and τ ′ bins in the past. c) At each time step during decoding, the log-likelihood of observing computed for each and every state in the library of neural trajectories. Log-likelihoods decompose across neurons and time bins into Poisson log-likelihoods that can be queried from a precomputed lookup table. A recursive procedure (not depicted) further improves computational efficiency. d) Two candidate neural states (purple and blue) are identified. The first is the state within the library of trajectories that maximizes the log-likelihood of the observed spikes. The second state similarly maximizes that log-likelihood, with the restriction that the second state must not come from the same trajectory as the first (i.e. must be from a different condition). Interpolation identifies an intermediate state that maximizes log-likelihood. e) The optimal interpolation is applied to candidate neural states – yielding the final neural-state estimate – and their corresponding behavioral states – yielding decoded behavior. Despite utilizing binned spiking observations, neural and behavioral states can be updated at millisecond resolution (Methods).

As expected given results above, neural and behavioral trajectories differ during reaching: neural trajectories are stacked loops while position trajectories are arranged radially (velocity trajectories are also radial but return to center as the reach ends). These different neural and behavioral geometries might seem like an impediment to decoding, or to argue for a different choice of target behavioral variables. Yet because neural and behavioral trajectories share the same indices c and k, decoding is straightforward regardless of which variables one wishes to decode.

The parameters of MINT are the libraries of neural trajectories (Ω) and behavioral trajectories (Φ). The trajectories in Ω form a scaffolding that outlines the expected neural manifold during decoding. Ω also specifies the directionality with which decoded states are expected to traverse that manifold. On long timescales, decoded trajectories may be very different from any library trajectory. Yet decoded trajectories will typically be composed of events resembling those from the library, reflecting the local stereotype described above. Prior work has utilized stereotyped behavioral trajectories [43, 47] or a mixture of condition-specific behavioral trajectory models that capture trial-by-trial movement variability [49]. For tractability, those methods assumed low-dimensional neural states that are roughly isometric to arm velocities. With MINT, neural trajectories can take any geometry, are learned empirically, and are related to behavioral trajectories via c and k rather than by an assumed geometric isometry.

During decoding, spike-based observations are binned and counted (Fig. 3b, Δ = 20 ms for all analyses unless otherwise specified). Suppose we are at time-bin t′ within a decoding session. Recent spiking observations are denoted by , where τ ′ specifies the number of previous time bins retained in memory. The foundation of MINT is the computation of the log-likelihood of the data, , for a variety of neural states. The decoded state is chosen with the goal of maximizing that log-likelihood. Not all decoders involve data likelihood computations, but likelihoods are desirable when it is possible to obtain them. One typically wishes to decode the state that maximizes data likelihood, or maximizes a cost function that involves likelihoods across multiple states. Knowledge of likelihoods can also indicate the trustworthiness of decoding, as will be discussed in a subsequent section.

To decode, we first compute the log-likelihood that we would have observed for each neural state in the library (Fig. 3c). Importantly, this computation is performed for relatively few states. Even with 1000 samples for each of the neural trajectories in Figure 3, there are only 4000 possible neural states for which log-likelihoods must be computed (in practice it is fewer still, see Methods). This is far fewer than if one were to naively consider all possible neural states in a typical rate- or factor-based subspace. It thus becomes tractable to compute log-likelihoods using a Poisson observation model. A Poisson observation model is usually considered desirable, yet can pose tractability challenges for methods that utilize a continuous model of neural states. For example, when using a Kalman filter, one is often restricted to assuming a Gaussian observation model to maintain computational tractability.

The library, Ω, makes it simple to compute data likelihoods that take spike-history into account. Without the constraint of a flow-field, a great many past trajectories could have led to the present neural state. Computing the likelihood of given all these possible histories, may be impractical during real-time decoding. Yet given a strong flow-field, such histories will be similar on short timescales. For example, if we are presently 240 ms into the purple trajectory in Figure 3a (i.e. c = 3 and k = 240) then we know (approximately) the rate of every neuron over the last few hundred milliseconds. One can thus compute the log-likelihood of the observed pattern of spikes, over the last few hundred milliseconds, given that we are presently at c = 3 and k = 240. This computation is aided by the fact that, under a Poisson model, spike counts in non-overlapping bins are conditionally independent given knowledge of the underlying rate. Similarly, spike counts for different neurons are conditionally independent given knowledge of their rates. Thus, the log-likelihood of , for a particular current neural state, is simply the sum of many individual log-likelihoods (one per neuron and time-bin). Each individual log-likelihood depends on only two numbers: the firing rate at that moment and the spike count in that bin. To simplify online computation, one can precompute the log-likelihood, under a Poisson model, for every plausible combination of rate and spike-count. For example, a lookup table of size 2001× 21 is sufficient when considering rates that span 0-200 spikes/s in increments of 0.1 spikes/s, and considering 20 ms bins that contain at most 20 spikes (only one lookup table is ever needed, so long as its firing-rate range exceeds that of the most-active neuron at the most active moment in Ω). Now suppose we are observing a population of 200 neurons, with a 200 ms history divided into ten 20 ms bins. For each library state, the log-likelihood of the observed spike-counts is simply the sum of 200 × 10 = 2000 individual log-likelihoods, each retrieved from the lookup table. In practice, computation is even simpler because many terms can be reused from the last time bin using a recursive solution (Methods). This procedure is lightweight and amenable to real-time applications. The assumption of locally stereotyped trajectories also enables neural states (and decoded behaviors) to be updated between time bins. While waiting for the next set of spiking observations to arrive, MINT simply assumes that each neural state advances deterministically according to the flow-field.

To decode stereotyped trajectories, one could simply obtain the maximum-likelihood neural state from the library, then render a behavioral decode based on the behavioral state with the same values of c and k. This would be appropriate for applications in which conditions are categorical, such as typing or handwriting. Yet in most cases we wish for the trajectory library to serve not as an exhaustive set of possible states, but as a scaffolding for the mesh of possible states. MINT’s operations are thus designed to estimate any neural trajectory – and any corresponding behavioral trajectory – that moves along the mesh in a manner generally consistent with the trajectories in Ω. To illustrate, suppose the subject made a reach that split the difference between the purple and blue reach conditions in Figure 3a. MINT identifies two candidate states with high likelihoods, along different neural trajectories (Fig. 3d) in the library. MINT then interpolates between these high-likelihood states, and determines whether there exists a higher-likelihood state between them. MINT does so by considering a line segment of neural states parameterized by α, ranging from 0 (at the purple state) to 1 (at the blue state). Because the log-likelihood of the observed spikes is a concave function of α, the optimal neural state can be rapidly identified online using Newton’s method. The decoded behavioral output is then produced by interpolation between the corresponding two behavioral states in Φ, using the same α associated with the optimal neural state (Fig. 3e).

The value of α may change with time. For example, this could occur if the true behavioral trajectory began close to the purple trajectory but became progressively more similar to the blue trajectory. The library trajectories that MINT interpolates between can also change with time. Thus, interpolation allows considerable flexibility. Not only is one not ‘stuck’ on a trajectory from Φ, one is also not stuck on trajectories created by weighted averaging of trajectories in Φ. For example, if cycling speed increases, the decoded neural state could move steadily up a scaffolding like that illustrated in Figure 1b (green). In such cases, the decoded trajectory might be very different in duration from any of the library trajectories. Thus, one should not think of the library as a set of possible trajectories that are selected from, but rather as providing a mesh-like scaffolding that defines where future neural states are likely to live and the likely direction of their local motion. The decoded trajectory may differ considerably from any trajectory within Ω. Nevertheless, individual decoded states are (empirically) close to states within Ω, as will be discussed in a subsequent section. That proximity makes it reasonable to use the linear interpolation step when decoding behavior.

Behavioral decoding when the ground-truth is known

We begin by assessing MINT’s performance when applied to the activity of an artificial recurrent network of spiking neurons, trained to generate the empirical activity of the deltoid [55]. Doing so provides a situation where the ground truth is known: we know the network’s true output and that spiking is approximately Poisson, resembling that of cortical neurons [55]. The network performed two tasks – reaching and cycling – governed by different local flow-fields in different regions of factor-space (as proposed by the flexibility-via-subspace hypothesis [56]). Indeed, the neural dimensions occupied during cycling are orthogonal to those occupied during reaching (Fig. 4b). There are also moments when behavioral output – instantaneous muscle activity – is identical across tasks even though the underlying neural state is very different, in agreement with the empirical many-to-one mapping (Fig. 2c,d). These features provide a test of whether MINT can handle geometry similar to that proposed in Figure 1b.

Examples of behavioral decoding provided by MINT for one simulated dataset and four empirically recorded datasets.

All decoding is causal; only spikes from before the decoded moment are used. a) MINT was applied to spiking data from an artificial spiking network. That network was trained to generate posterior deltoid activity and to switch between reaching and cycling tasks. Based on spiking observations, MINT approximately decoded the true network output at each moment. ‘R3’, ‘R4’, and ‘R6’ indicate three different reach conditions. ‘C’ indicates a cycling bout. MINT used no explicit task-switching, but simply tracked neural trajectories across tasks as if they were conditions within a task. b) Illustration of the challenging nature, from a decoding perspective, of network trajectories. Trajectories are shown for two dimensions that are strongly occupied during cycling. Trajectories for the 8 reaching conditions (pink) are all nearly orthogonal to the trajectory for cycling (brown) and thus appear compressed in this projection. c) Decoded behavioral variables (green) compared to actual behavioral variables (black) across four empirical datasets. MC_Cycle and MC_RTT show 10 seconds of continuous decoding. MC_Maze and Area2_Bump show randomly selected trials, demarcated by vertical dashed lines. d) Distribution of distances between each decoded neural state and the nearest state in the library Ω (green), for the MC_Cycle dataset. To provide a reference, we drew neural states from a Gaussian distribution whose mean and covariance matched the library states, and computed distances to the library states (purple). This emulates what would occur if the manifold were defined by the data covariance and associated neural subspace. The green distribution mostly contains non-zero values, but is much closer to zero than the purple distribution. Thus, MINT rarely decodes a neural state that exactly matches a library state, but most decoded states are close to the scaffolding provided by the library states. Neural distances are normalized by the average pairwise distance between library states.

Decoded muscle activity (Fig. 4a, green) was virtually identical to network output (black). This was true during reaches (R3, R4, R6) and bouts of cycling (C). Decoding seamlessly transitioned between tasks. Decoding R2 was.968 over ∼7.1 minutes of test trials based on ∼4.4 minutes of training data. This demonstrates that MINT performs well when data match its motivating assumptions, and also that it may be particularly useful in situations where decoding must automatically switch amongst tasks.

We leveraged this simulated data to compare MINT with a biomimetic decoder whose operations (unlike MINT) mimic the manner in which the true output is produced. The network’s true output is simply a weighted sum of the spikes of 1200 neurons. We thus employed a linear readout (learned via ridge regression) leveraging all 1200 neurons. The resulting decode was very good: R2 = .982 (and would have been unity if readout weights were inferred from connectivity rather than learned from activity). However, the decoding dimension – just like the actual output dimension – captured exceedingly little (∼2%) of the total variance of network activity, a feature that is likely true of many real networks as well [61]. Biomimetic decoding may therefore suffer when not all neurons are recorded. Indeed, when only using 5% of neurons in the network, performance of the biomimetic decoder dropped from R2 = 0.982 to R2 = 0.773. In contrast, MINT’s performance declined only slightly from R2 = 0.968 to R2 = 0.967. This illustrates that, even when biomimetic decoding is possible, it may be outperformed by methods that leverage all neural dimensions rather than just output dimensions. MINT can also decode variables – e.g. kinematics – that are not literally encoded by network activity. These observations call into question an often-implicit assumption: that the activity-to-behavior ‘model’ used during decoding should attempt to approximate the true activity-to-behavior mapping. This is presumably true if a great many neurons can be recorded and the true output of the neural population is the quantity one wishes to decode – yet that is often not the case in real-world BCI applications.

Behavioral decoding across multiple datasets

We investigated MINT’s decoding performance across four datasets in which primates performed different motor tasks. All datasets included simultaneously collected spiking activity and behavioral measurements. The MC_Cycle dataset involved cycling a hand-held pedal and provided the motor cortex data in Figure 2. The MC_Maze dataset involved recordings from motor cortex during straight and curved reaches [98]. The Area2_Bump dataset involved recordings from somatosensory cortex (area 2) during active and passive center-out-reaches [99]. The MC_RTT dataset involved motor cortex recordings during point-to-point reaches with no condition structure [90]. All decoding was offline, to facilitate comparison across methods using matched data. Many of these datasets are suspected to involve trajectories with properties matching those in Figure 1b. However, unlike for the network above, this is an empirical inference and is not known to be true for all datasets. It was thus unclear, a priori, how well MINT would perform. All decoding was causal; i.e. based on spikes from before the decoded time. All decoding examples and performance quantifications utilized held-out test sets. Selection of MINT’s (few) hyperparameters is documented in a later section.

For the MC_Cycle dataset (Fig. 4c, top row), MINT’s decode tracked pedal phase and angular velocity (also see Supp. Video). When the pedal was stationary, erroneous deviations from non-zero angular velocity sometimes occurred (e.g. Figure 4c, just before the middle cycling bout) but were brief and rare: decoded angular speed was <0.1Hz for ∼ 99% of non-moving times. Transitions between rest and cycling occurred with near-zero latency. During movement, angular velocity was accurately decoded and angular phase was decoded with effectively no net drift over time. This is noteworthy because angular velocity on test trials never perfectly matched any of the trajectories in Φ. Thus, if decoding were restricted to a library trajectory, one would expect growing phase discrepancies. Yet decoded trajectories only need to locally (and approximately) follow the flow-field defined by the library trajectories. Based on incoming spiking observations, decoded trajectories speed up or slow down (within limits).

This decoding flexibility presumably relates to the fact that the decoded neural state is allowed to differ from the nearest state in Ω. To explore, we computed, for each moment in the MC_Cycle dataset, the distance between the decoded neural state and the closest library state. Most distances were non-zero (Fig. 4d, green distribution). At the same time, distances were much smaller than expected if decoded states obeyed a Gaussian with the empirical covariance (purple). These small distances are anticipated under the view that the empirical manifold (which produces the spiking observations that drive decoding) differs from the distribution defined by the data’s covariance matrix.

For the MC_Maze dataset (Fig. 4c, second row), decoded velocity tracked actual reach velocity across a variety of straight and curved reaches (108 conditions). Decoded and actual velocities were highly correlated (Fig. 4c, middle row, Rx = .963 ±.001, Ry = .950 ±.002; ranges indicate standard errors, computed by resampling test trials with replacement). For comparison with the other reach-like datasets below, we computed the mean absolute error (MAE) between decoded and actual velocity. Absolute errors were low (MAEx = 4.5 ± 0.1 cm/s, MAEy = 4.7 ± 0.1 cm/s), and constituted only ∼2% of x- and y-velocity ranges (224.8 and 205.7 cm/s, respectively).

For the Area2_Bump dataset (Fig. 4c, third row), correlations with horizontal and vertical hand velocity were also high: Rx = .939 ±.007, Ry = .895 ±.015. Mean absolute errors were small (MAEx = 4.1 ± 0.2 cm/s, MAEy = 4.4 0.2 cm/s) and similar to those for the MC_Maze dataset. In relative terms, errors were slightly larger for Area2_Bump versus MC_Maze because the range of velocities was smaller (95.8 and 96.8 cm/s for x- and y-velocity). Errors are thus slightly clearer in the traces plotted in Figure 4c (compare second and third rows).

Decoding was acceptable, but noticeably worse, for the MC_RTT dataset (Fig. 4c, bottom row). In part this can be attributed to this dataset’s slower reaches (86.6 and 61.3 cm/s x- and y-velocity ranges). When analyzing only movement periods (reach speed >1 cm/s), the median speed across all movement times was 38.1, 16.8, and 5.9 cm/s for MC_Maze, Area2_Bump, and MC_RTT, respectively. Thus, despite being modest in absolute terms (MAEx = 2.7 ± 0.1 cm/s, MAEy = 2.1 ± 0.1 cm/s), errors resulted in noticeably weakened correlations between decoder output and behavior: Rx = .785 ±.013, Ry = .843 ±.013. As will be discussed below, every decode method achieved its worst estimates of velocity for the MC_RTT dataset. In addition to the impact of slower reaches, MINT was likely impacted by training data that made it challenging to accurately estimate library trajectories. Due to the lack of repeated trials, MINT used AutoLFADS to estimate the neural state during training. In principle this should work well. In practice AutoLFADS may have been limited by having only ∼ 10 minutes of training data. Because the random-target task involved more variable reaches, it may also have stressed the ability of all methods to generalize, perhaps for the reasons illustrated in Figure 1b. To aid interpolation amongst library trajectories (and possibly improve generalization), for this dataset we allowed MINT to consider multiple pairwise interpolations when finding the maximum-likelihood state. This aided performance, but only very modestly. We return to the issue of generalization in a later section.

Comparison to other decoders

We compared MINT with four other decode algorithms: the Kalman filter, Wiener filter, feedforward neural network, and recurrent neural network (GRU). The Kalman filter and Wiener filter are historically popular BCI algorithms [1, 50] that are computationally simple and make linear assumptions. The feedforward neural network and GRU are expressive nonlinear function approximators that have been proposed for neural decoding [87, 100] and can be implemented causally. At each decoding time, both the feedforward neural network and the GRU leverage recent spiking history. The primary difference between the two is whether that history is processed all at once (feedforward network) or sequentially (GRU).

Each method was evaluated on each of the four datasets and was asked to decode multiple behavioral variables on held-out test sets (Fig. 5). All decoders were provided with a trailing window of spiking observations, binned in 20 millisecond increments. The length of this trailing window was optimized separately for each decoder and dataset (and, like all hyperparameters, was chosen using validation data). For non-MINT decoders, hyperparameters were tuned, using Bayesian optimization, to maximize performance [101]. MINT’s few hyperparameters were less extensively optimized. For example, window length was optimized just once for a given dataset for MINT, rather than separately for each set of behavioral variables as was done for most other decoders (excepting the Kalman filter). Minimal hyperparameter optimization embraces an unusual and useful aspect of MINT: there is no need to retrain if one wishes to decode a different behavioral variable. Once the neural state has been estimated, all behavioral variables are readily decodable. In principle, less-extensive optimization could have put MINT at a relative disadvantage. In practice, MINT is robust across reasonable hyperparameter choices (Fig. S1); any disadvantage is thus likely very slight. By necessity and desire, all comparisons were made offline, enabling benchmarked performance across a variety of tasks and decoded variables, where each decoder had access to the exact same data and recording conditions.

Comparison of decoding performance, for MINT and four additional algorithms, for four datasets and multiple decoded variables.

On a given trial, MINT decodes all behavioral variables in a unified way based on the same inferred neural state. For non-MINT algorithms, separate decoders were trained for each behavioral group (with the exception of the Kalman filter, which used the same model to decode position and velocity). E.g. separate GRUs were trained to output position and velocity in MC_Maze. Parentheticals indicate the number of behavioral variables within a group. E.g. ‘position (2)’ has two components: x- and y-position. R2 is averaged across behavioral variables within a group. ‘Overall’ plots performance averaged across all behavioral groups. R2 for feedforward networks and GRUs are additionally averaged across runs for 10 random seeds. The Kalman filter is traditionally utilized for position- and velocity-based decoding and was therefore only used to predict these behavioral groups. Accordingly, the ‘overall’ category excludes the Kalman filter for datasets in which the Kalman filter did not contribute predictions for every behavioral group. Results are based on the following numbers of training / test trials: MC_Cycle (174 train, 99 test), MC_Maze (1721 train, 574 test), Area2_Bump (272 train, 92 test), MC_RTT (810 train, 268 test). Vertical offsets and vertical ticks are used to increase visibility of data when symbols are close due to similar values.

There were 15 total ‘behavioral groups’ across the four datasets. For example, MC_Maze yielded two behavioral groups: position and velocity. We computed performance, for each group, by averaging across group members (e.g. across horizontal and vertical velocity). For every behavioral group, across all datasets, MINT’s performance improved upon the ‘interpretable’ methods: the Kalman and Wiener filters. MINT’s performance was typically comparable to, and often better than, that of the expressive GRU and feedforward neural network. MINT yielded the best performance, of all methods, for 11 of 15 behavioral groups. For 3 of the 4 behavioral groups where MINT’s performance was not the highest, the deficit relative to the best method was negligible: ΔR2 < .004. Thus, in only 1 of 15 cases was MINT at a disadvantage relative to another decoder.

When decoding velocity, MINT provided a small but noticeable improvement over the expressive methods for two datasets (MC_Maze and Area2_Bump), was modestly worse for one dataset (MC_RTT), and was involved in essentially a three-way tie for one dataset (MC_Cycle). For all datasets where position was a behavioral group, MINT and both expressive methods provided extremely similar and very accurate performance, considerably better than that of the Wiener and Kalman filter. MINT tended to display the largest improvements, relative to the next-best decoder, when decoding muscle-related variables (EMG, force, muscle velocity, and muscle length) and phase. This is noteworthy because MINT used the same hyperparameters in all cases and was not re-optimized or re-trained across variables (unlike the other methods). Thus, the same MINT operations that produce a velocity decode naturally produce a position decode and a muscle-length decode.

For each dataset, we computed overall performance by averaging across behavioral groups. MINT had the best overall performance for three of four datasets: MC_Cycle, MC_Maze, and Area2_Bump. For two datasets (MC_Cycle and MC_Maze) MINT considerably outperformed the Kalman and Wiener filter, and had a tiny advantage over the expressive feedforward network and GRU. For the third (Area2_Bump), MINT’s performance was noticeably better than all other methods.

The only dataset where MINT did not perform the best overall was the MC_RTT dataset, where it was outperformed by the feedforward network and GRU. As noted above, this may relate to the need for MINT to learn neural trajectories from training data that lacked repeated trials of the same movement (a design choice one might wish to avoid). Alternatively, the less-structured MC_RTT dataset may strain the capacity to generalize; all methods experienced a drop in velocity-decoding R2 for this dataset compared to the others. MINT generalizes somewhat differently than other methods, and may have been at a modest disadvantage for this dataset. A strong version of this possibility is that perhaps the perspective in Figure 1a is correct, in which case MINT might struggle because it cannot use forms of generalization that are available to other methods (e.g. generalization based on neuron-velocity correlations). This strong version seems unlikely; MINT continued to significantly outperform the Wiener and Kalman filters, which make assumptions aligned with Figure 1a.

MINT’s generally high level of decoding performance is supported by Bayesian computations: the estimation of data likelihoods and the selection of a decoded state that maximizes likelihood. A natural question is thus whether a simpler Bayesian decoder would have yielded similar results. We explored this possibility by testing a Naïve Bayes regression decoder [87] using the MC_Maze dataset. This decoder performed poorly, especially when decoding velocity (R2 = .688 and .093 for hand position and velocity, respectively), indicating that the specific modeling assumptions that differentiate MINT from a naive Bayesian decoder are important drivers of MINT’s performance. We also investigated the possibility that MINT gained its performance advantage simply by having access to trial-averaged neural trajectories during training, while all other methods were trained on single-trial data. This difference arises from the fundamental requirements of the decoder architectures: MINT needs to estimate typical trajectories while other methods don’t. Yet it might still be the case that other methods would benefit from including trial-averaged data in the training set, in addition to single-trial data. Alternatively, this might harm performance by creating a mismatch, between training and decoding, in the statistics of decoder inputs. We found that the latter was indeed the case: all non-MINT methods performed better when trained purely on single-trial data.

How challenging is generalization?

The contrasting assumptions in Figure 1a and 1b yield different implications regarding generalization, illustrated in Figure 6a and 6b. In both views, generalization is possible when a newly observed neural trajectory is ‘composed’ of previously observed elements. However, the nature of those elements is very different. Under the view in Figure 6a, those elements are neural states within the manifold. New movements involve novel neural trajectories, but those trajectories are composed of states within the previously observed manifold. Because those neural states are assumed to correlate with behavioral variables (e.g. velocity) those variables can be decoded during novel movements.

Implications of neural geometry for decoder generalization, in principle and in practice.

a) Under a traditional perspective, training that explores the full range of to-be-decoded variables will also explore the relevant neural manifold, allowing future generalization. Gray dots indicate neural states observed during training. Colored traces indicate newly observed trajectories. Because these trajectories traverse the previously defined manifold, they can be decoded by leveraging the established correlations between manifold dimensions and kinematics. b) Under an emerging perspective, generalization is possible when new trajectories lie within a similar region of factor-space as training-set trajectories. Gray-dashed traces indicate training-set neural trajectories. Colored traces indicate newly observed trajectories. Green and blue trajectories largely follow the flow-field implied by training-set trajectories, allowing generalization. In contrast, the orange trajectory evolves far from any previously explored region, challenging generalization. This may occur even when behavioral outputs overlap with those observed during training. c) Basic decoding performance on the MC_PacMan task for MINT (green) and a Wiener filter (red). Test sets used held-out trials from the same conditions as training sets. Force-decoding was excellent both when considering all conditions and when considering only ramps (fast and slow, increasing and decreasing). Results are based on the following numbers of training and test trials: All (234 train, 128 test), Ramps (91 train, 57 test). d) Generalization when training employed the four ramp conditions and testing employed a sine or chirp. Results are based on 91 training trials (all cases) and the following numbers of test trials: Ramps (57), 0.25 Hz Sine (19), 1 Hz Sine (12), 3 Hz Sine (15), Chirp (25). e) Data likelihoods indicate strained generalization. We computed the log-likelihood of the spiking observations for each neural state decoded by MINT. Rightward values indicate that those observations were unlikely even for the state selected as most likely to have generated those observations. Training employed the four ramp conditions. Distributions are shown when test data employed ramps, a 0.25 Hz sine, or a 3 Hz sine. f) Examples of force decoding traces (for held-out trials) when decoding using the Wiener filter. Top row: all conditions were used during training. Bottom row: training employed the four ramp conditions. Horizontal and vertical scale bars correspond to 2 seconds and 16 Newtons, respectively. g) As above, but when decoding using MINT. Trials and scale bars are matched across f and g. Representative trials were selected as those that achieved median decoding performance by MINT within each condition.

While this scenario would be advantageous, empirical observations question its assumptions. The constraint of low tangling implies that it is not possible to traverse previously observed states in entirely new ways. Indeed, even when encouraged to do so under BCI control, monkeys are incapable of producing new neural trajectories that conflict with the flow-field implied by previously seen trajectories [80, 81]. This suggests that newly observed trajectories will fall into two categories. First, new trajectories may be composed of trajectory segments that largely obey the flow-field implied by previously observed trajectories (Fig. 6b, green and blue traces). For example, the green trajectory climbs the scaffolding of previously observed single-speed trajectories, mimicking what might occur in a new situation where cycling speed increases steadily [53]. Second, newly observed trajectories may sometimes evolve in previously unobserved portions of state-space (Fig. 6b, orange trajectories) where the flow-field has never been estimated. This could occur if the internal computation must change considerably [102], a scenario that may be difficult to anticipate externally.

Under Figure 6a, MINT would be at a disadvantage relative to traditional decode methods. In particular, the standard Wiener filter should generalize well because it leverages neuron-behavior correlations and is not limited by any assumptions about a flow-field. In contrast, MINT might often generalize poorly because it cannot leverage neuron-behavior correlations and is constrained by previously observed trajectories. For example, if the green trajectory in Figure 6a is in the library, then MINT would be predisposed against decoding the blue trajectory during generalization because the two trajectories approach one another with opposing derivatives. The Wiener filter would have no such limitation.

If the scenario in Figure 6b holds, generalization will hinge less on the decoder itself and more on how new neural trajectories (colored traces) relate to the previously observed scaffolding (dashed gray traces). When newly observed trajectories hew close to that scaffolding (e.g. green and blue) any good decode method should generalize well. When newly observed trajectories are far from the scaffolding (e.g. orange), all methods will struggle. MINT will struggle because it has no scaffolding in this region. Other methods will struggle because previously observed neuron-behavior correlations are no longer valid.

We desired a dataset where the predictions of Figure 6a and 6b are likely to be different. With the possible exception of MC_RTT, the datasets above lack this property; generalization to held-out trials is not particularly challenging and could be subserved by the features in Figure 6a or by decoding something like the green trajectory in Figure 6b. One way to challenge generalization, using existing data, is to restrict training to fewer conditions while ensuring that training still employs a broad range of values for each decoded variable. In the MC_Cycle dataset, the same set of x- and y-velocities occur during forward and backward cycling, just at different phases. If neural activity were dominated by correlations with velocity, traditional methods that leverage those correlations should generalize across cycling directions. Instead, we found that all tested methods failed to generalize when trained on one cycling direction and tested on the other. Most R2 values were close to zero or even negative. The likely reason is that neural trajectories for forward and backward cycling occupy nearly orthogonal subspaces (Fig. S2), especially for this particular dataset [20]. These observations are consistent with the view in Figure 6b. Of course, this did not have to be the case; the empirical neural trajectories during cycling could have obeyed the hypothesis in Figure 6a and thus supported generalization.

Whether one finds the above test conclusive depends upon which behavioral variables one considers primary. If only velocity (or only position) matters, then generalization should have been easy under Figure 6a because the same range of velocities (and positions) was present in training and test sets. Yet one may have different views regarding which variables are central. One may thus suspect that generalization failed because those variables (or their combinations) were insufficiently sampled during training. To address this concern, we explored generalization using a dataset where it is simple to confirm that the relevant behavioral variables are fully explored during training. The MC_PacMan dataset involved Neuropixels-based recordings of motor cortex spiking activity and simultaneous measurements of isometric force applied to an immovable handle. All forces were generated in a forward direction (away from the body) and the arm did not move. Force is thus the central behavioral variable in this task; all other behavioral variables (including muscle activity and cursor height) mirror force or its derivative. Every condition explores a similar range of forces, yet involves very different temporal force profiles. Under the scenario in Figure 6a, generalization to new force profiles should be straightforward (for traditional methods) because neural states associated with each force level have already been observed. New force profiles are simply new trajectories through previously observed states. This scenario thus predicts that a simple traditional decoder, such as a Wiener filter, will generalize well but that MINT will not. In contrast, under the scenario in Figure 6b, some forms of generalization should be possible (for any method) while others will be intrinsically challenging (for any method).

To test these predictions, we trained both MINT and a Wiener filter to decode force. Training data included fast and slow increasing and decreasing ramps, and thus spanned the full range of forces used in this task. Generalization was tested using sinusoids (0.25, 2, and 3 Hz) and a slow-to-fast chirp. In agreement with Figure 6b, generalization performance was strongly condition-specific and only weakly decoder-specific (Fig. 6d,f,g). Both MINT and the Wiener filter generalized well to the 0.25 Hz sinusoid: performance was almost as good as when training included all conditions. Yet both decoders generalized poorly at higher frequencies, especially the 3 Hz sinusoid. This was true even though the training data contained swiftly increasing and decreasing forces.

These empirical results make little sense under Figure 6a, because training data should have explored the relevant manifold. However, they do make sense under Figure 6b. Under this perspective, neural activity is not dominated by correlations with behavioral variables, but is shaped by the flow-field performing the underlying computation. Two conditions can involve different flow-fields, even if their behavioral outputs span similar ranges. Generalization will be strained in such situations, which may be difficult to anticipate ‘from the outside’.

In instances where generalization is strained, and performance is poor overall, different decoders may experience performance drops of different, and perhaps difficult-to-predict, magnitudes. For example, MINT outperformed the Wiener filter when generalizing from ramps to the chirp, while the Wiener filter outperformed MINT when generalizing from ramps to the 1 Hz sine. This observation does not suggest a meaningful advantage for either decoder. These advantages were idiosyncratic and, more importantly, overall performance was so poor that either decoder would have been essentially unusable without retraining that includes additional conditions. The MC_RTT task may also be an example of this effect: the two expressive methods outperformed MINT, but all decoders achieved their worst velocity-decode performance for this dataset (Fig. 5, bottom).

For these reasons, one may wish to know, during decoding, if generalization is becoming strained. Usefully, MINT computes the log-likelihood of spiking observations for its estimated neural state (which is the most-probable state it can find). A prediction of Figure 6b is that, when generalization is strained, these likelihoods should become unusually low because the true neural state (which generated the spiking observations) will often be far from any state that MINT can estimate. To test this prediction, we computed the distribution of negative log-likelihoods (Fig. 6e) for all estimated neural states during the test of generalization described above. The distribution for 0.25 Hz sinusoids (purple) was similar to that for ramps (blue). This is consistent with the set of true neural states, during the 0.25 Hz sinusoids, being similar to states that are ‘reachable’ by MINT’s interpolation step. In contrast, the distribution for 3 Hz sinusoids (orange) showed a second mode of spiking observations that were particularly unlikely. This is expected if the true neural state, at these moments, is far from any state that can be reached by MINT’s interpolation (as would be true for the orange trace in Figure 6b). Data log-likelihoods could thus potentially be used to indicate – without having to ask the participant – when decoding is strained because neural states are far from those observed during training. Training could then be modified to include additional conditions.

Modeling & preprocessing choices

MINT uses a direct mapping from neural states to behavioral states to instantiate highly nonlinear decoding, then uses locally linear interpolation to support decoding of behavioral states not observed during training. To determine the impact of these choices on performance, we ran a full factorial analysis that always used neural state estimates generated by MINT, but compared the direct neural-to-behavioral-state mapping to a linear neural-to-behavioral-state mapping (Fig. S3a), and assessed the impact of interpolation (Fig. S3b). Both choices did indeed improve performance. We also assessed the impact of acausal versus causal decoding (Fig. S3c). Causal decoding is important for real-time BCIs, and is thus employed for all our main decoding analyses. However, future applications could potentially introduce a small lag into online decoding, effectively allowing some ‘acausal decoding’ at the expense of longer-latency responses. As expected, acausal decoding provided modest performance increases, raising the possibility of an interesting tradeoff. Decoding can be zero-lag (causal), positive-lag (acausal), or even negative-lag. The latter would result in a very ‘snappy’ decoder that anticipated actions at the expense of some decoding accuracy. The lag hyperparameter wouldn’t need to be specified prior to training and could be freely adjusted at any time.

All datasets were curated to contain sorted spikes. Yet during online performance, decoding must typically rely on unsorted threshold crossings. Prior studies have found that moving from sorted spikes to threshold crossings produces a modest but tolerable drop in performance [103]. We similarly found that moving from sorted spikes to threshold crossings (selected to be from ‘good’ channels with reasonable signal-to-noise) produced only a small drop in performance (Fig. S3d). The operations of MINT highlight why such robustness is expected. When spikes from two neurons are combined, the resulting ‘unit’ acts statistically as if it were a high firing-rate neuron. Spiking is still approximately Poisson, because the sum of Poisson processes is a Poisson process. Thus, MINT’s statistical assumptions remain appropriate.

Neural state estimation

MINT’s success at behavioral decoding suggests underlying success in neural-state estimation. However, that would not necessarily have to be true. For example, the GRU also typically supported excellent behavioral decodes but doesn’t provide neural state estimation. MINT does (by its very construction) but whether those estimates are accurate remains to be evaluated.

To evaluate neural-state estimates, we submitted firing-rate predictions to the Neural Latents Benchmark [104], an online benchmark for latent variable models that compares acausal neural-state estimates across methods and datasets. The four primary datasets for the benchmark are MC_Maze, Area2_Bump, MC_RTT, and an additional dataset, DMFC_RSG, that contains neural recordings from dorsomedial frontal cortex while a monkey performed a cognitive timing task [62]. Three secondary datasets (MC_Maze-L, MC_Maze-M, and MC_Maze-S) contain additional sessions of the maze task with progressively fewer training trials (500, 250, 100, respectively).

By submitting to the Neural Latents Benchmark, we can ask how MINT performs relative to a set of ‘baseline’ methods. One baseline method (smoothed spikes) is traditional and simple. Other baseline methods, Gaussian Process Factor Analysis (GPFA) [105] and Switching Linear Dynamical System (SLDS) [106], are modern yet well-established. Finally, two baseline methods are quite new and cutting-edge: Neural Data Transformers (NDT) [92] and AutoLFADS [107]. These are expected to provide a high bar by which to judge the performance of other approaches.

Although the neural state is a well-defined quantity in spiking models [55], experimental data does not typically provide a ground-truth neural state against which estimates can be compared (unlike for behavioral-state decoding). Yet one can define attributes that an accurate neural-state decode should possess. A critical attribute is prediction of spiking activity. This attribute is assessed via bits per spike, which is closely related to the log-likelihood of spiking observations given a set of firing rates (assuming Poisson spiking). MINT performed similarly to AutoLFADS and was consistently slightly better than NDT on the four primary datasets (Fig. 7a, left). MINT performed slightly better than AutoLFADS and NDT on the largest of the secondary datasets. That advantage grew as dataset-size shrank (Fig. 7a, right). MINT outperformed the other three baseline methods by modest-to-sizeable margins for all seven datasets.

Evaluation of neural state estimates for seven datasets.

a) Performance quantified using bits per spike. The benchmark’s baseline methods have colored markers. All other submissions have gray markers. Vertical offsets and vertical ticks are used to increase visibility of data when symbols are close due to similar values. Results with negative values are designated by markers to the left of zero, but their locations don’t reflect the magnitude of the negative values. Data from Neural Latents Benchmark (https://neurallatents.github.io/). b) Performance quantified using PSTH R2. c) Performance quantified using velocity R2, after velocity was decoded linearly from the neural state estimate. A linear decode is used simply as a way of evaluating the quality of neural state estimates, especially in dimensions relevant to behavior.

MINT was not designed to maximize bits per spike – good performance in this regard simply results from the primary goal of estimating a neural state containing firing rates. This matters because there exist spiking features – e.g. synchrony beyond that due to correlated rates – that some methods may be able to leverage when predicting spike probabilities. Doing so is reasonable but separate from MINT’s goal of estimating a rate-based neural state. It is thus worth considering other attributes expected of good neural-state estimates. The Neural Latents Benchmark includes two such attributes. First, if state estimation is accurate, trial-averaged neural state estimates should resemble empirical trial-averaged firing rates (computed for test trials), an attribute assessed by PSTH R2. MINT achieved higher PSTH R2 values than every other method on every dataset (Fig. 7b). This is in some ways unsurprising: MINT estimates neural states that tend to resemble (at least locally) trajectories ‘built’ from training-set-derived rates, which presumably resemble test-set rates. Yet strong performance is not a trivial consequence of MINT’s design. MINT does not ‘select’ whole library trajectories; PSTH R2 will be high only if condition (c), index (k), and the interpolation parameter (α) are accurately estimated for most moments.

A second attribute is that, in sensorimotor areas during motor tasks, behavior should be decodable from the neural state. The benchmark thus measured R2 for a linear decode of velocity from the neural state. MINT outperformed all baseline approaches (Fig. 7c) for all datasets. (MINT could of course have achieved even better decoding using its typical direct association, but that would have undermined the utility of this comparison).

In addition to the five baseline methods, the Neural Latents Benchmark solicited submissions from other methods and many were submitted. These submissions spanned a broad range in terms of performance (gray vertical lines in Fig. 7), highlighting the challenging nature of this problem. Some submissions involved ensemble approaches, some of which outperformed both MINT and the baseline methods in terms of bits per spike. For the primary datasets, the highest bits per spike overall was achieved by an ensemble of SpatioTemporal Neural Data Transformers [108], achieving .386 bits per spike on the MC_Maze dataset, compared to .330 for MINT. For the smaller secondary datasets, the very best submissions provided minimal-to-modest improvement over MINT in terms of bits per spike (Fig. 7a, right), and the majority performed less well.

The ability of some expressive neural networks (and some ensembles of these networks) to improve the bits per spike metric indicates there exists variability in the neural recordings that MINT does not capture. That variability could arise from behaviorally irrelevant fluctuations in the neural state, from synchrony based effects, or from instabilities in the neural recordings. These are ‘real’ things and it is thus valid for a method to leverage them to predict spiking when that is the goal. Yet as noted above, they may often be incidental to one’s goals in estimating the neural state, highlighting the need to consider additional attributes such as PSTH R2 and velocity R2. For PSTH R2 (Fig. 7b), MINT outperformed all other methods for all six datasets. For velocity R2 (Fig. 7c), MINT had the best performance for three out of six datasets and was very close to the highest R2 values for the remaining three datasets (the largest deficit was ΔR2 = .012).

In summary, MINT’s strong decode performance (Fig. 5) does indeed follow from good neural-state estimation, as one would hope. MINT provided estimates that were competitive with the best present methods. This includes existing well-documented methods, as well as a bevy of additional submissions. This implies that MINT could be used in situations where neural-state estimation is the goal, possibly including monitoring of the neural state for clinical purposes. MINT uses simple operations and is readily implemented causally – indeed this is typically how we would expect to see it used. Yet MINT’s performance is typically similar to – and often better than – that of complex methods that would likely have to be limited to offline situations.

Practical considerations

Training & execution times

For MINT, ‘training’ simply means computation of standard quantities (e.g. firing rates) rather than parameter optimization. MINT is thus typically very fast to train (Table 1), on the order of seconds using generic hardware (no GPUs). This speed reflects the simple operations involved in constructing the library of neural-state trajectories: filtering of spikes and averaging across trials. At the same time we stress that MINT is a method for leveraging a trajectory library, not a method for constructing it. One may sometimes wish to use alternatives to trial-averaging, either of necessity or because they improve trajectory estimates. For example, for the MC_RTT task we used AutoLFADS to infer the library. Training was consequently much slower (hours rather than seconds) because of the time taken to estimate rates. Training time could be reduced back to seconds using a different approach – grouping into pseudo-conditions and averaging – but performance was reduced. Thus, training will typically be very fast, but one may choose time-consuming methods when appropriate.

Execution times were well below the threshold for real-time applications (Table 1), even using a laptop computer. State estimation in MINT is 𝒪(NTtotal), where Ttotal is the total number of times (across conditions) in the library of neural trajectories. This is 𝒪(NC) when making the simplifying assumption that all conditions are of equal length. Memory requirements are similarly 𝒪((N + M)C), where M is the number of behavioral variables (a value that will typically remain quite small). Thus, both execution times and memory requirements grow only linearly with neurons or with conditions. Additionally, much of the estimation procedure in MINT is condition-specific and therefore parallelizable. This makes MINT a good candidate for decoding in BCI applications that involve large neuron counts and/or large behavioral repertoires.

MINT performs well on small training sets

To investigate robustness to training with small trial-counts, we leveraged the fact that the four maze-reaching datasets – one primary (MC_Maze) and three secondary (MC_Maze-L, MC_Maze-M, and MC_Maze-S) – contained training sets of decreasing size. Because training involved computing trial-averaged rates, performance is expected to decline when fewer trials are available for averaging. For comparison, we chose the GRU and the feedforward neural network because of their generally good position and velocity decoding performance in the MC_Maze task. We assessed performance in terms of position and velocity R2 (Fig. 8a). MINT always performed at least as well as the two expressive methods. This gap was often (though not always) larger for smaller training sets. More broadly, MINT was quite robust to small training sets: an approximately five-fold reduction in training data led to a decline in R2 of only .037 (position) and .050 (velocity).

Decoding robustness in the face of small training sets and reduced neuron counts.

a) R2 values for MINT and two neural network decoders (GRU and the feedforward network) when decoding position and velocity from four maze datasets with progressively fewer training trials per condition. Results are based on the following numbers of training and testing trials: MC_Maze-S (75 train, 25 test), MC_Maze-M (188 train, 62 test), MC_Maze-L (375 train, 125 test), MC_Maze (1721 train, 574 test). MC_Maze contains 108 conditions and the other maze datasets each contain 27 conditions. b) Decoding performance in the face of undetected (dashed traces) and known (solid traces) loss of neurons. R2 values are shown for MINT (green) and the Wiener filter (red) when decoding position and velocity from the MC_Maze-L dataset (same train/test trials as in a). To simulate undetected neuron loss, we chose k neurons (randomly, with k being the number of lost neurons out of 162 total) and eliminated their spikes without adjusting the decoders. This procedure was repeated 50 times for each value of k. Traces show the mean and standard deviation of the sampling distribution (equivalent to the standard error). To simulate known neuron loss, we followed the above procedure but altered the decoder to account for the known loss. For MINT, this simply involved ignoring the lost neurons when computing data likelihoods. In contrast, the Wiener filter had to be retrained using training data with those neurons eliminated (as would be true for most methods). This was done separately for every set of lost neurons.

Performance when neurons are lost

It is desirable for a decoder to be robust to the unexpected loss of the ability to detect spikes from some neurons. Such loss might occur while decoding, without being immediately detected. Additionally, one desires robustness to a known loss of neurons / recording channels. For example, there may have been channels that were active one morning but are no longer active that afternoon. At least in principle, MINT makes it very easy to handle this second situation: there is no need to retrain the decoder, one simply ignores the lost neurons when computing likelihoods. This is in contrast to nearly all other methods, which require retraining because the loss of one neuron alters the optimal parameters associated with every other neuron.

To evaluate, we performed a neuron-dropping analysis using the MC_Maze-L dataset (Fig. 8b). For comparison, we also evaluated a Wiener filter. The Wiener filter provides a useful comparison because (like MINT) it is easy to retrain quickly, with reliable results, across many training sets (which was critical in the present situation). To simulate undetected neuron loss, we chose a random subset of neurons and removed all their spikes, without making any adjustment to the decoders. This simulates a neuron (or channel) falling silent. This was repeated many times with different random draws. To simulate known neuron loss, we did the same but allowed the decoders to adjust: by ignoring the lost neurons in the case of MINT, and by retraining all parameters in the case of the Wiener filter. We evaluated both position (Fig. 8b, top) and velocity (bottom) decoding.

For MINT, undetected loss of fewer than 20 (of 162) neurons caused essentially no performance deficit. Average position and velocity decoding R2 remained above 95% of peak performance until fewer than 113 and 109 neurons remained, respectively. Large undetected losses – e.g. of half the neurons – caused major deficits in decoding. Yet decoding remained accurate if those losses were known, with the lost neurons ignored during decoding. MINT was extremely robust in this situation: MINT’s average R2 for position decoding remained above 95% of peak performance until fewer than 58 neurons (36%) remained. Results were similar for velocity decoding: performance was >95% of its peak until fewer than 62 neurons (38%) remained. The Wiener filter was also fairly robust, but less so than MINT. To remain above 95% of peak performance, the Wiener filter needed approximately twice as many remaining neurons: 118 for position and 126 for velocity.

These results illustrate that MINT is reasonably robust to undetected neuron loss. In practice, the loss of many neurons would likely be detected. In such cases, MINT is very robust. Additionally, because there is no need for any true retraining, any adjustment to known neuron-loss can occur on the fly, with no need to pause decoding.

Interpretability

One form of interpretability relates to whether an algorithm uses ‘black box’ optimization to learn a solution. For example, neural networks can be highly expressive and learn constraints directly from data. However, when constraints are known, one may prefer to rely more on explicit assumptions and less on expressivity. Interpretable methods, such as the Kalman filter, rely on explicit assumptions. Such assumptions may (if they match the data) improve performance. They may also provide other potential advantages. For example, the Kalman filter is a canonical ‘interpretable’ algorithm, with a probabilistic graphical model that makes an explicit Gaussian assumption regarding distributions of spike count measurements. This explicitness may be helpful in understanding successes and failures (e.g. the Gaussian assumption can break down during low-rate periods).

The performance of interpretable methods should tend to reflect the extent to which their assumptions match the statistics of the data. To investigate, we considered the performance of both the Kalman filter and MINT on the MC_Maze reaching dataset. The Kalman filter produces a relatively poor decode (R2 = .609 for hand velocity, versus .914 for MINT). This doesn’t imply the Kalman filter is a poor algorithm, simply that its assumptions may not be well-aligned with the data. We confirmed this by generating synthetic data whose properties are closer to those assumed by the Kalman filter. The Kalman filter then performed quite competitively (Fig. S4). This result underscores that MINT does not outperform the Kalman filter because it is intrinsically a better method. Instead, MINT performs well because its assumptions are presumably a better match to the underlying properties of the data. If Figure 1a were the correct perspective, the Kalman filter would have been expected to perform at least as well as MINT. Under Figure 1b the Kalman filter is predicted to perform less well, which is indeed what we found.

MINT is also interpretable regarding why each prediction is made. Suppose that, at a given moment during decoding, MINT selects neural state x as the most likely state and renders its decode using the associated behavioral state z. As analyzed in Figure 6e above, selection of x comes with an associated likelihood that conveys information regarding how well the spiking data matched that neural state. Likelihoods are known not only for this ‘selected’ state, but for all non-selected states on all trajectories in the library. These likelihoods can be analyzed to understand why the selected state was chosen over the others, how close the algorithm was to selecting a different state, and which neurons contributed most strongly to state selection. More generally, because of its probability-based approach, MINT is readily modifiable and extensible in a variety of ways (Fig. S5).

Discussion

Implications regarding neural geometry

Our results argue that the empirical structure of motor-cortex activity is closer to the hypothesis in Figure 1b. We began with the prediction that, under that hypothesis, MINT should consistently outperform existing interpretable methods such as the Kalman and Wiener filter. This was indeed the case: MINT always performed better than these two methods, often by sizable margins. In contrast, MINT and the Kalman filter performed comparably on simulated data that better approximated the assumptions in Figure 1a. Thus, MINT is not a ‘better’ algorithm – simply better aligned with the empirical properties of motor cortex data. This highlights an important caveat. Although MINT performs well when decoding from motor areas, its assumptions may be a poor match in other areas (e.g. the hippocampus). MINT performed well on two non-motor-cortex datasets – Area2_Bump (S1) and DMFC_RSG (dorsomedial frontal cortex) – yet there will presumably be other brain areas and/or contexts where one would prefer a different method that makes assumptions appropriate for that area.

We also began with the prediction that, under the hypothesis in Figure 1b, MINT should be competitive with highly expressive approaches that can learn their constraints from data. MINT may even perform better in some cases because MINT can begin with appropriate constraints rather than having to learn them from training data. At the same time, MINT might sometimes be at a disadvantage if there are additional constraints or features that it does not take into account (a simple example would be drift in firing rates over time – one would desire a drift stabilization module to correct for this, as in Fig. S5). Empirically, decoding performance was often exceedingly similar for MINT and both expressive methods (the feedforward network and GRU). Consider velocity, which is probably the most commonly decoded parameter for motor BCIs. Across the four datasets in Figure 5, the average R2 for the velocity decode was .841 (MINT), .830 (GRU), and .837 (feedforward network). These values are all considerably higher than for the Kalman filter (.630) and Wiener filter (.719). There were multiple cases where MINT performed noticeably better than either expressive method, and one case (the MC_RTT dataset) where MINT performed noticeably worse. These results are consistent with the set of expectations outlined above: MINT should be highly competitive with expressive methods, often performing modestly better and sometimes performing modestly worse.

The hypothesis that MINT’s assumptions are well-aligned with the data is further supported by its performance during neural-state estimation. Of the well-established baseline methods for state-estimation, the highest performing were AutoLFADS and a Neural Data Transformer (NDT). Across a variety of datasets and performance metrics, MINT either performed similarly to these methods or performed noticeably better. MINT consistently outperformed simple lightweight state estimators, despite being a simple lightweight estimator itself. We find it noteworthy that MINT naturally performs very well as a neural-state estimator, despite being designed as a decoder of behavior. MINT does so with no need for modification. This relates to the observation that MINT can readily switch from decoding one parameter (e.g. cycling phase) to a very different one (muscle activity) with no need for additional training, an option not available with any other standard method.

Practical considerations for use of MINT as an online decoder

Across the four datasets we tested, the best decoders (MINT and the two expressive methods) were all very close in terms of performance. Given this, decoder-choice would likely be based on considerations beyond raw performance. Here MINT has many advantages. Because MINT’s computations are so simple, it is typically fast to train and always very fast to run. Its interpretability supports principled extensions (discussed below) and the ability to investigate the source of any failures (as in Figure 6e). These considerations, combined with the fact that MINT often yielded the best behavioral decode of all methods, makes MINT a compelling candidate for online decoding in scientific or clinical applications. With that goal in mind, there exist three important practical considerations. First, some decode algorithms experience a performance drop when used online. One presumed reason is that, when decoding is imperfect, the participant alters their strategy which in turn alters the neural responses upon which decoding is based. Because MINT produces particularly accurate decoding, this effect may be minimized, but this cannot be known in advance. If a performance drop does indeed occur, one could adapt the known solution of retraining using data collected during online decoding [13]. Another presumed reason (for a gap between offline and online decoding) is that offline decoders can overfit the temporal structure in training data [109]. This concern is somewhat mitigated by MINT’s use of a short spike-count history, but MINT may nevertheless benefit from data augmentation strategies such as including time-dilated versions of learned trajectories in the libraries.

A second practical consideration is that, for a participant with limited ability to move, training would involve asking subjects to observe and/or attempt movements. Both strategies engage neurons in motor cortex [110], but can also create uncertainty regarding the exact behavioral label for each moment (a challenge any decode method must face). MINT’s decoding performance benefits from good estimates of Ω and Φ, and will decline if those estimates are poor or misaligned. Fortunately, methods exist for temporally aligning neural data based on spiking activity alone [111] and the performance of such methods should only improve as more neurons are simultaneously recorded. Such methods would allow computation of trial-averaged rates (or single-trial rates) that could then be aligned with a canonical movement trajectory (e.g. the usual bell-shaped velocity profile). Conveniently, any alterations to that alignment are trivial to accomplish, including at runtime. For example, to shift velocity earlier, one would simply shift MINT’s library indices. More generally, MINT is amenable to a variety of methods – simple or sophisticated – for inferring the trajectory library. Possible future approaches include leveraging the typical neural geometry for a particular task. This could allow a participant-specific library to be learned using fewer trials. This might even be done in an unsupervised way (similar to Dyer et al. [112]): a trajectory library could be built without behavioral labels, then appropriately linked to a ‘reference’ behavioral library via prior knowledge of task-specific neural geometry. As a simple example, cycling across different speeds produces a tube-shaped manifold with different velocities at each end [53]. An empirically observed tube might be given behavioral labels based on this knowledge.

A final practical consideration concerns generalization. Because MINT uses a library of canonical trajectories, it is tempting to suppose that its decoded trajectories will simply be replicas of a limited set of stereotyped behaviors. By removing the interpolation step, MINT can indeed approximate this mode, which would be desirable in situations where the to-be-decoded behavior is itself stereotyped (e.g. producing keystrokes or handwritten characters). Yet one often wishes to decode behaviors beyond those observed in the training set. MINT can do so if the empirical neural trajectory mostly (and locally) obeys the flow-field implied by the library. This provides considerable, but not unlimited, flexibility. Might MINT thus be at a disadvantage, relative to other methods, with respect to generalization? This is predicted to be true under the hypothesis in Figure 1a and 6a. However, it is not expected to be true under the hypothesis in Figure 1b and 6b. Under that hypothesis, the forms of generalization that are unavailable to MINT will also be unavailable to standard methods. Consistent with that prediction, generalization success for the MC_PacMan dataset was condition-specific, not decoder-specific.

Multi-task decoding

The fact that generalization may often be challenging does not imply that decoding across multiple tasks is a challenge. Indeed MINT is ideally suited for this situation. For example, MINT successfully decoded motor output for both tasks performed by the network in Figure 4a, and across all conditions in the MC_PacMan dataset (when trained on all conditions, Fig. 6c). As long as trajectories relevant to both tasks are present in the library, ‘task switching’ occurs naturally and implicitly. Because MINT does not rely on correlations between neural activity and behavioral variables, task switching will not be impaired if those correlations change across tasks, something anticipated under Figure 1b and 6b. As the number of situations decoders are expected to handle increases, this may be a useful property.

Future avenues

MINT’s probability-based framework is amenable to principled extensions. Rather than simply decode the state with the highest data likelihood, one could incorporate priors. Priors could be fixed or guided by incoming information (e.g. an eye tracker). Decoding could also focus on utility functions rather than probabilities. This would respect the fact that a participant may value avoiding certain types of errors more than others. For example, erroneous transitions from stationary to moving may be jarring or unsafe, and could be dissuaded via an appropriately designed utility function. These extensions are natural because MINT computes data likelihoods for a broad range of candidate states before choosing the maximum-likelihood state. Converting likelihoods to probabilities (based on a prior) or to an estimated cost can thus be done in a principled manner.